---
title: "Microbiome composition"
bibliography: 
- bibliography.bib
output: 
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---
  <!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{microbiome tutorial - composition}
  %\usepackage[utf8]{inputenc}
  %\VignetteEncoding{UTF-8}  
-->
  
  
Also see [phyloseq barplot examples](http://joey711.github.io/phyloseq/plot_bar-examples.html).
  
Read example data from a [diet swap study](http://dx.doi.org/10.1038/ncomms7342):
  
```{r composition-example1}
# Example data
library(microbiome)
library(dplyr)
data(dietswap)

# Just use prevalent taxa to speed up examples
# (not absolute counts used in this example)
pseq <- core(dietswap, detection = 10^3, prevalence = 95/100)

# Pick sample subset
library(phyloseq)
pseq2 <- subset_samples(pseq, group == "DI" & nationality == "AFR" & timepoint.within.group == 1)
```

### Barplots for composition

Same with compositional (relative) abundances; for each sample (left), or averafged by group (right).
  
```{r composition-example4b, fig.width=12, fig.height=5, out.width="400px", fig.show="hold", warning=FALSE, message=FALSE}
# Try another theme
# from https://github.com/hrbrmstr/hrbrthemes
library(hrbrthemes)
library(gcookbook)
library(tidyverse)

# Limit the analysis on core taxa and specific sample group
p <- plot_composition(pseq2,
		      taxonomic.level = "OTU",
                      sample.sort = "nationality",
                      x.label = "nationality",
                      transform = "compositional") +
     guides(fill = guide_legend(ncol = 1)) +
     scale_y_percent() +
     labs(x = "Samples", y = "Relative abundance (%)",
                                   title = "Relative abundance data",
                                   subtitle = "Subtitle",
                                   caption = "Caption text.") + 
     theme_ipsum(grid="Y")
print(p)  

# Averaged by group
p <- plot_composition(pseq2,
                      average_by = "bmi_group", transform = "compositional")
print(p)
```



### Composition heatmaps


Heatmap for CLR-transformed abundances, with samples and OTUs sorted
with the neatmap method:
  
```{r composition-example7, fig.width=10, fig.height=4}
tmp <- plot_composition(pseq2, plot.type = "heatmap", transform = "clr",
                      sample.sort = "neatmap", otu.sort = "neatmap",
                      mar = c(6, 13, 1, 1))
```



### Plot taxa prevalence

We use the Dynamics IBD data set from [Halfvarson J., et al. Nature
Microbiology, 2017](http://www.nature.com/articles/nmicrobiol20174) as
downloaded from [Qiita ID
1629](https://qiita.ucsd.edu/study/description/1629). This function
allows you to have an overview of OTU prevalences alongwith their
taxonomic affiliations. This will aid in checking if you filter OTUs
based on prevalence, then what taxonomic affliations will be lost.

```{r plot_prev, fig.height=6, fig.width=8, dev="CairoPNG"}
data(DynamicsIBD)

# Use sample and taxa subset to speed up example
p0 <- subset_samples(DynamicsIBD, sex == "male" & timepoint == 1)

# Check the names of the taxonomic level 
colnames(tax_table(p0)) 

# Change the "Rank" label to taxonomic levels
colnames(tax_table(p0)) <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")

p0 <- core(p0, detection = 10, prevalence = 0)

# For the available taxonomic levels
plot_taxa_prevalence(p0, "Phylum")
```

