---
title: "Core microbiome"
bibliography: 
- bibliography.bib
output: 
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---
<!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{microbiome tutorial - core}
  %\usepackage[utf8]{inputenc}
  %\VignetteEncoding{UTF-8}  
-->

## Core microbiota analysis

See also related functions for the analysis of rare and variable taxa (noncore_members; noncore_abundance; rare_members; rare_abundance; low_abundance).

Load example data:

```{r core-prevalence}
# Load data
library(microbiome)
data(peerj32)

# Rename the data
pseq <- peerj32$phyloseq

# Calculate compositional version of the data
# (relative abundances)
pseq.rel <- microbiome::transform(pseq, "compositional")
```


### Prevalence of taxonomic groups

Relative population frequencies; at 1% compositional abundance threshold:

```{r core-prevalence2}
head(prevalence(pseq.rel, detection = 1, sort = TRUE))
```


Absolute population frequencies (sample count):

```{r core-prevalence2b}
head(prevalence(pseq.rel, detection = 1, sort = TRUE, count = TRUE))
```


### Core microbiota analysis

If you only need the names of the core taxa, do as follows. This returns the taxa that exceed the given prevalence and detection thresholds. 

```{r core-members, message=FALSE, warning=FALSE, eval = FALSE}
core.taxa.standard <- core_members(pseq.rel, detection = 0, prevalence = 50/100)
```


A full phyloseq object of the core microbiota is obtained as follows:

```{r core-data, message=FALSE, warning=FALSE}
pseq.core <- core(pseq.rel, detection = 0, prevalence = .5)
```


Retrieving the associated taxa names from the phyloseq object:

```{r core-taxa, message=FALSE, warning=FALSE}
core.taxa <- taxa(pseq.core)
```



### Core abundance and diversity

Total core abundance in each sample (sum of abundances of the core members):

```{r core-ab, message=FALSE, warning=FALSE}
core.abundance <- sample_sums(core(pseq.rel, detection = .01, prevalence = .95))
```


## Core visualization

### Core line plots

Determine core microbiota across various abundance/prevalence
thresholds with the blanket analysis [(Salonen et al. CMI, 2012)](http://onlinelibrary.wiley.com/doi/10.1111/j.1469-0691.2012.03855.x/abstract) based on various signal and prevalences.

```{r core2, fig.width=9, fig.heigth=6, out.width="400px", warning=FALSE}
# With compositional (relative) abundances
det <- c(0, 0.1, 0.5, 2, 5, 20)/100
prevalences <- seq(.05, 1, .05)
plot_core(pseq.rel, prevalences = prevalences, detections = det, plot.type = "lineplot") + xlab("Relative Abundance (%)")
```


### Core heatmaps

This visualization method has been used for instance in [Intestinal microbiome landscaping: Insight in community assemblage and implications for microbial modulation strategies](https://academic.oup.com/femsre/article/doi/10.1093/femsre/fuw045/2979411/Intestinal-microbiome-landscaping-insight-in#58802539). Shetty et al. _FEMS Microbiology Reviews_ fuw045, 2017.

Note that you can order the taxa on the heatmap with the order.taxa argument.

```{r core-example3, fig.width=8, fig.heigth=18, fig.show='hold', out.width = '200px', warning=FALSE}
# Core with compositionals:
prevalences <- seq(.05, 1, .05)
detections <- 10^seq(log10(1e-3), log10(.2), length = 10)

# Also define gray color palette
gray <- gray(seq(0,1,length=5))
p <- plot_core(pseq.rel, plot.type = "heatmap", colours = gray,
    prevalences = prevalences, detections = detections) +
    xlab("Detection Threshold (Relative Abundance (%))")
print(p)    


# Same with the viridis color palette
# color-blind friendly and uniform
# options: viridis, magma, plasma, inferno
# https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html
# Also discrete=TRUE versions available
library(viridis)
print(p + scale_fill_viridis())

# Core with absolute counts and horizontal view:
# and minimum population prevalence (given as percentage)
detections <- 10^seq(log10(1), log10(max(abundances(pseq))/10), length = 10)

library(RColorBrewer)
p <- plot_core(pseq, plot.type = "heatmap", 
       		 prevalences = prevalences,
       		 detections = detections,
		 colours = rev(brewer.pal(5, "Spectral")),
		 min.prevalence = .2, horizontal = TRUE)
print(p)
```

