---
title: "Installation of microbiome R tools"
bibliography: 
- bibliography.bib
output: 
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---
<!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{microbiome tutorial - Installation}
  %\usepackage[utf8]{inputenc}
  %\VignetteEncoding{UTF-8}  
-->


### Installing R

**If you do not already have R/RStudio installed**, do as follows. 

  1. Install [R](http://www.r-project.org/) 
  1. Install [RStudio](http://rstudio.org)
  1. With Windows, install also [RTools](http://cran.r-project.org/bin/windows/Rtools/) (version corresponding to your R version)


### Installation

Open R and install the package. If the installation fails, ensure from
the RStudio tools panel that you have access to the Bioconductor
repository.

```{r installation, message=FALSE, warning=FALSE, eval=FALSE}
library(BiocInstaller)
source("http://www.bioconductor.org/biocLite.R")
useDevel()
biocLite("microbiome")
```

Alternatively, to install the bleeding edge (potentially unstable)
development version, run in R:

```{r microbiomeinstall, message=FALSE, warning=FALSE, eval=FALSE}
library(devtools) # Load the devtools package
install_github("microbiome/microbiome") # Install the package
```

### Use 

Once the package has been installed, load it in R

```{r loading, eval=TRUE, message=FALSE}
library(microbiome)  
```

For a brief overview, see the [package vignette](https://bioconductor.org/packages/devel/bioc/vignettes/microbiome/inst/doc/vignette.html) and the more extensive [on-line tutorial](http://microbiome.github.io/microbiome/).



### Reproducible document generation

To test reproducible document generation:

 1. Start [RStudio](http://www.rstudio.com/)
 1. Open a new Rmarkdown (.html) file 
 1. Convert that .html file with the 'knit' button
 1. Modify the .html file and knit it again to make your own reproducible report


### Further reading

 * [Rmarkdown tips](http://rmarkdown.rstudio.com/)
 * [R cheat sheets](http://devcheatsheet.com/tag/r/)
 * [Using Github with R and RStudio](http://www.molecularecologist.com/2013/11/using-github-with-r-and-rstudio/)
 * [Molecular ecologist's view on code sharing](http://www.molecularecologist.com/2013/08/want-to-share-your-code/)


### General instructions to install R packages

Most R packages are maintained in CRAN, Bioconductor or Github. To
install the package from each, use:

```{r generalinstall, message=FALSE, warning=FALSE, eval=FALSE}
# Installing from Bioconductor
source("http://www.bioconductor.org/biocLite.R")
biocLite("MASS")

# Installing from CRAN
install.packages("sorvi")

# Installing from Github
library(devtools)
install_github("antagomir/netresponse")
```


