---
title: "Linear models and probabilistic programming"
author: "Leo Lahti and Ville Laitinen | VIB / KU Leuven, Belgium - Feb 27, 2018"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: true
    use_bookdown: false
    highlight: haddock
---
<!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{microbiome tutorial - rstanarm}
  %\usepackage[utf8]{inputenc}
  %\VignetteEncoding{UTF-8}  
-->


```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(devtools)
library(gridExtra)
library(dplyr)
library(reshape2)
library(tidyr)
library(cowplot)
library(ggplot2)

# Defaults
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path = "figs/")
knitr::opts_chunk$set(dev="CairoPNG")
knitr::opts_chunk$set(cache=FALSE)

# http://derekogle.com/fishR/2015-09-17-Figure-Table-Captions-in-Markdown
library(captioner)
tbls <- captioner(prefix="Table")
figs <- captioner(prefix="Fig.")
```


# Introduction

**Target group**: those who are willing to have their first encounters with Bayesian modeling and inference 

**Learning goals** This tutorial gets you started in probabilistic programming. The aim is to learn how basic linear models can be implemented and extended using probabilistic programming tools, and understanding the underlying motivations and challenges.

**Teaching method** This material provides pointers for experimentation and online resources. The material is intended to be used in combination with lectures and hands-on instruction. 

## Probabilistic versus frequentist models

Probabilistic programming provides alternatives to classical models. Often, the models are identical under particular modeling assumptions. The probabilistic framework provides expanded opportunities to modify the model structure, distributional assumptions, and prior information.

This is particularly useful in cases where analytical solutions are not available, and classical counterparts may not exist. When both alternatives are available, incorporation of prior information can help to avoid overfitting and deal with small sample sizes, adding robustness in probabilistic models.

In standard cases, the Bayesian and frequentist versions give mostly similar estimates for the model parameters. This is not always the case, however. An example is given in the [stan_lm](http://mc-stan.org/rstanarm/articles/lm.html) page. In that example, the use of Bayesian prior helps to avoid overfitting and finds a more accurate solution when the sample size is small compared to the model complexity.

Due to time limitations, comparisons to the classical/frequentist models are not considered in this workshop. Instead, we focus on the use and interpretation of the probabilistic version.


# The tools: rstanarm

The [rstanarm R package](http://mc-stan.org/rstanarm/) is based on the Stan probabilistic programming language, but uses syntax that is familiar to R users who are familiar with traditional linear models.

For a more comprehensive introduction, check the [rstanarm tutorial](http://mc-stan.org/rstanarm/articles/aov.html).

Let us load the necessary libraries in R:

```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(rstan)
library(rstanarm)
```

# Example data


You can use your own data set, ready-made R example data sets, or load the following example data. This data set from [Lahti et al. Nat. Comm. 5:4344, 2014](http://www.nature.com/ncomms/2014/140708/ncomms5344/full/ncomms5344.html) includes microbiota profiles for 130 genus-like taxa across 1006 normal western adults from [Data Dryad](http://doi.org/10.5061/dryad.pk75d). Load the data in R:

```{r data, warning=FALSE, message=FALSE, eval=FALSE}
# Install the example data set used in rstanarm tutorials
install.packages("HSAUR3")

# Load example data
library(microbiome)
data(dietswap)

# Some necessary data preparations
sample_data(dietswap)$time <- sample_data(dietswap)$timepoint

# Pick jist the baseline data points (first time point 1)
x <- baseline(dietswap)

# Pick metadata from the phyloseq object
df <- meta(x)

# Add microbiome diversity to the metadata
df$diversity <- alpha(x, "shannon")$shannon
```

You can study the effects of the various parameters in this data.frame
on microbiota diversity using ANOVA, linear models, mixed models, and
generalized linear models in rstanarm.

For further example data sets from microbial ecology, see [microbiome tutorial](http://microbiome.github.io/microbiome/rstanarm.html).


# Exercise session 1: introduction to rstanarm

- Check the tutorial, and try out rstanarm examples with [ANOVA in rstanarm](http://mc-stan.org/rstanarm/articles/aov.html). 

- Explore the use of [shinystan R package](http://mc-stan.org/users/interfaces/shinystan). This provides tools to diagnose stan models visually and numerically, including model parameters and convergence diagnostics for MCMC simulations.

- Compare the results between rstanarm and standard alternatives (lm vs. stan_lm or aov vs. stan_aov) etc.

- If the time allows, you can also make rstanarm examples with [linear models in rstanarm](http://mc-stan.org/rstanarm/articles/lm.html). 




# Exercise session 2: extend and vary the linear model

- Read about and explore posterior predictive checking (PPcheck) in ShinyStan

- Try out alternative formulations of the [linear model](http://mc-stan.org/rstanarm/articles/lm.html) (stan_lm; stan_glm; stan_lmer)

- Increase the number of predictors in the linear model (instead of gender, use gender, nationality, and BMI, for instance) and investigate how this affects the inference

- [Read about priors in linear models](http://mc-stan.org/rstanarm/articles/lm.html). Experiment with different priors and see how this affects the results (gaussian, cauchy, uniform, different parameter values). You can also read [more on priors in rstanarm](http://mc-stan.org/rstanarm/articles/priors.html).



# Exercise session 3: further resources

- Compare alternative models with the [loo package](https://cran.r-project.org/web/packages/loo/vignettes/loo-example.html)

- Experiment with more complex models in the [rstanarm tutorials](http://mc-stan.org/rstanarm/articles/index.html)

- You can also try the full (more involved but also more flexible) [rstan implementation of the linear model](https://www.r-bloggers.com/bayesian-regression-with-stan-part-1-normal-regression/)

# Troubleshooting

Ask the instructors, or see the [rstanarm tutorial](http://mc-stan.org/rstanarm/articles/rstanarm.html) for tips on common problems that may arise when altering the model.



# Further Resources and  literature

[R users will now inevitably become Bayesians](https://thinkinator.com/2016/01/12/r-users-will-now-inevitably-become-bayesians/)

[Online manual for rstanarm](http://mc-stan.org/rstanarm/)

[Video tutorial on rstanarm](https://www.youtube.com/watch?v=z7zOzL9Rrzs)

[Stan tutorials](http://mc-stan.org/users/documentation/tutorials)



# Contact and acknowledgement

[Feedback form](https://docs.google.com/forms/d/e/1FAIpQLSdMegb7DaxXz7gaSTd2CiOlSW07EsjMZ-4zCKKi3ZEn8wYzOg/viewform?usp=sf_link)

Lecture slides [PDF](https://github.com/antagomir/antagomir.github.io/blob/master/files/publications/slides/20180226-rstan-VIB.pdf); [OpenOffice](https://github.com/antagomir/antagomir.github.io/raw/master/files/publications/slides/20180226-rstan-VIB.odp)

Regarding the course content, you can contact [Leo Lahti](http://www.iki.fi/Leo.Lahti).

We are grateful for the developers of rstan, rstanarm, and shinystan. Regarding the external tutorials used in this assignment, contact the respective authors.




