---
title: "Tutorial"
bibliography: 
- bibliography.bib
output: 
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---
<!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{microbiome tutorial - Installation}
  %\usepackage[utf8]{inputenc}
  %\VignetteEncoding{UTF-8}  
-->


This tutorial gets You started with R tools for microbial ecology. In particular, to provide an introduction to

 * R tools for microbial ecology
 * Role of custom data formats and tools in data analytical workflows
 * Reproducible document generation 
 * Possibilities and challenges in population-level microbiome profiling studies


The tutorial slides are available [here](https://github.com/antagomir/antagomir.github.io/blob/master/files/publications/slides/20170913-Leuven-Tutorial-Lahti.odp?raw=true)

### Installation

Launch R/RStudio (see [installation instructions](http://microbiome.github.io/microbiome/Installation.html))

To initiate reproducible documentation, do the following in RStudio:

 1. Open a new Rmarkdown (.html) file 
 1. Convert that .html file with the 'knit' button
 1. Modify the file and knit again to make your own reproducible report


### Example data: intestinal microbiota of 1006 Western adults

Example data set will be the [HITChip Atlas](Atlas.html), which is available via the microbiome R package in phyloseq format. This data set from [Lahti et al. Nat. Comm. 5:4344, 2014](http://www.nature.com/ncomms/2014/140708/ncomms5344/full/ncomms5344.html) comes with 130 genus-like taxonomic groups across 1006 western adults with no reported health complications. Some subjects have also short time series. Load the data in R with:

```{r atlasdata, message=FALSE, warning=FALSE}
# Data citation doi: 10.1038/ncomms5344
library(microbiome)
data(atlas1006) 
print(atlas1006)

# Rename the data
pseq <- atlas1006
```

### Phyloseq data stucture for taxonomic profiling

Explore the phyloseq data format. See [examples on microbiome data manipulation](Preprocessing.html).


### Diversity

Explore the estimation and analysis of various [diversity indices](Diversity.html) and [taxonomic composition](Composition.html).

```{r global, message=FALSE, warning=FALSE}
tab <- global(pseq, index = "all")
kable(head(tab))
```

Assign the estimated diversity to sample metadata
```{r global2, message=FALSE, warning=FALSE}
sample_data(pseq)$diversity <- tab$diversities_shannon
```

Visualize the data

```{r diversity, message=FALSE, warning=FALSE}
p <- plot_regression(diversity ~ age, meta(pseq)) + xlab("Age") + ylab("Diversity")
```


### Technical biases

Explore potential technical biases in the data. DNA extraction method has a remarkable effect on sample grouping.

```{r dnaextraction, message=FALSE, warning=FALSE, fig.width=12, fig.height=8}
# Use relative abundance data
ps <- microbiome::transform(pseq, "compositional")

# For this example, choose samples with DNA extraction information available
ps <- subset_samples(ps, !is.na(DNA_extraction_method))

# Or: you could focus on a single DNA extraction method
# ps <- subset_samples(ps, DNA_extraction_method == "r")

# Illustrate sample similarities with PCoA (NMDS)
plot_landscape(ps, "NMDS", "bray", col = "DNA_extraction_method")
```


### Core microbiota

[Core microbiota](Core.html) refers to the set of species shared by (almost) all individuals.


A full phyloseq object with just the core taxa is obtained as follows:

```{r core-data, message=FALSE, warning=FALSE}
# Transform to compositional abundances
pseq.rel <- microbiome::transform(pseq, "compositional")

# Pick the core (>0.1% relative abundance in >50% of the samples)
pseq.core <- core(pseq.rel, detection = 0.1/100, prevalence = 50/100)
```

Visualizing the core

```{r core-example3, fig.width=8, fig.heigth=18, fig.show='hold', warning=FALSE, out.width = '200px', warning=FALSE}
# Core with compositionals:
prevalences <- seq(.05, 1, .05)
detections <- 10^seq(log10(1e-3), log10(.2), length = 10)

p <- plot_core(pseq.rel, plot.type = "heatmap",
    prevalences = prevalences, detections = detections) +
    xlab("Detection Threshold (Relative Abundance)")
print(p)    
```


### Other tools

Explore further tools in [microbiome tutorial](http://microbiome.github.io/microbiome/).


