---
title: "ANCOM test for differential abundance"
bibliography: 
- bibliography.bib
- references.bib
output: 
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---
<!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{microbiome tutorial - comparisons}
  %\usepackage[utf8]{inputenc}
  %\VignetteEncoding{UTF-8}  
-->

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Handle citations
require(knitcitations)
cleanbib()
bib <- read.bibtex("bibliography.bib")
knitr::opts_chunk$set(fig.path = "figure/", dev="CairoPNG")

# CSS Style
library(knitr)
### Import the style from CSS file 'css_path' 
style <- paste(readLines("bioc.css"), collapse = "\n")
opts_knit$set(self.contained = TRUE,
              upload.fun = image_uri, ## optional
              header = c(highlight = style))
```

## ANCOM

[ANCOM](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4450248/) has been claimed to outperform zero-inflated Gaussians and other recently popular models of differential abundance in microbiome studies. An R package is [available](https://www.niehs.nih.gov/research/resources/software/biostatistics/ancom/index.cfm) but not from standard repositories. Meanwhile, the ANCOM implementation is here modified from `r citep("10.1186/s40168-017-0237-y")`. The R code was obtained from the first author and included with permission in the microbiome package. For reference to the original ANCOM method by `r citep("10.3402/mehd.v26.27663")`.

Load example data:

```{r ancom, message=FALSE}
# Load the example data
library(microbiome)
data(dietswap)
pseq <- dietswap
```

Calculate adjusted p-values on OTU abundance for the nationality:

```{r ancom2, message=FALSE}
padj <- ancom(pseq, "nationality")
print(names(which(padj < 0.05)))
```


For validation purposes, compare ANCOM and Negative binomial adjusted p-values. Note that the ANCOM p-values in this example just take values 0 (significant) and 1 (non-significant). Note that this is a toy example, demonstrating that the significance estimates are correlated between the two tests. 

```{r ancom3, message=FALSE}
library(MASS)
pvs <- c()
for (tax in taxa(pseq)) { 
  # Pick the signal (abundance) for this tax
  sample_data(pseq)$signal <- get_sample(pseq, tax)

  # Negative binomial test with group and gender included
  if (length(unique(sample_data(pseq)$signal)) > 1) {
    res <- glm.nb(signal ~ nationality, data = meta(pseq))

    # Show the results
    pvs[[tax]] <- anova(res)["nationality", "Pr(>Chi)"]
  } else {
    pvs[[tax]] <- 1
  }
}
pvs <- p.adjust(pvs)
pvs <- pvs[names(padj)]
boxplot(log10(pvs)  ~ padj, ylab = "Adj P (neg. binomial log10)", xlab = "Adj P (ANCOM)", main = "ANCOM vs. Negative binomial")
```

### References

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
bibliography()
```

