## Intestinal microbiota diversity in 1006 western adults

Let us investigate an example data set from [Lahti et al. Nat. Comm. 5:4344, 2014](http://www.nature.com/ncomms/2014/140708/ncomms5344/full/ncomms5344.html). This contains microbiota profiling of 130 genus-like taxa across 1006 normal western adults from [Data Dryad](http://doi.org/10.5061/dryad.pk75d).


### Download HITChip Atlas data

[Load the HITChip Atlas microbiome profiling data in R](Data.md)

```{r data2}
# Download the required R packages and then the HITChip Atlas data set
library("rdryad")
library("microbiome")
d <- download_microbiome("atlas1006")
```


### Diversity 

### Estimating microbial diversity with different diversity measures

```{r div-example, warning=FALSE, message=FALSE, fig.path = "figure/"}
library(phyloseq)
div <- estimate_diversity(d, measures = c("Observed", "Shannon", "Simpson"))
kable(div)
```


### Diversity vs. obesity

```{r div-example2, warning=FALSE, message=FALSE, fig.path = "figure/"}
p <- plot_diversity(d, x = "bmi_group", measures = c("Observed", "Shannon", "Simpson"), det.th = 250)
print(p)
```


### Diversity vs. age

```{r atlas-example3, fig.width=8, fig.height=5, message=FALSE, warning=FALSE, fig.path = "figure/"}
library(microbiome)
library(sorvi)

# Pick the subset of RBB-preprocessed samples from time point 0
dsub <- subset_samples(d, time == 0 & DNA_extraction_method == "r")

# Collect variables into a data frame
df <- harmonize_fields(sample_data(dsub))

# Visualize
p <- sorvi::regression_plot(diversity~age, df)
print(p)
```


## Further resources

For further examples, see [microbiome tutorial](https://github.com/microbiome/microbiome/blob/master/vignettes/vignette.md)