---
title: "HITChip Atlas examples"
author: "Leo Lahti and Gerben Hermes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    theme: united
    highlight: pygments
---
<!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{HITChip Atlas examples}
  %\usepackage[utf8]{inputenc}
-->


HITChip Atlas examples
===========

This tutorial provides step-by-step examples on exploratory analysis
of large-scale population-level microbiota profiling data.


### Install the experimental tools in R

```{r install2, eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("microbiome/microbiome")
```

### Loading the package

```{r loading, eval=TRUE}
library(microbiome)  
```


### HITChip Atlas data set

An example data set from [Lahti et al. Nat. Comm. 5:4344, 2014](http://www.nature.com/ncomms/2014/140708/ncomms5344/full/ncomms5344.html) contains large-scale profiling of 130 genus-like taxa across 1006 normal western subjects. This data set is readily available for download from the open [Data Dryad](http://doi.org/10.5061/dryad.pk75d) repository.



Load the HITChip Atlas microbiome profiling data in R.

```{r data}
# Load Dryad tools
library("rdryad") # Use the install.packages("rdryad") if package not available

# Define the data URL
url <- download_url('10255/dryad.64665')

# Download the data
data <- read.table(url, sep = "\t", row.names = 1, header = TRUE)

# Fix some broken names from the original release..
# ie. replace 'Clostridium..sensu.stricto.les' with 'Clostridiales'
colnames(data) <- gsub("Clostridium..sensu.stricto.les", "Clostridiales", colnames(data))
```

Load the HITChip Atlas metadata in R. Note that some individuals have
multiple time points.

```{r meta}
url <- download_url('10255/dryad.64666')
meta <- read.table(url, sep = "\t", row.names = 1, header = TRUE)

# Add SampleIDs as a separate column from rownames
meta$SampleID <- rownames(meta)
```

### Abundance histograms

```{r hist, fig.width=16, fig.heigth=8, fig.show='hold'}
# Load tools
library(dplyr)

# Pick 
# 1. All samples (all time points and DNA extraction methods)
# 2. Samples at time point 0 that have specific DNA extraction method 
all.samples <- meta$SampleID
rbb.samples <- filter(meta, Time == "0" & DNA_extraction_method == "r")$SampleID

# Visualize
tax <- "Prevotella.melaninogenica.et.rel."
d <- data[all.samples, tax]
plot(density(d), main = paste(tax, "(Abs.; all samples)"))
plot(density(log10(d)), main = paste(tax, "(Log10: all samples)"))

d <- data[rbb.samples, tax]
plot(density(d), main = paste(tax, "(Abs.; RBB samples)"))
plot(density(log10(d)), main = paste(tax, "(Log10: RBB samples)"))
```



### Licensing and Citations

This work can be freely used, modified and distributed under the 
[Two-clause FreeBSD license](http://en.wikipedia.org/wiki/BSD\_licenses).

Kindly cite the work as 'Leo Lahti and Gerben Hermes
(2015). HITChip Atlas tutorial. URL: http://microbiome.github.com'.


### Session info

This vignette was created with

```{r sessioninfo}
sessionInfo()
```




