---
title: "Atlas workflow"
author: "Leo Lahti"
date: "`r Sys.Date()`"
output:
  md_document:
    toc: false
    variant: markdown_github
  html_document:
    toc: false
---
<!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Project Template}
  %\usepackage[utf8]{inputenc}
-->


## Intestinal microbiota diversity in 1006 western adults

We investigate here an example data set from [Lahti et al. Nat. Comm. 5:4344, 2014](http://www.nature.com/ncomms/2014/140708/ncomms5344/full/ncomms5344.html) that contains large-scale profiling of 130 genus-like taxa across 1006 normal western subjects. This data set is available for download from the open [Data Dryad](http://doi.org/10.5061/dryad.pk75d) repository.


### Download HITChip Atlas data

[Load the HITChip Atlas microbiome profiling data in R](Data.md)

```{r data2}
# Download the required R packages and then the HITChip Atlas data set
library("rdryad")
library("microbiome")
d <- download_microbiome("atlas1006")
```


### Diversity 

```{r div-example, warning=FALSE, message=FALSE, fig.path = "figure/"}
# Estimat diversity using the vegan package
# NOTE: data needs to be in absolute scale, not logarithmic!
# Also pay attention on having samples on the rows (not on columns)
di <- vegan::diversity(atlas$microbes, index = "shannon")
hist(di, main = "Diversity")
```

### Compare with known background factors

Diversity vs. obesity

```{r diversitywithmetadata, warning=FALSE, message=FALSE, fig.width=9, fig.heigth=6, fig.path = "figure/"}
par(mar = c(6, 4, 3, 1))
bmi <- atlas$meta$BMI_group
di <- vegan::diversity(atlas$microbes)
boxplot(di ~ bmi, las = 2, main = "Microbiota diversity vs. obesity")
```


Diversity vs. age with smoothed confidence intervals:

```{r visu-example3, fig.width=8, fig.height=5, message=FALSE, warning=FALSE, fig.path = "figure/"}
library(microbiome)
library(sorvi)
library(dplyr)

# Pick the subset of RBB-preprocessed samples from time point 0
rbb.samples <- filter(atlas$meta, Time == 0 & DNA_extraction_method == "r")$SampleID

# Collect variables into a data frame
df <- data.frame(Age = atlas$meta[rbb.samples, "Age"], Diversity = di[rbb.samples])

# Visualize
p <- sorvi::regression_plot(Diversity~Age, df, shade = TRUE, mweight = TRUE, verbose = FALSE)
print(p)
```

