---
title: "HITChip Atlas examples"
author: "Leo Lahti and Gerben Hermes"
date: "`r Sys.Date()`"
output: 
  md_document:
    variant: markdown_github
---
<!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{HITChip Atlas examples}
  %\usepackage[utf8]{inputenc}
-->


HITChip Atlas examples
===========

This tutorial provides step-by-step examples on exploratory analysis
of large-scale population-level microbiota profiling data.


### HITChip Atlas data set

An example data set from [Lahti et al. Nat. Comm. 5:4344, 2014](http://www.nature.com/ncomms/2014/140708/ncomms5344/full/ncomms5344.html) contains large-scale profiling of 130 genus-like taxa across 1006 normal western subjects. This data set is readily available for download from the open [Data Dryad](http://doi.org/10.5061/dryad.pk75d) repository.

Load the HITChip Atlas microbiome profiling data in R.

```{r data}
# Load Dryad tools
library("rdryad") # Use the install.packages("rdryad") if package not available

# Define the data URL
url <- download_url('10255/dryad.64665')

# Download the data
data <- read.table(url, sep = "\t", row.names = 1, header = TRUE)

# Fix some broken names from the original release..
# ie. replace 'Clostridium..sensu.stricto.les' with 'Clostridiales'
colnames(data) <- gsub("Clostridium..sensu.stricto.les", "Clostridiales", colnames(data))
```

Load the HITChip Atlas metadata in R. Note that some individuals have
multiple time points.

```{r meta}
url <- download_url('10255/dryad.64666')
meta <- read.table(url, sep = "\t", row.names = 1, header = TRUE)

# Add SampleIDs as a separate column from rownames
meta$SampleID <- rownames(meta)

# Order BMI groups in correct order
# (see README at http://datadryad.org/resource/doi:10.5061/dryad.pk75d for details)
meta$BMI_group <- factor(meta$BMI_group, levels = c("underweight", "lean", "overweight", "obese", "severeobese", "morbidobese"))
meta$SubjectID <- factor(meta$SubjectID)
```

### Abundance histograms

Different sample sets have different population distributions in
microbial abundance. It is also important to consider whether to use
absolute or logarithmic abundances!

```{r hist, fig.width=16, fig.heigth=8}
# Load microbiome package
library(microbiome)  

# Load tools
library(dplyr)

# 1. List all samples (all time points and DNA extraction methods)
all.samples <- meta$SampleID

# 2. List samples at time point 0 that have specific DNA extraction method 
rbb.samples <- filter(meta, Time == "0" & DNA_extraction_method == "r")$SampleID

# Visualize
#tax <- "Prevotella.melaninogenica.et.rel."
tax <- "Bifidobacterium"
d <- data[all.samples, tax]
par(mfrow = c(1, 2))
plot(density(d), main = paste(tax, "(All samples)"), xlab = "Abundance (Absolute HITChip signal)")
plot(density(log10(d)), main = paste(tax, "(All samples)"), xlab = "Abundance (Log10 HITChip signal)")

d <- data[rbb.samples, tax]
par(mfrow = c(1, 2))
plot(density(d), main = paste(tax, "(RBB samples)"), xlab = "Abundance (Absolute HITChip signal)")
plot(density(log10(d)), main = paste(tax, "(RBB samples)"), xlab = "Abundance (Log10 HITChip signal)")
```


### Microbiota diversity 

Diversity takes into account species richness and evenness ie. how
species abundances are distributed. We use here Shannon diversity.

```{r diversity, warning=FALSE, message=FALSE}
# Diversity using the vegan package
# NOTE: data needs to be in absolute scale, not logarithmic
library(vegan)
di <- vegan::diversity(data, index = "shannon")

# Diversity histogram across all samples
hist(di, main = "Diversity histogram", xlab = "(Shannon) Diversity", ylab = "Population frequency")
```

Compare diversity with known background factors. By the way, we use
here all samples. 

```{r diversitywithmetadata, warning=FALSE, message=FALSE, fig.width=16, fig.heigth=8}
par(mar = c(6, 4, 3, 1), mfrow = c(1, 2))
boxplot(di ~ meta$BMI_group, las = 2, main = "Microbiota diversity vs. obesity")
plot(meta$Age, di, main = "Microbiota diversity vs. Age", ylab = "Diversity", xlab = "Age (years)")
```

TASK: Try to use just single sample per subject (time point 0) and
perhaps a single DNA extraction method (r) - see above ?

### Plotting trends

Plot subject age versus phylotype abundance with smoothed confidence intervals:

```{r visu-example3, fig.width=8, fig.height=8, message=FALSE, warning=FALSE}
library(microbiome)
library(sorvi)
df <- data.frame(Age = meta[rbb.samples, "Age"], Diversity = di[rbb.samples])
p <- sorvi::regression_plot(Diversity~Age, df, shade = TRUE, mweight = TRUE, verbose = FALSE)
print(p)
```


### Relative abundancies

Estimate relative abundance of the taxa in each sample. Note: the
input data set needs to be in absolute scale (not logarithmic).

```{r diversity-example6}
rel <- relative.abundance(t(data))

# Rearrange the data for ggplot visualization tools
library(reshape)
dfm <- melt(rel)
colnames(dfm) <- c("Taxon", "SampleID", "RelativeAbundance")

# Provide barplot visualizations of relative abundances for some randomly selected samples
library(ggplot2)
dfmf <- filter(dfm, SampleID %in% c("Sample-1", "Sample-2", "Sample-3", "Sample-4", "Sample-5"))
p <- ggplot(dfmf, aes(x = SampleID, y = RelativeAbundance, fill = Taxon))
p <- p + geom_bar(position = "stack", stat = "identity")
print(p)

# Also note that taking relative abundances likely changes the abundance histograms
theme_set(theme_bw(20))
p <- ggplot(filter(dfm, Taxon == "Prevotella.melaninogenica.et.rel."), aes(x = 100*RelativeAbundance))
p <- p + geom_density(fill = "darkgray")
p <- p + scale_x_log10()
p <- p + xlab("Relative Abundance (%)")
print(p)
```

### Principal component analysis (PCA)

Visualize deviation of all bacteria from their population mean (smaller: blue; higher: red):

```{r density, message=FALSE, warning=FALSE, fig.width=15, fig.height=8}
# Let us focus on the most abundant and prevalent bacteria
# that are seen in >1% (>0.01) relative abundance in 
# >20% (>0.2) of the subjects
prevalent.taxa <- names(which(prevalence(t(rel), 0.01, sort = TRUE) > 0.2))

# Project data on 2D display with PCA (visualize subjects based on 20 random features)
set.seed(423542)
proj <- microbiome::project.data(log10(data[, prevalent.taxa]), type = "PCA")

# Visualize
p <- densityplot(proj, col = meta$DNA_extraction_method, legend = T)
print(p)

# Now do the same with RBB extracted samples only
# Project data on 2D display with PCA (visualize subjects based on 20 random features)
set.seed(4235423)
proj <- microbiome::project.data(log10(data[rbb.samples, prevalent.taxa]), type = "PCA")

# Visualize with DNA extraction method (now all samples have the same DNA extraction)
p <- densityplot(proj, col = meta[rbb.samples, "DNA_extraction_method"], legend = T)
print(p)

# Visualize with low/high Prevotella
# This shows that Prevotella (color) has ecosystem-level impact on microbiota composition
#high.prevotella <- log10(data[rbb.samples, "Prevotella.melaninogenica.et.rel."]) > 4
prevotella.abundance  <- log10(data[rbb.samples, "Prevotella.melaninogenica.et.rel."]) 
p <- densityplot(proj, col = prevotella.abundance, legend = T)
print(p)
```


PCA with ggplot2 - the above example gives a shortcut for the following:

```{r density2, message=FALSE, warning=FALSE, fig.width=15, fig.height=8}
# Arrange projected data onto a data frame
coms <- intersect(rownames(proj), rownames(meta))
df <- as.data.frame(cbind(proj[coms,], meta[coms,]))
names(df) <- c("x", "y", colnames(meta))

# Construct the figure with ggplot2
library(ggplot2)
theme_set(theme_bw(15))
p <- ggplot(df) 

# Add densities
p <- p + stat_density2d(aes(x = x, y = y, fill=..density..), geom="raster", stat_params = list(contour = F), geom_params = list()) 
p <- p + scale_fill_gradient(low="white", high="black") 

# Add points
p <- p + geom_point(aes(x = x, y = y, color = Sex), size = 1.5) 

# Add labels
p <- p + xlab("PCA 1") + ylab("PCA 2") + ggtitle("Density plot")
p <- p + scale_colour_discrete(name = "Sex")

# Plot the figure
print(p)
```





### Licensing and Citations

This work can be freely used, modified and distributed under the 
[Two-clause FreeBSD license](http://en.wikipedia.org/wiki/BSD\_licenses).

Kindly cite the work as 'Leo Lahti and Gerben Hermes
(2015). HITChip Atlas tutorial. URL: http://microbiome.github.com'.


### Session info

This vignette was created with

```{r sessioninfo}
sessionInfo()
```




