---
title: "Beta diversity"
author: "Leo Lahti"
date: "`r Sys.Date()`"
bibliography: 
- bibliography.bib
- references.bib
output: 
  rmarkdown::html_vignette
---
<!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{microbiome tutorial - variability}
  %\usepackage[utf8]{inputenc}
  %\VignetteEncoding{UTF-8}  
-->



## Beta diversity (microbiome heterogeneity)

Load example data

```{r heterogeneity-example2, message=FALSE}
library(microbiome)
data(peerj32)
pseq <- peerj32$phyloseq
```

### Quantifying group heterogeneity / spread 

Beta diversity within a given sample set can be quantified as the average dissimilarity of each sample from the group mean. This was applied in group-level comparisons for instance in [Salonen et al. ISME J 2014](http://www.nature.com/ismej/journal/v8/n11/full/ismej201463a.html) (they focused on homogeneity using inverse correlation, whereas here we focus on heterogeneity using correlation but the measure is essentially the same). 

Calculate group diversities within the LGG (probiotic) and Placebo groups

```{r heterogeneity-example2bb, message=FALSE}
b.pla <- group_diversity(subset_samples(pseq, group == "Placebo"))
b.lgg <- group_diversity(subset_samples(pseq, group == "LGG"))
```

Use these to compare microbiota heterogeneity within each group. The LGG group tends to have smaller values, indicating that the samples are more similar to the group mean, and the LGG group is less heterogeneous (has smaller spread / is more homogeneous):

```{r heterogeneity-example2bbb, message=FALSE, out.width="300px"}
boxplot(list(LGG = b.lgg, Placebo = b.pla))
```

The **inter- and intra-invididual stability** (or homogeneity) measures are obtained as 1-b where b is the group diversity with the anticorrelation method ([Salonen et al. ISME J 2014](http://www.nature.com/ismej/journal/v8/n11/full/ismej201463a.html)). 



### Intra-individual heterogeneity 

Quantify beta diversity within subjects over time (as in [Salonen et al. ISME J 2014](http://www.nature.com/ismej/journal/v8/n11/full/ismej201463a.html) for intra-individual stability)

```{r homogeneity-example2c, message=FALSE, out.width="300px"}
betas <- list()
groups <- as.character(unique(meta(pseq)$group))
for (g in groups) {
  df <- meta(subset_samples(pseq, group == g))
  beta <- c()

  for (subj in df$subject) {
    # Pick the samples for this subject
    dfs <- subset(df, subject == subj)
    # Check that the subject has two time points
    if (nrow(dfs) == 2) {
      s <- as.character(dfs$sample)
      # Here with just two samples we can calculate the
      # beta diversity directly
      beta[[subj]] <- 1-cor(abundances(pseq)[, s[[1]]],
      		            abundances(pseq)[, s[[2]]],
			    method = "spearman")
    }
  }
  betas[[g]] <- beta
}

boxplot(betas)
```


### Beta diversity within individual over time

Calculate change in beta diversity (community dissimilarity) over time within a single individual

```{r homogeneity-example2d, message=FALSE, out.width="300px"}
data(atlas1006)
pseq <- atlas1006

# Identify subject with the longest time series (most time points)
s <- names(which.max(sapply(split(meta(pseq)$time, meta(pseq)$subject), function (x) {length(unique(x))})))

# Pick the metadata for this subject and sort the
# samples by time
library(dplyr)
df <- meta(subset_samples(pseq, subject == s)) %>% arrange(time)

# Calculate the beta diversity between each time point and
# the baseline (first) time point
beta <- c(0, 0) # Baseline similarity
s0 <- subset(df, time == 0)$sample
for (tp in df$time[-1]) {
  # Pick the samples for this subject
  # If the same time point has more than one sample,
  # pick one at random
  st <- sample(subset(df, time == tp)$sample, 1)
  a <- abundances(pseq)
  b <- 1 - cor(a[, s0], a[, st], method = "spearman")
  beta <- rbind(beta, c(tp, b))
}
colnames(beta) <- c("time", "beta")
beta <- as.data.frame(beta)

p <- ggplot(beta, aes(x = time, y = beta)) +
       geom_point() + geom_line()
print(p)       
```

