---
title: "Bimodality"
author: "Leo Lahti"
date: "`r Sys.Date()`"
bibliography: 
- bibliography.bib
- references.bib
output: 
  rmarkdown::html_vignette
---
<!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{microbiome tutorial - stability}
  %\usepackage[utf8]{inputenc}
  %\VignetteEncoding{UTF-8}  
-->


## Bimodality analysis

Get example data - [HITChip Atlas of 130 genus-like taxa across 1006 healthy western adults](http://www.nature.com/ncomms/2014/140708/ncomms5344/full/ncomms5344.html). A subset of 76 subjects have also short time series available for temporal stability analysis:

```{r bistability, message=FALSE}
# Load the example data
library(microbiome)
data(atlas1006)

# Rename the example data
pseq <- atlas1006

# Focus on specific DNA extraction method
pseq <- pseq %>% subset_samples(DNA_extraction_method == "r")

# Keep prevalent taxa (HITChip signal >3 in >20 percent of the samples)
pseq <- core(pseq, detection = 10^3, prevalence = 20)

# Use relative abundances
pseq <- transform(pseq, "compositional")

# For cross-sectional analysis, include
# only the zero time point:
pseq0 <- subset_samples(pseq, time == 0)
```


### Bimodality indicators

Bimodality of the abundance distribution provides an indirect
indicator of bistability, although other explanations such as sampling
biases etc. should be controlled. Multiple bimodality scores are
available.


Multimodality score using [potential analysis with
bootstrap](http://www.nature.com/ncomms/2014/140708/ncomms5344/full/ncomms5344.html)


```{r bimodality2, message=FALSE, warning=FALSE}
# Bimodality is better estimated from log10 abundances
pseq0.log10 <- transform(pseq0, "log10")
bimodality.pb <- bimodality(pseq0.log10, method = "potential_analysis")
```

Sarle's bimodality coefficient:

```{r bimodality3, message=FALSE, warning=FALSE}
bimodality.sarle <- bimodality(pseq0.log10, method = "Sarle.finite.sample")
```

DIP test is another standard multimodality test, available via the
[diptest](https://cran.r-project.org/web/packages/diptest/index.html)
package. Use the 1-p as the score here

```{r bimodality3bbb, message=FALSE, warning=FALSE}
library(diptest)
bimodality.dip <- apply(abundances(pseq0.log10), 1, function (x) {1 - unname(dip.test(x)$p.value)})
```

Compare the alternative bimodality scores

```{r bimodalitycomp, message=FALSE, warning=FALSE, out.width="400px"}
b <- cbind(DIP = bimodality.dip, Potential = bimodality.pb, Sarle = bimodality.sarle)
pairs(b)
```


### Visualize population densities for unimodal and bimodal groups

```{r stability2, message=FALSE, warning=FALSE, fig.width=12, fig.height=5, out.width="500px"}
# Pick the most and least bimodal taxa as examples
bimodality <- bimodality.pb
unimodal <- names(which.min(bimodality))
bimodal  <- names(which.max(bimodality))

# Visualize population frequencies
library(ggplot2)
theme_set(theme_bw(20))
p1 <- plot_density(pseq, variable = unimodal, log10 = TRUE) 
p2 <- plot_density(pseq, variable = bimodal,  log10 = TRUE) 
library(gridExtra)
library(ggplot2)
grid.arrange(p1, p2, nrow = 1)
```


## Tipping point detection

Identify potential minima in cross-section population data as tipping
point candidates (note that [longitudinal analysis](Stability.md)
would be necessary to establish bistability).

```{r stability-tipping, message=FALSE, warning=FALSE}
# Log10 abundance for a selected taxonomic group
tax <- bimodal

# Detect tipping points detection at log10 abundances 
x <- log10(abundances(pseq)[tax,])

# Potential analysis to identify potential minima
library(earlywarnings)
res <- livpotential_ews(x)

# Identify the potential minimum location as a tipping point candidate
# and cast the tipping back to the original (non-log) space:
tipping.point <- 10^res$min.points

print(tipping.point)
```


## Variation lineplot and bimodality hotplot

Pick subset of the [HITChip Atlas data set](http://doi.org/10.5061/dryad.pk75d) and plot the subject abundance variation lineplot (**Variation tip plot**) and **Bimodality hotplot** for a given taxon as in [Lahti et al. 2014](http://www.nature.com/ncomms/2014/140708/ncomms5344/full/ncomms5344.html). The bi-stable Dialister has bimodal population distribution and reduced temporal stability within subjects at intermediate abundances.

```{r stability-variationplot, message=FALSE, warning=FALSE, fig.show='hold', out.width="430px"}
# Bimodality hotplot:
# Consider a unique sample from each subject: the baseline time point 
p <- hotplot(pseq0, tax, tipping.point = tipping.point)
print(p)

pv <- tipplot(pseq, tax, tipping.point = tipping.point)
print(pv)
```


