## Microbiota composition


### Barplot visualizations

Also see [phyloseq barplot examples](http://joey711.github.io/phyloseq/plot_bar-examples.html) and [HITChip Barplots](Barplots.md)


Load example data:

```{r composition-example1, fig.path = "figure/", fig.width=8, fig.height=5}
# Example data
library(microbiome)
pseq0 <- download_microbiome("dietswap")
# Pick sample subset
pseq <- subset_samples(pseq0, group == "DI" & nationality == "AFR")
```

Show OTU absolute abundance in each sample. Plot absolute taxon
abundances (Note: on HITChip data the Phylum level is only
approximate):

```{r composition-example1b, fig.path = "figure/", fig.width=8, fig.height=5}
plot_abundance(pseq, taxonomic.level = "Phylum")
```

Same with relative abundances:

```{r composition-example3, fig.path = "figure/", fig.width=8, fig.height=5}
p <- plot_abundance(pseq, taxonomic.level = "Phylum", relative.abundance = TRUE)
p <- p + guides(fill = guide_legend(nrow = 12))
print(p)
```


Arrange by sample variable and use custom X axis labels. Americans have more Bacteroides, and Africans more Prevotella (as expected):

```{r composition-example4, fig.path = "figure/", fig.width=12, fig.height=5}
# Subset taxa and samples
pseq <- subset_samples(pseq0, group == "DI" & timepoint.within.group == 1)
pseq <- prune_taxa(c("Prevotella melaninogenica et rel.", "Bacteroides fragilis et rel.", "Akkermansia"), pseq)
p <- plot_abundance(pseq, relative.abundance = TRUE, sort.by = "nationality", x.label = "nationality")
p <- p + theme(legend.position = "bottom")
p <- p + guides(fill = guide_legend(ncol = 1))
print(p)
```

