## Microbiota composition


### Barplot visualizations

Also see [phyloseq barplot examples](http://joey711.github.io/phyloseq/plot_bar-examples.html) and [HITChip Barplots](Barplots.md)


Load example data:

```{r composition-example1, fig.path = "figure/", fig.width=8, fig.height=5}
# Example data
library(microbiome)
pseq0 <- download_microbiome("dietswap")
# Pick sample subset
pseq <- subset_samples(pseq0, group == "DI" & nationality == "AFR")
```

Show OTU absolute abundance in each sample. Plot absolute taxon
abundances (Note: on HITChip data the Phylum level is only
approximate):

```{r composition-example1b, fig.path = "figure/", fig.width=8, fig.height=5}
plot_composition(pseq, taxonomic.level = "Phylum")
```

Same with relative abundances:

```{r composition-example3, fig.path = "figure/", fig.width=8, fig.height=10}
p <- plot_composition(pseq, taxonomic.level = "Phylum", relative.abundance = TRUE)
p <- p + guides(fill = guide_legend(nrow = 12, byrow = TRUE))
p <- p + theme(legend.position = "bottom")
print(p)
```


Arrange by sample variable and use custom X axis labels. Africans have more Prevotella as expected:

```{r composition-example4, fig.width=12, fig.height=5}
# Subset taxa and samples
pseq <- subset_samples(pseq0, group == "DI" & timepoint.within.group == 1)
# Pick the top OTUs only
pseq <- prune_taxa(names(sort(taxa_sums(pseq), TRUE)[1:5]), pseq)
p <- plot_composition(pseq, relative.abundance = TRUE, sort.by = "nationality", x.label = "nationality")
p <- p + guides(fill = guide_legend(ncol = 1))
p <- p + theme(legend.position = "bottom")
print(p)
```

### Coloured Barplots

The following example visualizes samples, colored by Phylum
percentages (in this example data the Phylum is approximated by 16S
sequence similarity, not exactly Phylum):

```{r composition-example5, warning=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.path = "figure/"}
pseq <- subset_samples(pseq0, group == "DI")
p <- plot_bar(pseq, x = "timepoint.within.group", fill = "Phylum", facet_grid = ~nationality)
print(p)
```

