
## Group-wise comparisons

Read example data from a [diet swap study](http://dx.doi.org/10.1038/ncomms7342):

```{r comparisons-example, warning=FALSE, message=FALSE}
library(microbiome)
data("dietswap")
pseq <- dietswap
```

### Comparing of two or more groups with a parametric test (linear model; ANOVA)

Note that in practice it will be necessary to check ANOVA modeling assumptions before testing:

```{r comparisons2}
# Convert to relative abundances
pseq <- transform_phyloseq(pseq, "relative.abundance")

# 1-way ANOVA p-values for the multi-group comparison across time groups
anova.results <- check_anova(pseq, "group", p.adjust.method = "BH")
kable(head(anova.results))
```


Negative binomial test example ([read more](http://www.ats.ucla.edu/stat/r/dae/nbreg.htm)):

```{r comparisons2b, eval=FALSE}
library(MASS)
taxa <- taxa_names(x)[1:2]
x <- atlas1006
df <- as(sample_data(x), "data.frame")
for (tax in taxa) {
  df$signal <- get_sample(x, tax)
  res <- glm.nb(signal ~ bmi_group + gender, data = df)
  print(coef(summary(res)))
}
```

### Wilcoxon test (two-group comparisons)

If the data remarkably violates Gaussian assumptions use
non-parametric test. Wilcoxon is one option for two group
comparison. Here we compare males and females in the example data.

```{r comparisons-exampless}
pval <- check_wilcoxon(pseq, "sex")
```


### Comparison with random effect subject term

```{r comparisons-lmer, eval=FALSE}
pseq <- peerj32$phyloseq

# We need taxa x samples matrix
mydata <- otu_table(x)@.Data
if (!taxa_are_rows(x)) {mydata <- t(mydata)}

tax <- "Dialister"
dfs <- sample_data(x)
dfs$signal <- mydata[tax, rownames(dfs)]
dfs$group <- dfs[[group]]

# Paired comparison
library(lme4)
out <- lmer(signal ~ group + (1|subject), data = dfs)
out0 <- lmer(signal ~ (1|subject), data = dfs)
comp <- anova(out0, out)
pv <- comp[["Pr(>Chisq)"]][[2]]
```

### PERMANOVA

PERMANOVA is used to assess significance of community differences between groups. Here let us evaluate whether nationality has a significant effect on gut microbiota.

```{r comparisons-permanova, fig.width=5, fig.height=6, message=FALSE, warnings=FALSE}
# Example data
data("dietswap")
x <- dietswap
group <- "nationality"

# Use relative abundances for simpler visualizations
x <- transform_phyloseq(x, "relative.abundance")
otu <- get_sample(x)
meta <- as(sample_data(x), "data.frame")
meta$group <- meta[[group]]

# PERMANOVA: samples x species as input
library(vegan)
permanova <- adonis(t(otu) ~ group, data=meta, permutations=99)
pv <- as.data.frame(permanova$aov.tab)["group", "Pr(>F)"]

# P-value
print(pv)

# However note the assumption of similar
# multivariate spread among the groups
# ie. analogous to variance homogeneity
# Here the groups have signif. different spreads and
# permanova result may be explained by that.
dist <- vegdist(t(otu))
anova(betadisper(dist,meta$group))

# Coefs for the top taxa separating the groups
coef <- coefficients(permanova)["group1",]
top.coef <- coef[rev(order(abs(coef)))[1:20]]
par(mar = c(3, 14, 2, 1))
barplot(sort(top.coef), horiz = T, las = 1, main = "Top taxa")
```

