---
title: "Microbiome composition"
bibliography: 
- bibliography.bib
- references.bib
output: 
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---
  <!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{microbiome tutorial - composition}
  %\usepackage[utf8]{inputenc}
  %\VignetteEncoding{UTF-8}  
-->
  
  
## Microbiota composition
  

Also see [phyloseq barplot examples](http://joey711.github.io/phyloseq/plot_bar-examples.html).
  
Read example data from a [diet swap study](http://dx.doi.org/10.1038/ncomms7342):
  
```{r composition-example1}
# Example data
library(microbiome)
library(dplyr)
data(dietswap)

# Just use prevalent taxa to speed up examples
# (not absolute counts used in this example)
pseq <- core(dietswap, detection = 10^3, prevalence = 95/100)

# Pick sample subset
library(phyloseq)
pseq2 <- subset_samples(pseq, group == "DI" & nationality == "AFR" & timepoint.within.group == 1)
```

### Barplots for composition

Same with compositional (relative) abundances:
  
```{r composition-example4b, fig.width=12, fig.height=5, out.width="300px", fig.show="hold", eval=FALSE}
# Try another theme
# from https://github.com/hrbrmstr/hrbrthemes
library(hrbrthemes)
library(gcookbook)
library(tidyverse)

# Limit the analysis on core taxa and specific sample group
p <- plot_composition(pseq2,
		      taxonomic.level = "Genus",
                      sample.sort = "nationality",
                      x.label = "nationality",
                      transform = "compositional") +
     guides(fill = guide_legend(ncol = 1)) +
     scale_y_percent() +
     labs(x = "Samples", y = "Relative abundance (%)",
                                   title = "Relative abundance data",
                                   subtitle = "Subtitle",
                                   caption = "Caption here 'g'") + 
     theme_ipsum(grid="Y")
print(p)  
```


Averaged by group:
  
```{r composition-example4c, fig.width=12, fig.height=5, eval=FALSE}
p <- plot_composition(pseq2,
                      average_by = "nationality", transform = "compositional")
print(p)
```



### Composition heatmaps


Heatmap for CLR-transformed abundances, with samples and OTUs sorted with the neatmap method:
  
```{r composition-example7, fig.width=10, fig.height=3, eval = FALSE}
p <- plot_composition(pseq2, plot.type = "heatmap", transform = "clr"
                      sample.sort = "neatmap", otu.sort = "neatmap",
                      mar = c(6, 13, 1, 1))
```



### Plot taxa prevalence

Here we use the Dynamics IBD data set from [Halfvarson J., et al. Nature Microbiology, 2017](http://www.nature.com/articles/nmicrobiol20174) as downloaded from [Qiita ID 1629](https://qiita.ucsd.edu/study/description/1629). 

```{r, fig.height=6, fig.width=8}
data(DynamicsIBD)
p0 <- DynamicsIBD
# Improve the taxonomic information
p0.f <- format_phyloseq(p0)
# For the available taxonomic levels, see tax_table(p0.f)
p <- plot_taxa_prevalence(p0.f, 'Phylum')
print(p)
```

