---
title: "Comparisons"
author: "Leo Lahti"
date: "`r Sys.Date()`"
bibliography: 
- bibliography.bib
- references.bib
output: 
  rmarkdown::html_vignette
---
<!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{microbiome tutorial - comparisons}
  %\usepackage[utf8]{inputenc}
  %\VignetteEncoding{UTF-8}  
-->


## Group-wise comparisons

### Boxplots

```{r boxplot-example, warning=FALSE, message=FALSE, fig.width=8, fig.height=5}
# Load libraries
library(microbiome)
library(ggplot2)
library(dplyr)

# Probiotics intervention example data from
# https://peerj.com/articles/32/
data("peerj32")

# Abundance boxplot
p <- boxplot_abundance(peerj32$phyloseq, x = "time", y = "Akkermansia", line = "subject", color = "gender")
print(p)
```







### Negative binomial example

[Read more](http://www.ats.ucla.edu/stat/r/dae/nbreg.htm)

```{r comparisons2b, eval=FALSE}
library(MASS)
taxa <- taxa_names(x)[1:2]
x <- atlas1006
df <- as(sample_data(x), "data.frame")
for (tax in taxa) {
  df$signal <- get_sample(x, tax)
  res <- glm.nb(signal ~ bmi_group + gender, data = df)
  print(coef(summary(res)))
}
```


### Comparisons for individual taxa with random effect subject term

```{r comparisons-lmer, eval=FALSE}
# Get taxa x samples abundance matrix
x <- peerj32$phyloseq

# Get the data
mydata <- get_taxa(x)
tax <- "Dialister"
dfs <- sample_data(x)
dfs$signal <- mydata[tax, rownames(dfs)]
dfs$group <- dfs[[group]]

# Paired comparison
library(lme4)
out <- lmer(signal ~ group + (1|subject), data = dfs)
out0 <- lmer(signal ~ (1|subject), data = dfs)
comp <- anova(out0, out)
pv <- comp[["Pr(>Chisq)"]][[2]]
```


## Linear models with limma

Identify most significantly different taxa between males and females.

For further details, see [limma
homepage](http://bioinf.wehi.edu.au/limma/) and [limma User's
guide](http://www.lcg.unam.mx/~lcollado/R/resources/limma-usersguide.pdf). For
discussion on why limma is preferred over t-test, see [this
article](http://www.plosone.org/article/info:doi/10.1371/journal.pone.0012336).

```{r limma-example, warning=FALSE}
# Get example data
library(microbiome)
data("peerj32")
pseq <- peerj32$phyloseq
otu <- abundances(transform_phyloseq(pseq, "log10"))
meta <- sample_data(pseq)
grouping.variable <- "gender" 

# Compare the two groups with limma
library(limma)

# Prepare the design matrix which states the groups for each sample
# in the otu
design <- cbind(intercept = 1, Grp2vs1 = meta[[grouping.variable]])
rownames(design) <- rownames(meta)
design <- design[colnames(otu), ]

# NOTE: results and p-values are given for all groupings in the design matrix
# Now focus on the second grouping ie. pairwise comparison
coef.index <- 2
     
# Fit the limma model
fit <- lmFit(otu, design)
fit <- eBayes(fit)

# Limma P-values
pvalues.limma = fit$p.value[, 2]

# Limma effect sizes
efs.limma <-  fit$coefficients[, "Grp2vs1"]

# Summarise 
kable(topTable(fit, coef = coef.index, p.value=0.1), digits = 2)
```

**Q-Q plot for limma**

```{r limma-qq, warning=FALSE}
qqt(fit$t[, coef.index], df = fit$df.residual + fit$df.prior)
abline(0,1)
```

**Volcano plot for limma**

```{r limma-volcano, warning=FALSE}
volcanoplot(fit, coef = coef.index, highlight = coef.index)
```


### PERMANOVA

PERMANOVA can be also used to assess community-level differences
between groups. Here let us evaluate whether nationality has a
significant effect on gut microbiota.

```{r comparisons-permanova, fig.width=5, fig.height=6, message=FALSE, warnings=FALSE}
# Example data
data("dietswap")
x <- dietswap
group <- "nationality"

# Use relative abundances for simpler visualizations
x <- transform_phyloseq(x, "compositional")
otu <- get_sample(x)
meta <- as(sample_data(x), "data.frame")
meta$group <- meta[[group]]

# PERMANOVA: samples x species as input
library(vegan)
permanova <- adonis(t(otu) ~ group, data=meta, permutations=99, method = "bray")
pv <- as.data.frame(permanova$aov.tab)["group", "Pr(>F)"]

# P-value
print(pv)

# Note the assumption of similar
# multivariate spread among the groups
# ie. analogous to variance homogeneity
# Here the groups have signif. different spreads and
# permanova result may be explained by that.
dist <- vegdist(t(otu))
anova(betadisper(dist,meta$group))

# Coefs for the top taxa separating the groups
coef <- coefficients(permanova)["group1",]
top.coef <- coef[rev(order(abs(coef)))[1:20]]
par(mar = c(3, 14, 2, 1))
barplot(sort(top.coef), horiz = T, las = 1, main = "Top taxa")
```

