---
title: "Core"
author: "Leo Lahti"
date: "`r Sys.Date()`"
bibliography: 
- bibliography.bib
- references.bib
output: 
  rmarkdown::html_vignette
---
<!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{microbiome tutorial - core}
  %\usepackage[utf8]{inputenc}
  %\VignetteEncoding{UTF-8}  
-->

## Core microbiota analysis

See also related functions for the analysis of rare taxa.

Load example data:

```{r core-prevalence}
# Load data
library(microbiome)
data(peerj32)

# Rename the data
pseq <- peerj32$phyloseq

# Calculate compositional version of the data
# (relative abundances)
pseq.rel <- transform(pseq, "compositional")
```


### Prevalence of taxonomic groups

Relative population frequencies; at 1% compositional abundance threshold:

```{r core-prevalence2}
head(prevalence(pseq.rel, detection = 1, sort = TRUE))
```


Absolute population frequencies (sample count):

```{r core-prevalence2b}
head(prevalence(pseq.rel, detection = 1, sort = TRUE, count = TRUE))
```



### Core microbiota analysis

Core taxa above a given detection and prevalences:

```{r core-members, message=FALSE, warning=FALSE}
core.taxa <- taxa(core(pseq.rel, detection = 1, prevalence = 95))
```


Pick the core microbiota subset of the data. This corresponds to taxa that exceed the given prevalence and detection thresholds. Alternatively, use bootstrap analysis to identify the core members (for added robustness; see [Salonen et al. CMI (2012)](http://onlinelibrary.wiley.com/doi/10.1111/j.1469-0691.2012.03855.x/abstract). The core members may be sensitive to sampling effects; bootstrap aims to estimate how frequently a particular group is determined to be a core member in random bootstrap subsets of the data. The bootstrap provides more robust core but is slower to estimate.

```{r core-data, message=FALSE, warning=FALSE}
pseq.core <- core(pseq.rel, detection = 1, prevalence = 50)
pseq.core.bootstrap <- core(pseq.rel, detection = 1, prevalence = 50, method = "bootstrap")
```

### Core abundance and diversity

Total core abundance in each sample (sum of abundances of the core members):

```{r core-ab, message=FALSE, warning=FALSE}
core.abundance <- sample_sums(core(pseq.rel, detection = 1, prevalence = 95))
```


Total core and diversity (number of core members exceeding the detection limit in each samples):

```{r core-ab2, message=FALSE, warning=FALSE}
# Estimate all diversity measures at once
divs <- microbiome::diversity(pseq.core, detection = 1)

# Pick the desired richness / diversity indices
r <- divs$Observed # Observed richness (above the given detection threshold)
d <- divs$Shannon  # Shannon diversity
e <- divs$Evenness # Pielou's evenness
```





## Core visualization

### Core line plots

Determine core microbiota across various abundance/prevalence
thresholds with the blanket analysis [(Salonen et al. CMI, 2012)](http://onlinelibrary.wiley.com/doi/10.1111/j.1469-0691.2012.03855.x/abstract) based on various signal and prevalences.

```{r core-example2, fig.width=9, fig.heigth=6, fig.show='hold', out.width="430px", warning=FALSE}
# With absolute read counts
det <- c(0, 1, 2, 5, 10, 20, 50, 100, 200, 500, 1000, 2000)
prevalences <- seq(5, 100, 5)
p <- plot_core(pseq, prevalences = prevalences, detections = det, plot.type = "lineplot")
p + xlab("Abundance (OTU read count)")

# With compositional (relative) abundances
det <- c(0, 0.1, 0.2, 0.5, 1, 2, 5, 10, 20)
p <- plot_core(pseq.rel, prevalences = prevalences, detections = det, plot.type = "lineplot")
p + xlab("Relative Abundance (%)")
```


### Core heatmaps

This visualization method has been used for instance in [Intestinal microbiome landscaping: Insight in community assemblage and implications for microbial modulation strategies](https://academic.oup.com/femsre/article/doi/10.1093/femsre/fuw045/2979411/Intestinal-microbiome-landscaping-insight-in#58802539). Shetty et al. _FEMS Microbiology Reviews_ fuw045, 2017.

```{r core-example3, fig.width=8, fig.heigth=18, fig.show='hold', out.width = '430px'}
# Core with compositionals:
prevalences <- seq(5, 100, 5)
detections <- 10^seq(log10(1e-3), log10(20), length = 20)

# Also define gray color palette
gray <- gray(seq(0,1,length=5))
p <- plot_core(pseq.rel, plot.type = "heatmap", colours = gray,
    prevalences = prevalences, detections = detections) +
    xlab("Detection Threshold (Relative Abundance (%))")
print(p)    

# Core with absolute counts and horizontal view:
# and minimum population prevalence (given as percentage)
detections <- 10^seq(log10(1), log10(max(abundances(pseq))/10), length = 20)

library(RColorBrewer)
plot_core(pseq, plot.type = "heatmap", 
       		 prevalences = prevalences,
       		 detections = detections,
		 colours = rev(brewer.pal(5, "Spectral")),
		 min.prevalence = 20, horizontal = TRUE)
```


