### Prevalence of taxonomic groups


```{r prevalence, fig.path = "figure/"}
# Load example data
library(microbiome)
data("peerj32")
pseq <- peerj32$phyloseq

# Calculate relative abundances
pseq.rel <- transform_phyloseq(pseq, "relative.abundance", "OTU")
```


List prevalence for each group at 1 percent relative abundance abundance threshold:

```{r prevalence2, fig.path = "figure/"}
head(prevalence(pseq.rel, detection.threshold = 1, sort = FALSE))
```


List the taxa that are present at the given detection threshold (1% relative abundance) at a given prevalence (80%) (fraction of the samples):

```{r prevalence3, fig.path = "figure/"}
prevalent.taxa <- prevalent_taxa(pseq.rel, detection.threshold = 1, prevalence.threshold = 80)
```


### Core microbiota

Determine core microbiota with the [blanket
analysis](http://onlinelibrary.wiley.com/doi/10.1111/j.1469-0691.2012.03855.x/abstract)
based on various signal and prevalence thresholds.
 
```{r core-example, message=FALSE, warning=FALSE, fig.path = "figure/"}
det <- c(0, 0.1, 0.2, 0.5, 1, 2, 5, 10, 20)
prev <- seq(10, 100, 10)
core <- core_matrix(pseq.rel, prevalence.intervals = prev, detection.thresholds = det)
```

### Core 2D line plots

```{r core-example2, fig.width=8, fig.heigth=6, fig.path = "figure/", fig.show='hold', out.width="430px"}
# Core lineplot with absolute read counts
det <- c(0, 1, 2, 5, 10, 20, 50, 100, 200, 500, 1000, 2000, 5000, 1e4)
res <- plot_core(pseq, prevalence.intervals = prev, detection.thresholds = det, plot.type = "lineplot", plot = FALSE)
res$plot + xlab("Abundance (OTU read count)")

# Core lineplot with relative abundances
det <- c(0, 0.1, 0.2, 0.5, 1, 2, 5, 10, 20)
res <- plot_core(pseq.rel, prevalence.intervals = prev, detection.thresholds = det, plot.type = "lineplot", plot = FALSE)
res$plot + xlab("Relative Abundance (%)")

# Retrieve the core count data matrix
coremat <- res$data
print(coremat)
```


### Core heatmaps

Core with relative abundances:

```{r core-example3, fig.width=8, fig.heigth=13, fig.path = "figure/"}
prevalence.intervals <- seq(5, 100, 5)
detection.thresholds <- 10^seq(log10(1e-3), log10(20), length = 20)		 
res <- plot_core(pseq.rel, plot.type = "heatmap", palette = "bw", prevalence.intervals = prevalence.intervals, detection.thresholds = detection.thresholds, plot = FALSE) 
print(res$plot + xlab("Detection Threshold (Relative Abundance (%))"))


# Retrieve the core prevalence data matrix
prevalences <- res$data
```

Core with absolute counts:

```{r core-example3b, fig.width=8, fig.heigth=13, fig.path = "figure/"}
prevalence.intervals = seq(5, 100, 5)
detection.thresholds <- 10^seq(log10(1), log10(max(otu_table(pseq))/10), length = 20)		 
res <- plot_core(pseq, plot.type = "heatmap", palette = "bw", prevalence.intervals = prevalence.intervals,
       		       detection.thresholds = detection.thresholds, min.prevalence = NULL)$plot
```

Zoom in on the core region by filtering out rows and columns not passing min prevalence (given as percentages):

```{r core-example3bb, fig.width=8, fig.heigth=13, fig.path = "figure/"}
res <- plot_core(pseq, plot.type = "heatmap", palette = "bw", prevalence.intervals = prevalence.intervals,
		detection.thresholds = detection.thresholds, min.prevalence = 10, plot = TRUE)
```


Color scale core plot:

```{r core-example4, fig.width=8, fig.heigth=13, fig.path = "figure/"}
res <- plot_core(pseq, plot.type = "heatmap", palette = "spectral", prevalence.intervals = prevalence.intervals,
		detection.thresholds = detection.thresholds, min.prevalence = 0, plot = TRUE)

# Also check the data used for plotting
kable(head(res$data))
```

