### Prevalence of taxonomic groups

```{r core-prevalence}
# Load example data
library(microbiome)
data("peerj32")
pseq <- peerj32$phyloseq

# Calculate relative abundances
pseq.rel <- transform_phyloseq(pseq, "relative.abundance", "OTU")
```


List prevalence for each group at 1 percent relative abundance abundance threshold:

```{r core-prevalence2}
head(prevalence(pseq.rel, detection.threshold = 1, sort = FALSE))
```


List the taxa that are present at the given detection threshold (1% relative abundance) at a given prevalence (80%) (fraction of the samples):

```{r core-prevalence3, fig.path = "figure/"}
prevalent.taxa <- prevalent_taxa(pseq.rel, detection.threshold = 1, prevalence.threshold = 80)
```


### Core microbiota

Determine core microbiota with the [blanket
analysis](http://onlinelibrary.wiley.com/doi/10.1111/j.1469-0691.2012.03855.x/abstract)
based on various signal and prevalence thresholds. See also the the
bootstrap_microbes function.
 
```{r core-example, message=FALSE, warning=FALSE, fig.path = "figure/"}
det <- c(0, 0.1, 0.2, 0.5, 1, 2, 5, 10, 20)
prev <- seq(10, 100, 10)
core <- core_matrix(pseq.rel, prevalence.intervals = prev, detection.thresholds = det)
```

A shortcut to determine the core taxa with a given detection and prevalence thresholds:

```{r core-members, message=FALSE, warning=FALSE}
core.taxa <- core(pseq.rel, detection.threshold = 1, prevalence.threshold = 95)
```

Total core abundance:

```{r core-members, message=FALSE, warning=FALSE}
total.core.abundance <- core_abundance(pseq.rel, detection.threshold = 1, prevalence.threshold = 95)
```



### Core 2D line plots

```{r core-example2, fig.width=8, fig.heigth=6, fig.show='hold', out.width="430px"}
# With absolute read counts
det <- c(0, 1, 2, 5, 10, 20, 50, 100, 200, 500, 1000, 2000, 5000, 1e4)
res <- plot_core(pseq, prevalence.intervals = prev, detection.thresholds = det, plot.type = "lineplot")
res$plot + xlab("Abundance (OTU read count)")

# With relative abundances
det <- c(0, 0.1, 0.2, 0.5, 1, 2, 5, 10, 20)
res <- plot_core(pseq.rel, prevalence.intervals = prev, detection.thresholds = det, plot.type = "lineplot")
res$plot + xlab("Relative Abundance (%)")

# Retrieve the core count data matrix
coremat <- res$data
print(coremat)
```


### Core heatmaps

```{r core-example3, fig.width=8, fig.heigth=15, fig.show='hold', out.width = '430px'}
# Core with relative abundances:
prevalence.intervals <- seq(5, 100, 5)
detection.thresholds <- 10^seq(log10(1e-3), log10(20), length = 20)
# Also define gray color palette
gray <- gray(seq(0,1,length=5))
res <- plot_core(pseq.rel, plot.type = "heatmap", colours = gray,
    prevalence.intervals = prevalence.intervals, detection.thresholds = detection.thresholds) 
print(res$plot + xlab("Detection Threshold (Relative Abundance (%))"))

res <- plot_core(pseq.rel, plot.type = "heatmap", colours = gray, 
    prevalence.intervals = prevalence.intervals, detection.thresholds = detection.thresholds) 

# Core with absolute counts:
prevalence.intervals = seq(5, 100, 5)
detection.thresholds <- 10^seq(log10(1), log10(max(otu_table(pseq))/10), length = 20)		 
plot_core(pseq, plot.type = "heatmap", colours = gray,
       		 prevalence.intervals = prevalence.intervals,
       		 detection.thresholds = detection.thresholds,
		 min.prevalence = NULL)$plot
```


Zoom in on the core region by filtering out rows and columns not passing min prevalence (given as percentages):

```{r core-example3bb, fig.width=8, fig.heigth=15, fig.path = "figure/", fig.show='hold', out.width='430px'}
res <- plot_core(pseq, plot.type = "heatmap", colours = gray,
                 prevalence.intervals = prevalence.intervals,
		 detection.thresholds = detection.thresholds,
		 min.prevalence = 10)
print(res$plot)		 


library(RColorBrewer)
res <- plot_core(pseq, plot.type = "heatmap",
                 #colours = colorRampPalette(rev(brewer.pal(11, "Spectral")))(5),
		 colours = brewer.pal(5, "Spectral"),
                 prevalence.intervals = prevalence.intervals,
		 detection.thresholds = detection.thresholds,
		 min.prevalence = 0)		 
print(res$plot)		 
```


Retrieve the core prevalence data matrix

```{r core-example4, fig.width=8, fig.heigth=13, fig.path = "figure/"}
prevalences <- res$data
kable(head(prevalences))
```

