## Visualizing cross-hybridization

To visualize cross-hybridization between selected taxa on HITChip (or
other chips), use the following.

By default the groups with no cross-hyb are filtered out for clarity. Rows and columns are ordered by hierarchical clustering. The cross-hyb is shown in percentages, rounded as indicated by the rounding argument. The percentage indicates which fraction of probes for a row taxon overlaps with probes of a column taxon. This is not symmetric if the row and col taxon have a different total number of probes. For details, see help(PlotCrosshyb).

```{r chyb, warning=FALSE}
library(microbiome, quietly = TRUE)

# Check cross-hyb between all L1 groups
res <- PlotCrosshyb(tax.level = "L1", rounding = 1, show.plot = FALSE)
    
# Pick the crosshyb table and figure
crosshyb.table <- res$data
p <- res$plot

# Plot the figure    
print(p)

# Organize the Crosshyb table
s <- subset(res$data, crosshyb > 0)
s <- s[rev(order(s$crosshyb)),]
head(s)
#write.table(s, file = "crosshyb.tab", sep = "\t", quote = F, row.names = F)

```


### Further examples

Investigate species-species cross-hybridization within the Dialister L2 group

```{r chyb2, warning=FALSE}
# Pick the phylogeny which was used to summarize probes to species level
phylogeny.info <- GetPhylogeny("HITChip", "filtered") 

# Select species belonging to Dialister L2 group
mytaxa <- levelmap("Dialister", level.from = "L2", level.to = "species", phylogeny.info = phylogeny.info)[[1]]

# Check cross-hyb between Dialister species
res <- PlotCrosshyb(tax.level = "species", selected.taxa = mytaxa, rounding = 0, phylogeny.info = phylogeny.info)

# Check the cross-hyb data as well
head(res$data)
    
```

