---
title: "Data"
author: "Leo Lahti"
date: "`r Sys.Date()`"
bibliography: 
- bibliography.bib
- references.bib
output: 
  rmarkdown::html_vignette
---
<!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{microbiome tutorial - data}
  %\usepackage[utf8]{inputenc}
  %\VignetteEncoding{UTF-8}  
-->


## Microbiome example data sets



### Intestinal microbiota profiling of 1000 Western adults

[The HITChip Atlas](Atlas.md) data set is available via the microbiome
R package in phyloseq format, and also in [Data
Dryad](http://doi.org/10.5061/dryad.pk75d). This data set from [Lahti
et al. Nat. Comm. 5:4344,
2014](http://www.nature.com/ncomms/2014/140708/ncomms5344/full/ncomms5344.html)
comes with 130 genus-like taxonomic groups across 1006 western adults
with no reported health complications. Some subjects have also short
time series. Load data in R with:

```{r atlasdata}
library(microbiome)
data("atlas1006") 
pseq <- atlas1006
```


### Diet swap between Rural and Western populations

A two-week diet swap study between western (USA) and traditional
(rural Africa) diets, reported in [O'Keefe et al. Nat. Comm. 6:6342,
2015](http://dx.doi.org/10.1038/ncomms7342). The data is also
available for download from [Data
Dryad](http://dx.doi.org/10.5061/dryad.1mn1n). Load in R with:

```{r dietswap2}
data(dietswap) 
```


### Intestinal microbiota versus blood metabolites

Data set from [Lahti et al. PeerJ 1:e32,
2013](https://peerj.com/articles/32/) characterizes associations
between human intestinal microbiota and blood serum lipids. This data
set contains a list with the microbiome profiling data phyloseq
object, and additional data matrix of lipid species. Load the data
in R with:

```{r peerj2}
data(peerj32)
```


## Importing and preprocessing microbiome data in R

To import standard microbiome data formats (mothur, qiime etc.) into R
as phyloseq object, use the tools from the
[phyloseq](http://joey711.github.io/phyloseq/import-data) R package.

Examples on data preprocessing (filtering, subsetting etc.) are
available in the [preprocessing tutorial](Preprocessing.md).




