
### HITChip Atlas data set

An example data set from [Lahti et al. Nat. Comm. 5:4344, 2014](http://www.nature.com/ncomms/2014/140708/ncomms5344/full/ncomms5344.html) contains large-scale profiling of 130 genus-like taxa across 1006 normal western subjects. This data set is readily available for download from the open [Data Dryad](http://doi.org/10.5061/dryad.pk75d) repository.

Load the HITChip Atlas microbiome profiling data in R.

```{r data}
# Load Dryad tools
library("rdryad") # Use the install.packages("rdryad") if package not available

# Define the data URL
url <- download_url('10255/dryad.64665')

# Download the data
data <- read.table(url, sep = "\t", row.names = 1, header = TRUE)

# Fix some broken names from the original release..
# ie. replace 'Clostridium..sensu.stricto.les' with 'Clostridiales'
colnames(data) <- gsub("Clostridium..sensu.stricto.les", "Clostridiales", colnames(data))

# Convert to matrix 
data <- as.matrix(data)
```

Load the HITChip Atlas metadata in R. Note that some individuals have
multiple time points.

```{r meta}
url <- download_url('10255/dryad.64666')
meta <- read.table(url, sep = "\t", row.names = 1, header = TRUE)

# Add SampleIDs as a separate column from rownames
meta$SampleID <- rownames(meta)

# Order BMI groups in correct order
# (see README at http://datadryad.org/resource/doi:10.5061/dryad.pk75d for details)
meta$BMI_group <- factor(meta$BMI_group, levels = c("underweight", "lean", "overweight", "obese", "severeobese", "morbidobese"))
meta$SubjectID <- factor(meta$SubjectID)
```

Collect the atlas data and metadata into a single object:

```{r atlas}
atlas <- list(microbes = data, meta = meta)
```


### PeerJ example data set

An example data set from Lahti et al. [PeerJ 1:e32, 2013](https://peerj.com/articles/32/) concerns associations between human intestinal microbiota and blood serum lipids. Load the data in R:

```{r exdata}
library(microbiome)
data(peerj32)
names(peerj32)
```

### Load example data

Load simulated example data of the human gut microbiota. With HITChip,
[fRPA](http://www.computer.org/csdl/trans/tb/2011/01/ttb2011010217-abs.html)
is the recommended preprocessing method.

```{r datapeer, eval=TRUE, message=FALSE, warning=FALSE}
library(microbiome)
# Define data path (you can replace data.directory with your own path)
data.directory <- system.file("extdata", package = "microbiome")
print(data.directory)

# Read HITChip data matrix (genus-level (L2) log10 values)
level <- "L1"
method <- "frpa"
l1.data <- read.profiling(level = level, 
	     		       method = method, 
              		       data.dir = data.directory, 
	      	       	       log10 = TRUE)  

# Read HITChip probe level data (absolute values - no log10)
oligo.data <- read.profiling(level = "oligo", 
                             data.dir = data.directory, 
			     log10 = FALSE)  

# Probe-taxon mapping table
phylogeny.info <- read.profiling(level = "phylogeny.full", 
                           	 data.dir = data.directory)

# Phylogeny that is used to summarize the probes to phylotype/genus/phylum levels
phylogeny.info.filtered <- read.profiling(level = "phylogeny.filtered", 
                           	 data.dir = data.directory)
```


### Reading metadata

An easy way to provide sample metadata is to create a tab-separated metadata file. You can create the file in Excel and export it to tab-separated csv format. The standard (and self-explanatory) field names include 'sampleID', 'time', 'subjectID', 'group', 'gender', 'diet', 'age'. You can leave these out or include further fields. See this [example file](https://raw.github.com/microbiome/microbiome/master/inst/extdata/metadata.xls). Read the metadata with:

```{r metapeer, eval=TRUE, message=FALSE, warning=FALSE}
# Read simulated example metadata
library(gdata)
metadata.file <- paste(data.directory, "/metadata.xls", sep = "")
metadata <- read.xls(metadata.file, as.is = TRUE)
rownames(metadata) <- metadata$sampleID
```


### Estimating relative abundancies

Estimate relative abundance of the taxa in each sample. Note: the
input data set needs to be in absolute scale (not logarithmic).

```{r diversity-example6}
rel <- relative.abundance(oligo.data, det.th = min(na.omit(oligo.data)))
```

