## Example data sets

This page shows how to import HITChip data in R, how to convert HITChip data into phyloseq format, and how to load some published example data sets for microbiome analyses in R. For further microbiome data sets in phyloseq format, check [this](http://joey711.github.io/phyloseq/download-microbio.me.html).

For examples on preprocessing the data (filtering, subsetting etc.), see the [preprocessing tutorial](Preprocessing.md).


### HITChip Atlas data 

Data from [Lahti et al. Nat. Comm. 5:4344, 2014](http://www.nature.com/ncomms/2014/140708/ncomms5344/full/ncomms5344.html) contains large-scale profiling of 130 genus-like taxa across 1006 normal western adults. Some subjects have also short time series. This data set is available in [Data Dryad](http://doi.org/10.5061/dryad.pk75d). [Downloading the HITChip Atlas in R phyloseq format](Atlas.md):

```{r atlasdata}
library(microbiome)
data.atlas <- download_microbiome("atlas1006")
```


### Diet swap data set

Data from [O'Keefe et al. Nat. Comm. 6:6342, 2015](http://dx.doi.org/10.1038/ncomms7342) from [Data Dryad](http://dx.doi.org/10.5061/dryad.1mn1n). This is a two-week diet swap study between western (USA) and traditional (rural Africa) diets including microbiota profiling.

```{r dietswapdata}
library(microbiome)
data.dietswap <- download_microbiome("dietswap")
```


### Intestinal microbiota and blood serum lipid metabolites

Data from [Lahti et al. PeerJ 1:e32, 2013](https://peerj.com/articles/32/) characterizes associations between human intestinal microbiota and blood serum lipids. This data set is not readily provided in phyloseq format since it also contains additional data matrix of lipid species. Loading the data in R:

```{r peerjdata}
library(microbiome)
data.peerj32 <- download_microbiome("peerj32")
```


### Importing HITChip data

Importing HITChip data from data folder. With HITChip,
[fRPA](http://www.computer.org/csdl/trans/tb/2011/01/ttb2011010217-abs.html)
is the recommended preprocessing method. You can provide sample
metadata by adding new fields in the template metadata file your
HITChip data folder and exporting it again to tab-separated .tab
format. Some standard, self-explanatory field names include 'sample',
'time', 'subject', 'group', 'gender', 'diet', 'age'. You can leave
these out or include further fields. See this [example
file](https://raw.github.com/microbiome/microbiome/master/inst/extdata/meta.tab).


```{r datapeer, eval=TRUE, message=FALSE, warning=FALSE}
# Define example data path (replace here data.directory with your own path)
library(microbiome)
data.directory <- system.file("extdata", package = "microbiome")
print(data.directory)
```

## HITChip to phyloseq format

The [phyloseq](https://github.com/joey711/phyloseq) R package provides
many additional tools for microbiome analyses. See [phyloseq demo
page](http://joey711.github.io/phyloseq-demo/).

Import HITChip phylotype-level data in
[phyloseq](https://github.com/joey711/phyloseq) format (note: the
precalculated matrices are calculated with detection.threshold = 0):

```{r pseqconversion}
pseq <- read_hitchip(data.directory, method = "frpa")$pseq
```

Get higher taxonomic levels, use (on HITChip we use L1/L2 instead of Phylum/Genus):

```{r pseqconversion4}
pseq.L2 <- aggregate_taxa(pseq, level = "L2")
pseq.L1 <- aggregate_taxa(pseq, level = "L1")
```

Importing HITChip probe-level data and taxonomy from HITChip
output directory (these are not available in the phyloseq object):

```{r pseqconversion3}
probedata <- read_hitchip(data.directory, method = "frpa")$probedata
taxonomy.full <- read_hitchip(data.directory, method = "frpa")$taxonomy.full
```

Convert your own data into phyloseq format as follows:

```{r pseqconversion2, eval = FALSE}
# We need to choose the HITChip data level to be used in the analyses
# In this example use HITChip L2 data (note: this is in absolute scale)
res <- read_hitchip(method = "frpa", data.dir = data.directory)

# Species-level data matrix
otu <- otu_table(res$pseq)@.Data 

# Corresponding sample metadata
meta <- res$meta

# Taxonomy
taxonomy <- GetPhylogeny("HITChip", "filtered")
taxonomy <- unique(as.data.frame(taxonomy[, c("L1", "L2", "species")]))
rownames(taxonomy) <- as.vector(taxonomy[, "species"])

# Merging data matrices into phyloseq format:
pseq <- hitchip2physeq(t(otu), meta, taxonomy)
```


### Picking data from phyloseq  

Assuming your data is in the phyloseq format, many standard tools can directly operate on that data. If you need to pick specific data sets separately, you can mimic these examples.

Get example data in phyloseq format:

```{r pseqconversion22}
library(microbiome)
pseq <- download_microbiome("atlas1006")
```

Pick sample metadata:

```{r data21}
meta <- sample_data(pseq)
```

Pick taxonomy table

```{r data22}
tax.table <- tax_table(pseq)
```

Pick taxa abundance data matrix. In this example the OTU level corresponds to genus-like groups (the function name otu_table is somewhat misleading):

```{r data23}
abundance.table <- otu_table(pseq)@.Data
```

Aggregate the abundance matrix to higher-level taxa on HITChip:

```{r data24}
pseq2 <- aggregate_taxa(pseq, "Phylum") # Aggregate into phyloseq object
dat <- otu_table(pseq2)@.Data # Pick aggregated abundance table
```

