---
title: "Diversity"
bibliography: 
- bibliography.bib
- references.bib
output: 
  rmarkdown::html_vignette
---
<!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{microbiome tutorial - diversity}
  %\usepackage[utf8]{inputenc}
  %\VignetteEncoding{UTF-8}  
-->


## Alpha richness and diversity

See a separate page on [Beta diversity](Betadiversity.html).

Load example data:

```{r divdata, warning=FALSE, message=FALSE, fig.height=3, fig.width=5}
library(microbiome)
data(atlas1006)
pseq <- atlas1006
```


### Diversity estimation

```{r divest, warning=FALSE, message=FALSE, fig.height=3, fig.width=5}
# Estimate diversity (table with various diversity measures):
diversity <- microbiome::diversity(pseq, detection = 0)

# Show beginning of the diversity table
head(kable(diversity))
```



### Visualization

Show diversity:

```{r div-example2, warning=FALSE, message=FALSE, fig.width=8, fig.height=4}
library(ggplot2)
theme_set(theme_bw(20)) # Set bw color scheme
p <- ggplot(diversity, aes(x = Shannon)) + geom_histogram()
print(p)
```

Visualize diversity measures w.r.t. discrete variable (or check more generic [group-wise comparison tools](Comparisons.html))

```{r div-example2bb, warning=FALSE, message=FALSE, fig.width=8, fig.height=4}
p <- plot_diversity(pseq, "bmi_group", measures = c("Chao1", "Shannon"), indicate.subjects = FALSE)
print(p)
```

### Group-wise comparison

To indicate time as discrete variable, order it as a factor. If a
subject column is available in sample data the subjects will be
indicated by lines across the groups

```{r div-groupwise, warning=FALSE, message=FALSE, fig.width=8, fig.height=4}
# Visualize diversities across groups
p <- plot_diversity(pseq, "gender", measures = c("Chao1", "Shannon"), indicate.subject = TRUE)
print(p)
```


Diversity vs. continuous variable:

```{r diversity-example13, fig.width=8, fig.height=5, message=FALSE, warning=FALSE, dev="CairoPNG"}
library(dplyr)
pseq <- atlas1006

# Add diversity into sample metadata
sample_data(pseq)$diversity <- microbiome::diversity(pseq)$Shannon

# Visualize
p <- plot_regression(diversity ~ age, meta(pseq))
print(p)
```

