---
title: "Example"
author: "Leo Lahti"
date: "`r Sys.Date()`"
bibliography: 
- bibliography.bib
- references.bib
output: 
  rmarkdown::html_vignette
---
<!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{microbiome tutorial - example}
  %\usepackage[utf8]{inputenc}
  %\VignetteEncoding{UTF-8}  
-->


# Example document

This Rmarkdown document shows how to carry out some basic HITChip analysis with RStudio. [Download this file](https://raw.githubusercontent.com/microbiome/microbiome/master/index./Example..html), open it in RStudio, and then press the 'Knit HTML button'. This will generate and open a HTML file with some analysis with an example data set. You can then modify this file to use your own data (see below), or add new analyses.


## Update the microbiome package

Run this to make sure you have the latest version of the microbiome package:

```{r example-1, echo=TRUE, message=FALSE, cache=TRUE}
# Updating microbiome package
library(devtools)
install_github("microbiome/microbiome")
```

Load the tools and example data

```{r example-2, echo=TRUE, message=FALSE}
library(microbiome)
data("dietswap")
print(dietswap)
```

