## Heatmaps

See [Composition](Composition.md) page for phyloseq microbiota composition heatmaps. Load some example data:

```{r heatmap-example-data, fig.width=16, fig.height=5, warning=FALSE}
library(microbiome)
data("dietswap")
pseq <- dietswap
library(phyloseq)

# Pick data subset
bacteroidetes <- map_levels(from = "Phylum", to = "Genus",
                   data = tax_table(pseq))$Bacteroidetes
pseq2 <- prune_taxa(bacteroidetes, pseq)
pseq2 <- subset_samples(pseq2, group == "DI")
```


### Matrix heatmaps

Alternatively, pick abundance matrix separately and use matrix
visualization tools. Z-transforming OTUs ie. visualize deviation of
all bacteria from their population mean (smaller: blue; higher: red):

```{r heatmap-matvisu-example, fig.width=16, fig.height=5, fig.path = "figure/"}
# Z transform
pseqz <- transform_phyloseq(pseq2, "Z", "OTU")

# Pick OTU table
x <- taxa_abundances(pseqz)

# Plot heatmap
tmp <- plot_matrix(x, type = "twoway", mar = c(5, 14, 1, 1))
```


Find visually appealing order for rows and columns with Neatmap sorting:

```{r neatmap3, message=FALSE, warning=FALSE, fig.path = "figure/", fig.width=16, fig.height=5}
# Use the original phyloseq object for ordering as negative values are not allowed
data(peerj32)
x <- peerj32$microbes # Matrix
xz <- scale(x) # Z-transformed matrix

# Sorting rows (or columns) if just the sample order is needed rather than the matrix
sorted.rows <- neatsort(xz, "rows", method = "NMDS", distance = "euclidean") 

# Or just arrange the matrix directly
xo <- neat(xz, method = "NMDS", distance = "euclidean") # Sorted matrix
tmp <- plot_matrix(t(xo), type = "twoway", mar = c(5, 12, 1, 1))
```



### Cross-correlating data sets

Cross-correlate columns of two data sets from related to microbiome and blood serum lipids associations ([PeerJ 1:e32](https://peerj.com/articles/32/)).

The function returns correlations, raw p-values, and fdr estimates (not strictly proper as the comparisons are not independent). Here robust biweight midcorrelation ('bicor') from the [WGCNA package](http://labs.genetics.ucla.edu/horvath/CoexpressionNetwork/Rpackages/WGCNA/). Keep only those elements that have at least only one significant correlation (n.signif):

```{r heatmap-crosscorrelate4, message=FALSE, warning=FALSE, fig.keep='none', fig.path = "figure/"}
# Load example data 
library(microbiome)
data(peerj32)
data.peerj32.otu <- peerj32$microbes 
data.peerj32.lipids <- peerj32$lipids 

# Define data sets to cross-correlate
# OTU Log10 matrix # Microbiota (44 samples x 130 bacteria)
x <- log10(data.peerj32.otu)
y <- as.matrix(data.peerj32.lipids) # Lipids (44 samples x 389 lipids)

# Cross correlate data sets
correlations <- cross_correlate(x, y, method = "bicor", mode = "matrix", p.adj.threshold = 0.05, n.signif = 1)
```

Arrange the results in handy table format: 

```{r correlation-table2, message=FALSE, warning=FALSE, fig.path = "figure/"}
correlation.table <- cmat2table(correlations)
kable(head(correlation.table))
```

### Correlation heatmaps

Rearrange the data and plot the heatmap and mark significant correlations with stars to reproduce microbiota-lipidome heatmap from [this article](https://peerj.com/articles/32/) (the ordering of rows and columns may be different): 

```{r heatmap-example-stars2, message=FALSE, warning=FALSE, fig.keep='none', fig.path = "figure/"}
p <- correlation_heatmap(correlation.table, "X1", "X2", fill = "Correlation", star = "p.adj", p.adj.threshold = 0.05) 
```
```{r heatmap-example-stars3, fig.width=12, fig.height=8, message=FALSE, warning=FALSE, fig.path = "figure/"}
print(p)
```


### Heatmaps with ggplot2

The above examples provide handy shortcuts for heatmap visualization. You can also directly modify the ggplot2 routines. This time, let us set q-value threshold also for cell coloring: 

```{r heatmap-example-stars, message=FALSE, warning=FALSE, fig.path = "figure/"}
# Order the rows and columns with levels argument if needed:
correlation.table$X1 <- factor(correlation.table$X1, levels = unique(as.character(correlation.table$X1)))
correlation.table$X2 <- factor(correlation.table$X2, levels = unique(as.character(correlation.table$X2)))

# Set black-and-white theme
library(ggplot2)
theme_set(theme_bw())

# Pick only the correlations with q<0.05
# Note: this will leave other cells empty
library(dplyr)
subtable <- filter(correlation.table, p.adj < 0.05)

# Arrange the figure
p <- ggplot(subtable, aes(x = X1, y = X2, fill = Correlation))
p <- p + geom_tile() 
p <- p + scale_fill_gradientn("Correlation", 
       	 		       breaks = seq(from = -1, to = 1, by = 0.2), 
			       colours = c("darkblue", "blue", "white", "red", "darkred"), 
			       limits = c(-1,1)) 

# Polish texts
p <- p + theme(axis.text.x=element_text(angle = 90))
p <- p + xlab("") + ylab("")

# Mark the most significant cells with stars
p <- p + geom_text(data = subset(correlation.table, p.adj < 0.02), 
       	 	   aes(x = X1, y = X2, label = "+"), col = "white", size = 5)

# Plot
print(p)
```

### Heatmap with text

For detailed information, might be handy to print the actual values on
top of the heatmap:

```{r heatmap-example-text, message=FALSE, warning=FALSE, fig.width=14, fig.height=16, fig.path = "figure/"}
theme_set(theme_bw(20))
df <- microbiome::cmat2table(correlations)
df$X1 <- factor(df$X1)
df$X2 <- factor(df$X2)
p <- ggplot(df, aes(X1, X2, group=X2)) 
p <- p + geom_tile(aes(fill = Correlation)) 
p <- p + geom_text(aes(fill = df$Correlation, label = round(df$Correlation, 1)), size = 2) 
p <- p + scale_fill_gradientn("Correlation", 
       	 		      breaks = seq(from = -1, to = 1,  by = 0.25), 
       	 		      colours = c("blue", "white", "red"), 
			      limits = c(-1, 1))
p <- p + theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) 
p <- p + xlab("") + ylab("")
print(p)
```

### Links

For further examples, see [phyloseq tutorial](http://joey711.github.io/phyloseq/plot_heatmap-examples.html) and [Neatmaps](http://www.biomedcentral.com/1471-2105/11/45)