---
title: "HITChip Atlas Overview"
author: "Leo Lahti, Jarkko Salojarvi, Anne Salonen, Willem M de Vos"
date: "`r Sys.Date()`"
bibliography: 
- bibliography.bib
- references.bib
output: 
  md_document:
    variant: markdown_github
---
<!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{microbiome tutorial}
  %\usepackage[utf8]{inputenc}
-->

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Handle citations
require(knitcitations)
cleanbib()
#options("citation_format" = "pandoc")
bib <- read.bibtex("bibliography.bib")
#opts_chunk$set(fig.width=4, fig.height=3, par=TRUE, out.width='2in', fig.pos='H')
knitr::opts_chunk$set(fig.path = "figure/", dev="CairoPNG")
```


HITChip Atlas Overview
===========

We use the [phyloseq](http://joey711.github.io/phyloseq/import-data)
class, a standard representation format in R for taxonomic
profiling. This package provides extensions and convenient wrappers
for many standard tasks encountered in microbiome studies. 


## Intestinal microbiota diversity 

This data set has microbiota profiling of 130 genus-like taxa across over 10,000 samples.

Download the data in R:

```{r data2}
# Download the required R packages and then the HITChip Atlas data set
library("microbiome")
#data("atlas1006")
#pseq <- atlas1006
# Load the data (not yet public)
atlas = readRDS("/home/lei/proj/hitchip-atlas-annotation-scripts/Atlas.RData") # atlas
```


### Diversity vs. age

```{r atlas-example3, fig.width=8, fig.height=5, message=FALSE, warning=FALSE, fig.path = "figure/"}
# Pick the subset of RBB-preprocessed samples from time point 0
pseq <- subset_samples(atlas, time == 0 & DNA_extraction_method == "rbb")

# Visualize
p <- plot_regression(diversity~age, sample_data(pseq))
print(p)
```



### Licensing and Citations

This material can be freely used, modified and distributed under the
[Two-clause FreeBSD
license](http://en.wikipedia.org/wiki/BSD\_licenses). Kindly cite as
'Leo Lahti, Jarkko Salojarvi, Anne Salonen and Willem M de
Vos. HITChip Atlas. URL: http://microbiome.github.com'.


### References

```{r, echo=FALSE, message=FALSE}
#You can embed citations, for example: `r citep(bib[["lahti14natcomm"]])`
#You can embed citations, for example2: @lahti14natcomm
#Cite with DOI: `r citep("10.1890/11-0011.1")`
#Cite URL `r citep("http://knowledgeblog.org/greycite")`
#For automated markdown citations, check [this](http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html), [this](https://github.com/cboettig/knitcitations), and [this](http://www.carlboettiger.info/2012/03/24/citations-in-markdown-using-knitr.html). 
#write.bibtex(file="references.bib")
```

```{r, echo=FALSE, message=FALSE, results='asis'}
bibliography()
```

