### Installing R

If you do not already have R/RStudio installed, do the
following. 

  1. Install [R](http://www.r-project.org/) 
  1. Consider installing [RStudio](http://rstudio.org); GUI for R
  1. If you use Windows, install also [RTools](http://cran.r-project.org/bin/windows/Rtools/) (version corresponding to your R version)


### Install dependencies

Open R and install dependencies from the Tools panel, or run the
following commands:

```{r installation, message=FALSE, warning=FALSE, eval=FALSE}
source("http://www.bioconductor.org/biocLite.R")
biocLite(c("ade4", "fastcluster", "df2json", "rjson", "gplots", "devtools", 
  	   "ggplot2","MASS", "Matrix", "minet","mixOmics", "plyr","qvalue","reshape2",
	   "RPA","svDialogs","vegan","WGCNA", "rpart"))
```		   

If some of these installations fail, ensure from the RStudio tools
panel that you have access to CRAN and Bioconductor repositories. If
you cannot install some packages, some functionality in microbiome may
not work.


### Install/update the microbiome package

To install microbiome package and recommended dependencies, run in R:

```{r microbiomeinstall, message=FALSE, warning=FALSE, eval=FALSE}
library(devtools) # Load the devtools package
install_github("microbiome/microbiome") # Install the package
install_github("ropensci/rdryad") # Install proposed package
install_github("antagomir/netresponse") # Install proposed package
```

### Loading the package

Once the package has been installed, you can load it in R with:

```{r loading, eval=TRUE, message=FALSE}
library(microbiome)  
```


## Install HITChipDB package 

The HITChipDB package contains additional routines to fetch and
preprocess HITChip (or MIT/PITChip) data from the MySQL database. Note
that this package is not needed by most users and the data is
protected by password/IP combinations. Ask details from
admins. Install the package in R with:

```{r HITChipDBinstall, message=FALSE, warning=FALSE, eval=FALSE}
library(devtools) # Load the devtools package
install_github("microbiome/HITChipDB") # Install the package
# Also install RMySQL, multicore and tcltk !
source("http://www.bioconductor.org/biocLite.R")
biocLite("RMySQL") # multicore, tcltk?
# Test installation by loading the microbiome package in R
library("HITChipDB")
```

## General installation instructions

R packages are maintained in three distinct repositories: CRAN, Bioconductor and Github. You need to somehow find out which repository your desired package is in, or just try out the three alternatives:

```{r generalinstall, message=FALSE, warning=FALSE, eval=FALSE}
# Installing from Bioconductor
source("http://www.bioconductor.org/biocLite.R")
biocLite("MASS")

# Installing from CRAN
install.packages("sorvi")

# Installing from Github
library(devtools)
install_github("antagomir/netresponse")
```

