### Limma analysis

Example of limma analysis. Identify most significantly different taxa between males and females. For further details, see [limma homepage](http://bioinf.wehi.edu.au/limma/) and [limma User's guide](http://www.lcg.unam.mx/~lcollado/R/resources/limma-usersguide.pdf). For discussion on why limma is preferred over t-test, see [this article](http://www.plosone.org/article/info:doi/10.1371/journal.pone.0012336).

```{r limma-example, warning=FALSE, fig.path = "figure/"}
# Get example data
library(microbiome)
data("peerj32")
pseq <- peerj32$phyloseq
otu <- log10(otu_table(pseq)@.Data)
meta <- sample_data(pseq)
groups <- meta$gender

# Compare the two groups with limma
library(limma)

# Prepare the design matrix which states the groups for each sample
# in the otu
design <- cbind(intercept = 1, Grp2vs1 = groups)
rownames(design) <- rownames(meta)
design <- design[colnames(otu), ]

# NOTE: results and p-values are given for all groupings in the design matrix
# Now focus on the second grouping ie. pairwise comparison
coef.index <- 2
     
# Fit the limma model
fit <- lmFit(otu, design)
fit <- eBayes(fit)

# Summarise 
kable(topTable(fit, coef = coef.index), digits = 2)
```


Adjusted p-values; show all significant ones:

```{r limma-padj, warning=FALSE, fig.path = "figure/"}
pvalues.limma <- p.adjust(fit$p.value[, coef.index], method = "fdr")
names(pvalues.limma) <- rownames(fit$p.value)
print(sort(pvalues.limma[pvalues.limma < 0.1]))
```


### Q-Q plot


```{r limma-qq, warning=FALSE, fig.path = "figure/"}
qqt(fit$t[, coef.index], df = fit$df.residual + fit$df.prior)
abline(0,1)
```

### Volcano plot

```{r limma-volcano, warning=FALSE, fig.path = "figure/"}
volcanoplot(fit, coef = coef.index, highlight = coef.index)
```



### Comparison between limma and t-test

Order the taxa with t-test for comparison and validation purposes. The
differences are small in this simulated example, but [can be
considerable in real
data](http://www.plosone.org/article/info:doi/10.1371/journal.pone.0012336).

```{r limma-compairson, warning=FALSE, fig.path = "figure/"}
# Compare the two groups with t-test
library(dplyr)
pvalues.ttest <- c()
male.samples <- dplyr::filter(meta, gender == "male")$sample
female.samples <- dplyr::filter(meta, gender == "female")$sample
for (tax in rownames(otu)) {
  pvalues.ttest[[tax]] <- t.test(otu[tax, male.samples], otu[tax, female.samples])$p.value
}
# Multiple testing correction
pvalues.ttest <- p.adjust(pvalues.ttest, method = "fdr")

# Compare p-values between limma and t-test
taxa <- rownames(otu)
plot(pvalues.ttest[taxa], pvalues.limma[taxa])
abline(0,1,lty = 2)
```

### Continuous variables

Quantify continuous associations with lm_phyloseq function. This uses
the limma model to generate a table of P-values and effect sizes. Note
that no confounding variables taken into account in this wrapper. See
the [limma homepage](http://bioinf.wehi.edu.au/limma/) for more
detailed analyses.

```{r limma-lm-phyloseq, warning=FALSE, fig.path = "figure/"}
data("atlas1006")
tab <- lm_phyloseq(atlas1006, "age")
kable(head(tab))
```


