### Limma analysis

Example of limma analysis with simulated random data. For further details, see [limma homepage](http://bioinf.wehi.edu.au/limma/) and [limma User's guide](http://www.lcg.unam.mx/~lcollado/R/resources/limma-usersguide.pdf). For discussion on why limma is preferred over t-test, see [this article](http://www.plosone.org/article/info:doi/10.1371/journal.pone.0012336).


```{r limma-example, warning=FALSE}

# Get example data
library(microbiome, quietly = TRUE)

# Define here your own HITChip data folder
#data.directory <- system.file("extdata", package = "microbiome")
data.directory <- "~/R/x86_64-pc-linux-gnu-library/3.0/microbiome/extdata"

# Read HITChip data
hitchip.matrix <- read.profiling(level = "L2", 
                       data.dir = data.directory, log10 = TRUE)


# Define two random groups for demonstration purpose
g1 <- sample(colnames(hitchip.matrix), 10)
g2 <- setdiff(colnames(hitchip.matrix), g1)
# Modify hitchip matrix so that there are a few significant differences
altered.taxa <- sample(rownames(hitchip.matrix), 10)
hitchip.matrix[altered.taxa, g1] <- hitchip.matrix[altered.taxa, g1] + 2

# Compare the two groups with limma
library(limma)

# Prepare the design matrix which states the groups for each sample
# in the hitchip.matrix
design <- cbind(intercept=1, Grp2vs1=c(rep(0, length(g1)), rep(1, length(g2))))
rownames(design) <- c(g1, g2)
design <- design[colnames(hitchip.matrix), ]

# NOTE: results and p-values are given for all groupings in the design matrix
# Now focus on the second grouping ie. pairwise comparison
coef.index <- 2
     
# Fit the limma model
fit <- lmFit(hitchip.matrix, design)
fit <- eBayes(fit)

# Summarise or plot the results
topTable(fit, coef = coef.index)

# Q-Q plot
qqt(fit$t[, coef.index], df = fit$df.residual + fit$df.prior)
abline(0,1)

# Volcano plot
volcanoplot(fit, coef = coef.index, highlight = coef.index)

# Adjusted p-values; show all significant ones
pvalues.limma <- p.adjust(fit$p.value[, coef.index], method = "fdr")
names(pvalues.limma) <- unname(unlist(fit$genes))
print(sort(pvalues.limma[pvalues.limma < 0.05]))

```

### Comparison between limma and t-test

Order the taxa with t-test for comparison and validation purposes. The
differences are small in this simulated example, but [can be
considerable in real
data](http://www.plosone.org/article/info:doi/10.1371/journal.pone.0012336).

```{r limma-compairson, warning=FALSE}
# Compare the two groups with t-test
pvalues.ttest <- c()
for (tax in rownames(hitchip.matrix)) {
  pvalues.ttest[[tax]] <- t.test(hitchip.matrix[tax, g1], hitchip.matrix[tax, g2])$p.value
}
# Multiple testing correction
pvalues.ttest <- p.adjust(pvalues.ttest, method = "fdr")


# Order the taxa based on the p-values
taxa <- rownames(hitchip.matrix)
plot(pvalues.ttest[taxa], pvalues.limma[taxa])
abline(0,1,lty = 2)
```