---
title: "Taxonomic network visualization"
bibliography: 
- bibliography.bib
- references.bib
output: 
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---
<!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{microbiome tutorial - networks}
  %\usepackage[utf8]{inputenc}
  %\VignetteEncoding{UTF-8}  
-->




See also the [phyloseq tutorial](http://joey711.github.io/phyloseq/plot_network-examples)

Load [example data](Data.html):

```{r networks1, message=FALSE, warning=FALSE}
library(microbiome)
data(dietswap)
pseq <- dietswap
```


### Sample similarities

Network with phyloseq tools:

```{r networks3, warning=FALSE, message=FALSE}
theme_set(theme_bw(20))
p <- plot_net(pseq, maxdist = 0.2,
              shape = "group", color = "nationality",
	      distance = "bray", laymeth = "auto") +
     scale_colour_brewer(palette = "Accent")
print(p)		 
```


### Taxonomic network reconstruction 

The widely reported compositionality bias in similarity measures can
be fixed with SpiecEasi or SparCC; the implementations are available
via the [SpiecEasi package](https://github.com/zdk123/SpiecEasi). Note
that the execution is slow.

```{r networks4, warning=FALSE, message=FALSE, fig.width=10, fig.height=10}
library(SpiecEasi) #install_github("zdk123/SpiecEasi")
library(phyloseq)

# Pick the OTU table
otu <- t(abundances(pseq))

# SPIEC-EASI network reconstruction
# In practice, use more repetitions
net <- spiec.easi(otu, method='mb', lambda.min.ratio=1e-2, 
                  nlambda=20, icov.select.params=list(rep.num=20))

## Create graph object
n <- net$refit
colnames(n) <- rownames(n) <- colnames(otu)

# Network format
library(network)
netw <- network(as.matrix(n), directed = FALSE)
```


Investigate degree distribution with the following:

```{r degree, warning=FALSE, message=FALSE, fig.width=10, fig.height=7, eval=FALSE}
dd <- degree.distribution(ig)
plot(0:(length(dd)-1), dd, ylim=c(0,.35), type='b', 
      ylab="Frequency", xlab="Degree", main="Degree Distributions")
```


Visualize the network with [ggnet2](https://briatte.github.io/ggnet):

```{r networks5, warning=FALSE, message=FALSE, fig.width=12, fig.height=7}
library(GGally)
library(ggnet)
library(network)
library(sna)
library(ggplot2)
library(intergraph) # ggnet2 works also with igraph with this

phyla <- map_levels(colnames(otu), from = "Genus", to = "Phylum",
           tax_table(pseq))

netw %v% "Phylum" <- phyla
p <- ggnet2(netw, color = "Phylum", label = TRUE, label.size = 2)
print(p)
```


