## Ordination examples

Some examples with HITChip data. See also [phyloseq ordination tutorial](http://joey711.github.io/phyloseq/plot_ordination-examples.html).

Load example data:

```{r ordination1, message=FALSE, warning=FALSE, eval=TRUE}
library(microbiome)
library(phyloseq)
library(ggplot2)
data("atlas1006")
pseq <- atlas1006

# Convert signal to relative abundances
pseq.rel <- transform_phyloseq(pseq, "relative.abundance")

# Pick OTUs that are present with >1 percent relative abundance 
# in >10 percent of the samples
pseq2 <- filter_taxa(pseq.rel, function(x) sum(x > 1) > (0.1*nsamples(pseq.rel)), TRUE)
```


### Sample ordination

Project the samples with the given method and distance:

```{r ordination2, message=FALSE, warning=FALSE, fig.path = "figure/", results="hide"}
set.seed(423542)
library(phyloseq)
pseq.ord <- ordinate(pseq2, "NMDS", "bray")
# Just pick the projected data (first two columns + metadata)
proj <- plot_ordination(pseq2, pseq.ord, justDF = T)
```


Visualize and highlight. In addition to densityplot, see plot_ordn from the microbiome package and plot_ordination from the phyloseq package.

```{r ordination4, message=FALSE, warning=FALSE, fig.width=7, fig.height=5, fig.path = "figure/"}
# Highlight gender
library(microbiome)
p <- densityplot(proj[, 1:2], col = proj$gender, legend = T)
print(p)

# Highlight low/high Prevotella subjects
prevotella.abundance  <- as.vector(log10(otu_table(pseq2)["Prevotella melaninogenica et rel.",]) )
p <- densityplot(proj[, 1:2], col = prevotella.abundance, legend = T)
print(p)
```

Projection with sample names:

```{r visu-example2, message=FALSE, message=FALSE, fig.height=7, fig.width=5, fig.path = "figure/"}
ax1 <- names(proj)[[1]]
ax2 <- names(proj)[[2]]
ggplot(aes_string(x = ax1, y = ax2, label = "sample"), data = proj) + geom_text(size = 2)
```


Ordinate the taxa in NMDS plot with Bray-Curtis distances

```{r pca-ordination21, message=FALSE, message=FALSE, fig.height=6, fig.width=12}
p <- plot_ordination(pseq2, pseq.ord, type = "taxa", color = "Phylum", title = "Taxa ordination")
print(p)
```

Grouping the plots by Phylum

```{r pca-ordination22, message=FALSE, message=FALSE, fig.height=10, fig.width=13}
p + facet_wrap(~Phylum, 5)
```


### Multidimensional scaling (MDS / PCoA)

```{r ordinate23, warning=FALSE, message=FALSE, fig.width=10, fig.height=7}
plot_ordination(pseq, ordinate(pseq, "MDS"), color = "DNA_extraction_method") + geom_point(size = 5)
```

### RDA

See a separate page on [RDA](RDA.md).


### Canonical correspondence analysis (CCA)

With samples:

```{r ordinate24a, warning=FALSE, message=FALSE, fig.width=9, fig.height=5}
p <- plot_ordination(pseq, ordinate(pseq, "CCA"), type = "samples", color = "gender")
p + geom_point(size = 5)
```

With taxa:

```{r ordinate24b, warning=FALSE, message=FALSE, fig.width=10, fig.height=5}
p <- plot_ordination(pseq, ordinate(pseq, "CCA"), type = "taxa", color = "Phylum")
p <- p + geom_point(size = 4)
print(p)
```


### Split plot

```{r ordinate25, warning=FALSE, message=FALSE, fig.width=10, fig.height=5}
plot_ordination(pseq, ordinate(pseq, "CCA"), type = "split", shape = "gender", 
    color = "Phylum", label = "gender")
```


### Ordination biplot

```{r ordinate26, warning=FALSE, message=FALSE, fig.width=10, fig.height=5}
plot_ordination(pseq, ordinate(pseq, "CCA"), type = "biplot", color = "Phylum")
```






