---
title: "Ordination"
author: "Leo Lahti"
date: "`r Sys.Date()`"
bibliography: 
- bibliography.bib
- references.bib
output: 
  rmarkdown::html_vignette
---
<!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{microbiome tutorial - ordination}
  %\usepackage[utf8]{inputenc}
  %\VignetteEncoding{UTF-8}  
-->


## Ordination examples

Some examples with HITChip data. See also [phyloseq ordination tutorial](http://joey711.github.io/phyloseq/plot_ordination-examples.html).

Load example data:

```{r ordination1, message=FALSE, warning=FALSE, eval=TRUE}
library(microbiome)
library(phyloseq)
library(ggplot2)
data(dietswap)
pseq <- dietswap

# Convert signal to relative abundances
pseq.rel <- transform_phyloseq(pseq, "compositional")

# Pick OTUs that are present with >1 percent relative abundance 
# in >10 percent of the samples
pseq2 <- filter_taxa(pseq.rel,
           function(x) sum(x > 1) > (0.1*nsamples(pseq.rel)), TRUE)
```


### Sample ordination

Project the samples with the given method and distance. See also plot_ordination from the phyloseq package.


```{r ordination2, message=FALSE, warning=FALSE, results="hide"}
set.seed(423542)
pseq.ord <- ordinate(pseq2, "NMDS", "bray")
# Just pick the projected data (first two columns + metadata)
proj <- plot_ordination(pseq2, pseq.ord, justDF = T)
```

Then visualize the projected data:

```{r ordination4, message=FALSE, warning=FALSE, fig.width=7, fig.height=6, fig.show="hold", out.width="400px"}
# Highlighting nationality
p <- densityplot(as.matrix(proj[, 1:2]), col = proj$nationality, legend = T)
print(p)

# Projection with sample names:
ax1 <- names(proj)[[1]]
ax2 <- names(proj)[[2]]
p <- ggplot(aes_string(x = ax1, y = ax2, label = "sample"), data = proj) +
       geom_text(size = 2)
print(p)
```


Ordinate the taxa in NMDS plot with Bray-Curtis distances

```{r ordination-pca-ordination21, message=FALSE, message=FALSE, fig.height=5, fig.width=8}
p <- plot_ordination(pseq2, pseq.ord, type = "taxa", color = "Phylum", title = "Taxa ordination")
print(p)
```

Grouping by Phyla

```{r ordination-pca-ordination22, message=FALSE, message=FALSE, fig.height=10, fig.width=9}
p + facet_wrap(~Phylum, 5)
```


### Multidimensional scaling (MDS / PCoA)

```{r ordination-ordinate23, warning=FALSE, message=FALSE, fig.width=10, fig.height=6}
plot_ordination(pseq, ordinate(pseq, "MDS"), color = "DNA_extraction_method") +
                geom_point(size = 5)
```

### RDA

See a separate page on [RDA](RDA.md).


### Canonical correspondence analysis (CCA)


```{r ordination-ordinate24a, warning=FALSE, message=FALSE, fig.width=8, fig.height=6, fig.show="hold", out.width="400px"}
# With samples
p <- plot_ordination(pseq, ordinate(pseq, "CCA"),
       type = "samples", color = "nationality")
p <- p + geom_point(size = 4)
print(p)

# With taxa:
p <- plot_ordination(pseq, ordinate(pseq, "CCA"),
       type = "taxa", color = "Phylum")
p <- p + geom_point(size = 4)
print(p)
```


### Split plot

```{r ordination-ordinate25, warning=FALSE, message=FALSE, fig.width=14, fig.height=5}
plot_ordination(pseq, ordinate(pseq, "CCA"),
		      type = "split", shape = "nationality", 
    		      color = "Phylum", label = "nationality")
```


### Ordination biplot

```{r ordination-ordinate26, warning=FALSE, message=FALSE, fig.width=11, fig.height=5}
plot_ordination(pseq, ordinate(pseq, "CCA"), type = "biplot", color = "Phylum")
```






