## Ordination examples

Here some examples with HITChip data. See also [phyloseq ordination tutorial](joey711.github.io/phyloseq/plot_ordination-examples.html).

Load example data:

```{r ordination1, message=FALSE, warning=FALSE, eval=TRUE}
library(microbiome)
library(ggplot2)
pseq <- download_microbiome("atlas1006")

# Convert signal to relative abundances
pseq.rel <- transform_sample_counts(pseq, function (x) {x/sum(x)})

# Pick OTUs that are present with >1 percent relative abundance 
# in >10 percent of the samples
pseq2 <- filter_taxa(pseq.rel, function(x) sum(x > .01) > (0.1*nsamples(pseq.rel)), TRUE)
```


### Principal component analysis (PCA)

Project data on the first two PCA axes:

```{r ordination2, message=FALSE, warning=FALSE, fig.path = "figure/"}
set.seed(423542)
x <- t(otu_table(pseq2)@.Data)
proj <- project.data(log10(x), type = "PCA") # MDS.classical etc.
```


Visualize and highlight:

```{r ordination4, message=FALSE, warning=FALSE, fig.width=11, fig.height=8, fig.path = "figure/"}
# Highlight gender
p <- densityplot(proj, col = sample_data(pseq2)$gender, legend = T)
print(p)

# Highlight low/high Prevotella subjects
prevotella.abundance  <- log10(x[, "Prevotella melaninogenica et rel."]) 
p <- densityplot(proj, col = prevotella.abundance, legend = T)
print(p)
```

Projection with sample names:

```{r visu-example2, message=FALSE, message=FALSE, fig.height=8, fig.width=8, fig.path = "figure/"}
ggplot(aes(x = Comp.1, y = Comp.2, label = rownames(proj)), data = proj) + geom_text(size = 2)
```


### Unifrac with PCoA

Not implemented with HITChip but popular with sequencing data. See [phyloseq tutorial](http://joey711.github.io/phyloseq/plot_ordination-examples.html). 


### NMDS with Bray-Curtis distances

```{r ordination3, message=FALSE, warning=FALSE, fig.width=15, fig.height=8}
pseq.ord <- ordinate(pseq2, "NMDS", "bray")
```

Ordinate the taxa in NMDS plot with Bray-Curtis distances

```{r pca-ordination21, message=FALSE, message=FALSE, fig.height=8, fig.width=12}
p <- plot_ordination(pseq2, pseq.ord, type = "taxa", color = "Genus", title = "taxa")
print(p)
```

Grouping the plots by Phylum

```{r pca-ordination22, message=FALSE, message=FALSE, fig.height=8, fig.width=10}
p + facet_wrap(~Phylum, 5)
```


### Multidimensional scaling (MDS)

```{r ordinate23, warning=FALSE, message=FALSE, fig.width=10, fig.height=7}
plot_ordination(pseq, ordinate(pseq, "MDS"), color = "DNA_extraction_method") + geom_point(size = 5)
```


### Canonical correspondence analysis (CCA)

With samples:

```{r ordinate24a, warning=FALSE, message=FALSE, fig.width=9, fig.height=7}
p <- plot_ordination(pseq, ordinate(pseq, "CCA"), type = "samples", color = "gender")
p + geom_point(size = 5)
#p <- p + geom_polygon(aes(fill = gender))
```

With taxa:

```{r ordinate24b, warning=FALSE, message=FALSE, fig.width=10, fig.height=7}
p <- plot_ordination(pseq, ordinate(pseq, "CCA"), type = "taxa", color = "Phylum")
p <- p + geom_point(size = 4)
print(p)
```


### Split plot

```{r ordinate25, warning=FALSE, message=FALSE, fig.width=12, fig.height=7}
plot_ordination(pseq, ordinate(pseq, "CCA"), type = "split", shape = "gender", 
    color = "Phylum", label = "gender")
```


### Ordination biplot

```{r ordinate26, warning=FALSE, message=FALSE, fig.width=10, fig.height=7}
plot_ordination(pseq, ordinate(pseq, "CCA"), type = "biplot", color = "Phylum")
```






