### Principal component analysis (PCA)

Visualize deviation of all bacteria from their population mean (smaller: blue; higher: red):

```{r density, message=FALSE, warning=FALSE, fig.width=15, fig.height=8, fig.path = "figure/"}
# Example data
library(microbiome)
data(peerj32)
x <- peerj32$microbes
m <- peerj32$meta

# Convert HITChip signal to relative abundances
rel <- relative.abundance(t(x))

# Let us focus on the most abundant and prevalent bacteria
# that are seen in >1% (>0.01) relative abundance in 
# >20% (>0.2) of the subjects
prevalent.taxa <- names(which(prevalence(t(rel), 0.01, sort = TRUE) > 0.2))

# Project data on 2D display with PCA (visualize subjects based on 20 random features)
set.seed(423542)
proj <- microbiome::project.data(log10(x[, prevalent.taxa]), type = "PCA")

# Visualize; show gender with colors
p <- densityplot(proj, col = m$gender, legend = T)
print(p)

# Visualize with low/high Prevotella
# This shows that Prevotella (color) has ecosystem-level impact on microbiota composition
prevotella.abundance  <- log10(x[, "Prevotella melaninogenica et rel."]) 
p <- densityplot(proj, col = prevotella.abundance, legend = T)
print(p)
```


PCA with ggplot2 - the above example gives a shortcut for the following:

```{r density2, message=FALSE, warning=FALSE, fig.width=15, fig.height=8, fig.path = "figure/"}
# Arrange projected data onto a data frame
coms <- intersect(rownames(proj), rownames(m))
df <- as.data.frame(cbind(proj[coms,], m[coms,]))
names(df) <- c("x", "y", colnames(m))

# Construct the figure with ggplot2
library(ggplot2)
theme_set(theme_bw(15))
p <- ggplot(df) 

# Add densities
p <- p + stat_density2d(aes(x = x, y = y, fill=..density..), geom="raster", stat_params = list(contour = F), geom_params = list()) 
p <- p + scale_fill_gradient(low="white", high="black") 

# Add points
p <- p + geom_point(aes(x = x, y = y, color = gender), size = 1.5) 

# Add labels
p <- p + xlab("PCA 1") + ylab("PCA 2") + ggtitle("Density plot")
p <- p + scale_colour_discrete(name = "Sex")

# Plot the figure
print(p)
```


### Other projections

Project high-dimensional data on two-dimensional plane by various
methods including PCA, MDS, Sammons mapping etc. (for visualization
purposes; see help(project.data) for details):

```{r visu-example, message=FALSE, message=FALSE, fig.height=8, fig.width=8, fig.path = "figure/"}
# Example data
library(microbiome)
data(peerj32)
x <- peerj32$microbes

# Nonmetric MDS projection
proj <- project.data(x, type = "MDS.nonmetric")

# Plot the projection
library(ggplot2)
theme_set(theme_bw(15))
ggplot(aes(x = Comp.1, y = Comp.2), data = proj) + geom_point()
```

Plot the projection with sample names:

```{r visu-example2, message=FALSE, message=FALSE, fig.height=8, fig.width=8, fig.path = "figure/"}
ggplot(aes(x = Comp.1, y = Comp.2, label = rownames(proj)), data = proj) + geom_text()
```

