---
title: "Comparisons"
author: "Leo Lahti"
date: "`r Sys.Date()`"
bibliography: 
- bibliography.bib
- references.bib
output: 
  rmarkdown::html_vignette
---
<!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{microbiome tutorial - comparisons}
  %\usepackage[utf8]{inputenc}
  %\VignetteEncoding{UTF-8}  
-->


## PERMANOVA for community-level multivariate comparisons

Load example data:

```{r boxplot-example, warning=FALSE, message=FALSE, fig.width=8, fig.height=5}
# Load libraries
library(microbiome)
library(ggplot2)
library(dplyr)

# Probiotics intervention example data 
data(peerj32) # Source: https://peerj.com/articles/32/
pseq <- peerj32$phyloseq # Rename the example data
```


PERMANOVA can be used to assess multivariate community-level
differences between groups. Here let us evaluate whether probiotics treatment
has a significant effect on overall gut microbiota composition.

Pick relative abundances (compositional) and sample metadata 

```{r comparisons_permanova_data, fig.width=5, fig.height=6, message=FALSE, warnings=FALSE}
pseq.rel <- transform_phyloseq(pseq, "compositional")
otu <- abundances(pseq.rel)
meta <- meta(pseq.rel)
```

Visualize the population density and highlight sample groups

```{r comparisons_permanova_visu, fig.width=5, fig.height=6, message=FALSE, warnings=FALSE}
plot_landscape(pseq.rel, method = "NMDS", distance = "bray", col = "group", size = 3)
print(p)
```


Perform PERMANOVA: 

```{r comparisons_permanova_analyse, fig.width=5, fig.height=6, message=FALSE, warnings=FALSE}
# samples x species as input
library(vegan)
permanova <- adonis(t(otu) ~ group,
               data = meta, permutations=99, method = "bray")

# P-value
print(as.data.frame(permanova$aov.tab)["group", "Pr(>F)"])
```

Check that variance homogeneity assumptions hold (to ensure the reliability of the results):

```{r comparisons-permanova2, fig.width=5, fig.height=6, message=FALSE, warnings=FALSE}
# Note the assumption of similar multivariate spread among the groups
# ie. analogous to variance homogeneity
# Here the groups have signif. different spreads and
# permanova result may be potentially explained by that.
dist <- vegdist(t(otu))
anova(betadisper(dist, meta$group))

# Coefs for the top taxa separating the groups
coef <- coefficients(permanova)["group1",]
top.coef <- coef[rev(order(abs(coef)))[1:20]]
par(mar = c(3, 14, 2, 1))
barplot(sort(top.coef), horiz = T, las = 1, main = "Top taxa")
```

