# HITChip data analysis in phyloseq format

The [phyloseq](https://github.com/joey711/phyloseq) is an external high-quality R package that provides additional tools for microbiome data analysis. These examples show how to convert HITChip data into phyloseq format and perform some standard analyses. For more info on phyloseq tools, see [phyloseq demo page](http://joey711.github.io/phyloseq-demo/).

Download example data from [O'Keefe et al. Nat. Comm. 6:6342, 2015](http://dx.doi.org/10.1038/ncomms7342) from [Data Dryad](http://dx.doi.org/10.5061/dryad.1mn1n). This is a two-week diet swap study between western (USA) and traditional (rural Africa) diets including microbiota profiling:

```{r phyloseqdata, fig.width=10, fig.height=7, warning=FALSE, message=FALSE}
library(microbiome)
data.dietswap <- download_microbiome("dietswap")
```


## Filtering and pruning

```{r prune, warning=FALSE, message=FALSE}
f1 <- filterfun_sample(topp(0.1))
wh1 <- genefilter_sample(physeq, f1, A = round(0.5 * nsamples(physeq)))
ex2 <- prune_taxa(wh1, physeq)
r <- transform_sample_counts(physeq, function(x) x/sum(x))
f <- filter_taxa(r, function(x) var(x) > 1e-05, TRUE)
```

## Networks

[Networks](http://joey711.github.io/phyloseq/plot_network-examples)

```{r networks, warning=FALSE, message=FALSE}
plot_net(physeq, maxdist = 0.45, point_label = "group")

ig <- make_network(physeq, max.dist = 0.45)
plot_network(ig, physeq, color = "gender", shape = "group", line_weight = 0.4, label = NULL)
```



For more, see [phyloseq demo page](http://joey711.github.io/phyloseq-demo/phyloseq-demo.html)

```{r more}
#ntaxa(physeq)
#rank_names(physeq)
#sample_names(physeq)
#taxa_names(physeq)
#sample_variables(physeq)
#merge_phyloseq
#get_variable(physeq, sample_variables(physeq)[5])[1:10]
#get_taxa_unique(physeq, "Phylum")
#sample_sums(physeq)
#taxa_sums(physeq)
#get_taxa(physeq, sample_names(physeq)[5])
#get_sample(physeq, taxa_names(physeq)[5])
#plot_bar(ent10, "Genus", fill = "Genus", facet_grid = SeqTech ~ Enterotype)
#GP <- prune_taxa(taxa_sums(physeq) > 0, physeq)
#sample_data(GP)$human = ...
#eso <- rarefy_even_depth(physeq)
# UniFrac(physeq) # This requires tree that we do not have
#GP.chl = subset_taxa(GlobalPatterns, Phylum == "Chlamydiae")
```
