## Standard data processing operations

The high-quality [phyloseq package](http://joey711.github.io/phyloseq/) provides a complete set of tools for data subsetting, aggregating and filtering.

Download example data from [O'Keefe et al. Nat. Comm. 6:6342, 2015](http://dx.doi.org/10.1038/ncomms7342) from [Data Dryad](http://dx.doi.org/10.5061/dryad.1mn1n). This is a two-week diet swap study between western (USA) and traditional (rural Africa) diets including microbiota profiling:

```{r phyloseqdata, fig.width=10, fig.height=7, warning=FALSE, message=FALSE}
library(microbiome)
pseq <- download_microbiome("dietswap")
```


### Sample operations

Sample names and variables

```{r preprocess2b, warning=FALSE, message=FALSE}
head(sample_names(pseq))
```

Sample sums

```{r preprocess2c, warning=FALSE, message=FALSE}
head(sample_sums(pseq))
```

Abundance of a given species in each sample

```{r preprocess2d, warning=FALSE, message=FALSE}
head(get_sample(pseq, taxa_names(pseq)[1]))
```

Filter samples

```{r preprocess3, warning=FALSE, message=FALSE}
f1 <- filterfun_sample(topp(0.1))
taxa <- genefilter_sample(pseq, f1, A = round(0.5 * nsamples(pseq)))
```

### Estimating relative abundancies

Estimate relative abundance of the taxa in each sample. Note: the
input data set needs to be in absolute scale (not logarithmic).


Calculating relative abundances for a phyloseq object:

```{r data10}
pseq <- transform_sample_counts(pseq, function(x) x/sum(x))
```


Calculating relative abundance for standard abundance matrix:

```{r diversity-example6}
dat <- otu_table(pseq)@.Data
rel <- relative.abundance(dat, det.th = 0)
```


Pick samples by specific metadata fields

```{r data-subsetting}
pseq.subset <- subset_samples(pseq, nationality == "AFR")
```


### Variable operations

Sample variable names

```{r preprocess4, warning=FALSE, message=FALSE}
sample_variables(pseq)
```

Pick variable values for a given variable

```{r preprocess4b, warning=FALSE, message=FALSE}
head(get_variable(pseq, sample_variables(pseq)[1]))
# Assign fields to sample metadata
# sample_data(GP)$human <- ..
```

### Taxa operations


Number of taxa

```{r preprocess3e, warning=FALSE, message=FALSE}
ntaxa(pseq)
```


Names

```{r preprocess3f, warning=FALSE, message=FALSE}
ranks <- rank_names(pseq)
taxa <- taxa_names(pseq)
```


Prune taxa

```{r preprocess3b, warning=FALSE, message=FALSE}
taxa <- levelmap(NULL, "Phylum", "Genus", tax_table(pseq))$Bacteroidetes

# With given taxon names
ex2 <- prune_taxa(taxa, pseq)

# Taxa with positive sum across samples
ex3 <- prune_taxa(taxa_sums(pseq) > 0, pseq)
```

Subset taxa

```{r preprocess3c, warning=FALSE, message=FALSE}
pseq <- subset_taxa(pseq, Phylum == "Bacteroidetes")
```


Filter by user-specified function values (here variance)

```{r preprocess3d, warning=FALSE, message=FALSE}
f <- filter_taxa(pseq, function(x) var(x) > 1e-05, TRUE)
```


List unique phylum-level groups: 

```{r preprocess3g, warning=FALSE, message=FALSE}
head(get_taxa_unique(pseq, "Phylum"))
```

Pick detected taxa by sample name:

```{r preprocess3gg, warning=FALSE, message=FALSE}
samplename <- sample_names(pseq)[[1]]
tax.abundances <- get_taxa(pseq, samplename)
```


Taxa sums

```{r preprocess3h, warning=FALSE, message=FALSE}
head(taxa_sums(pseq))
```


### Merging operations for phyloseq objects


Aggregate OTUs to higher taxonomic levels, use (on HITChip we use L1/L2 instead of Phylum/Genus):

```{r preprocessing-aggregate-taxa}
pseq <- read_hitchip(data.directory, method = "frpa")$pseq
pseq.L2 <- aggregate_taxa(pseq, level = "L2")
```

Merging phyloseq objects

```{r preprocess-merge, warning=FALSE, message=FALSE, eval=FALSE}
merge_phyloseq(pseq1, pseq2)
```


### Rarification

```{r preprocess-rarify, warning=FALSE, message=FALSE}
pseq.rarified <- rarefy_even_depth(pseq)
```


