## Filtering and pruning

The external high-quality [phyloseq package](http://joey711.github.io/phyloseq/) provides a complete set of tools for data preprocessing and filtering purposes. Download example data from [O'Keefe et al. Nat. Comm. 6:6342, 2015](http://dx.doi.org/10.1038/ncomms7342) from [Data Dryad](http://dx.doi.org/10.5061/dryad.1mn1n). This is a two-week diet swap study between western (USA) and traditional (rural Africa) diets including microbiota profiling:

```{r phyloseqdata, fig.width=10, fig.height=7, warning=FALSE, message=FALSE}
library(microbiome)
physeq <- download_microbiome("dietswap")
```

### Sample operations

```{r preprocess2, warning=FALSE, message=FALSE}
# Filter samples
f1 <- filterfun_sample(topp(0.1))
wh1 <- genefilter_sample(physeq, f1, A = round(0.5 * nsamples(physeq)))

# Transform sample counts
r <- transform_sample_counts(physeq, function(x) x/sum(x))

# Sample names and variables
sample_names(physeq)

# Sample sums
sample_sums(physeq)
get_sample(physeq, taxa_names(physeq)[5])
```


### Variable operations

```{r preprocess4, warning=FALSE, message=FALSE}
sample_variables(physeq)
get_variable(physeq, sample_variables(physeq)[5])[1:10]

# Assign fields to sample metadata
# sample_data(GP)$human <- ..
```

### Taxa operations

```{r preprocess3, warning=FALSE, message=FALSE}
# Prune taxa
ex2 <- prune_taxa(wh1, physeq)
ex3 <- prune_taxa(taxa_sums(physeq) > 0, physeq)

# Subset taxa
pseq <- subset_taxa(pseq, Phylum == "Bacteroidetes")

# Filter taxa
f <- filter_taxa(r, function(x) var(x) > 1e-05, TRUE)

# Number of taxa
ntaxa(physeq)

# Rank names
rank_names(physeq)

# Taxa names
taxa_names(physeq)
get_taxa_unique(physeq, "Phylum")
get_taxa(physeq, sample_names(physeq)[5])

# Taxa sums
taxa_sums(physeq)
```


### Merging phyloseq operations

```{r preprocess-merge, warning=FALSE, message=FALSE}
# merge_phyloseq()
# merge_taxa()
# tax_glom()
```


### Rarification

```{r preprocess-rarify, warning=FALSE, message=FALSE}
physeq.rarified <- rarefy_even_depth(physeq)
```

