---
title: "Preprocessing"
bibliography: 
- bibliography.bib
- references.bib
output: 
  rmarkdown::html_vignette
---
<!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{microbiome tutorial - Preprocessing}
  %\usepackage[utf8]{inputenc}
  %\VignetteEncoding{UTF-8}  
-->


## Processing taxonomic profiling data

Instructions to manipulate microbiome data sets using tools from the [phyloseq package](http://joey711.github.io/phyloseq/) and some extensions from the [microbiome package](https://github.com/microbiome/microbiome), including subsetting, aggregating and filtering.


Load example data:

```{r data21}
library(phyloseq)
library(microbiome)

data(atlas1006)   # Load the data
pseq <- atlas1006 # Rename the data
```


### Retrieving data elements from a phyloseq object  

A phyloseq object contains OTU table (taxa abundances), sample
metadata, taxonomy table (mapping between OTUs and higher-level
taxonomic classifications), and phylogenetic tree (relations between
the taxa). Some of these are optional.


Pick metadata:

```{r data21b}
meta <- sample_data(pseq)
```

Taxonomy table:

```{r data22}
taxonomy <- tax_table(pseq)
```


Abundances for taxonomic groups ('OTU table') as a TaxaxSamples matrix:

```{r data23}
# Absolute abundances
otu.absolute <- abundances(pseq)

# Relative abundances
otu.relative <- abundances(pseq, "compositional")
```


Melt phyloseq data for easier plotting:

```{r data25}
df <- psmelt(pseq)
kable(head(df))
```



### Sample operations

Sample names and variables

```{r preprocess2b, warning=FALSE, message=FALSE}
head(sample_names(pseq))
```

Total OTU abundance in each sample

```{r preprocess2c, warning=FALSE, message=FALSE}
head(sample_sums(pseq))
```

Abundance of a given species in each sample

```{r preprocess2d, warning=FALSE, message=FALSE}
head(get_sample(pseq, "Akkermansia"))
```

Filter samples

```{r preprocess3, warning=FALSE, message=FALSE}
f1 <- filterfun_sample(topp(0.1))
taxa <- genefilter_sample(pseq, f1, A = round(0.5 * nsamples(pseq)))
```

Select samples by specific metadata fields

```{r data-subsetting}
pseq.subset <- subset_samples(pseq, nationality == "US")
```


Pick samples at the baseline time points only:

```{r baseline, message=FALSE, warning=FALSE}
pseq0 <- baseline(pseq)
```



### Data transformations

The microbiome package provides a wrapper for standard sample/OTU transforms. For arbitrary transforms, use the transform_sample_counts function in the phyloseq package.

Log10 transform (log(1+x) if the data contains zeroes). Also "Z",
"clr", and "hellinger" are available as common transforms.

```{r preprocessing-logtrans}
pseq.log <- transform(pseq, "log10")
```

Relative abundances (the input data needs to be in absolute scale, not logarithmic!):

```{r data10}
pseq1 <- microbiome::transform(pseq, "compositional", "OTU")
pseq2 <- phyloseq::transform_sample_counts(pseq, function(x) x/sum(x))
```


### Variable operations

Sample variable names

```{r preprocess4, warning=FALSE, message=FALSE}
sample_variables(pseq)
```

Pick values for a given variable

```{r preprocess4b, warning=FALSE, message=FALSE}
head(get_variable(pseq, sample_variables(pseq)[1]))
```

Assign new fields to metadata

```{r preprocess4b2, warning=FALSE, message=FALSE}
# Calculate diversity for samples
div <- microbiome::diversity(pseq, measures = "Shannon")$Shannon

# Assign this to sample metadata
sample_data(pseq)$diversity <- div
```

### Taxa operations


Number of taxa

```{r preprocess3e, warning=FALSE, message=FALSE}
n <- ntaxa(pseq)
```


Most abundant taxa

```{r toptaxa, warning=FALSE, message=FALSE}
topx <- top_taxa(pseq, n = 10)
```


Names

```{r preprocess3f, warning=FALSE, message=FALSE}
ranks <- rank_names(pseq)  # Taxonomic levels
taxa  <- taxa(pseq)        # Taxa names at the analysed level
```


Subset taxa:

```{r preprocess3c, warning=FALSE, message=FALSE}
pseq.bac <- subset_taxa(pseq, Phylum == "Bacteroidetes")
```


Prune (select) taxa:

```{r preprocess3b, warning=FALSE, message=FALSE}
# List of Genera in the Bacteroideted Phylum
taxa <- map_levels(NULL, "Phylum", "Genus", pseq)$Bacteroidetes

# With given taxon names
ex2 <- prune_taxa(taxa, pseq)

# Taxa with positive sum across samples
ex3 <- prune_taxa(taxa_sums(pseq) > 0, pseq)
```


Filter by user-specified function values (here variance):

```{r preprocess3d, warning=FALSE, message=FALSE}
f <- filter_taxa(pseq, function(x) var(x) > 1e-05, TRUE)
```


List unique phylum-level groups: 

```{r preprocess3g, warning=FALSE, message=FALSE}
head(get_taxa_unique(pseq, "Phylum"))
```

Pick the taxa abundances for a given sample:

```{r preprocess3gg, warning=FALSE, message=FALSE}
samplename <- sample_names(pseq)[[1]]

# Two ways to pick abundances for a particular taxon
tax.abundances <- get_taxa(pseq, samplename)
tax.abundances2 <- abundances(pseq)[, samplename]
```


### Merging operations

Aggregate taxa to higher taxonomic levels. This is particularly useful if the phylogenetic tree is missing. When it is available, see [merge_samples, merge_taxa and tax_glom](http://joey711.github.io/phyloseq/merge.html))

```{r data24}
pseq2 <- aggregate_taxa(pseq, "Phylum") 
```


Merging phyloseq objects

```{r preprocess-merge, warning=FALSE, message=FALSE, eval=FALSE}
merge_phyloseq(pseqA, pseqB)
```


### Rarification

```{r preprocess-rarify, warning=FALSE, message=FALSE, eval=FALSE}
pseq.rarified <- rarefy_even_depth(pseq)
```


### Taxonomy 

Convert between taxonomic levels (here from Genus (Akkermansia) to
Phylum (Verrucomicrobia):

```{r phylogeny-example2, warning=FALSE, message=FALSE}
m <- map_levels("Akkermansia", "Genus", "Phylum", tax_table(pseq))
print(m)
```


### Metadata

Visualize frequencies of given factor (sex) levels within the
indicated groups (group):

```{r phylogeny-example3, warning=FALSE, message=FALSE, fig.width=10, fig.height=6}
res <- plot_frequencies(sample_data(pseq), "bmi_group", "gender")
print(res$plot)

# Retrieving the actual data values:
kable(head(res$data), digits = 2)
```

