## Filtering and pruning

The external high-quality [phyloseq package](http://joey711.github.io/phyloseq/) provides a complete set of tools for data preprocessing and filtering purposes. Download example data from [O'Keefe et al. Nat. Comm. 6:6342, 2015](http://dx.doi.org/10.1038/ncomms7342) from [Data Dryad](http://dx.doi.org/10.5061/dryad.1mn1n). This is a two-week diet swap study between western (USA) and traditional (rural Africa) diets including microbiota profiling:

```{r phyloseqdata, fig.width=10, fig.height=7, warning=FALSE, message=FALSE}
library(microbiome)
pseq <- download_microbiome("dietswap")
```

### Sample operations


Transform sample counts. Here, calculate relative abundances:

```{r preprocess2, warning=FALSE, message=FALSE}
r <- transform_sample_counts(pseq, function(x) x/sum(x))
```

Sample names and variables

```{r preprocess2b, warning=FALSE, message=FALSE}
head(sample_names(pseq))
```

Sample sums

```{r preprocess2c, warning=FALSE, message=FALSE}
head(sample_sums(pseq))
```

Abundance for species ‘i’ in each sample

```{r preprocess2d, warning=FALSE, message=FALSE}
head(get_sample(pseq, taxa_names(pseq)[1]))
```


### Variable operations

Sample variable names

```{r preprocess4, warning=FALSE, message=FALSE}
sample_variables(pseq)
```

Pick variable values for a given variable

```{r preprocess4b, warning=FALSE, message=FALSE}
head(get_variable(pseq, sample_variables(pseq)[3]))
# Assign fields to sample metadata
# sample_data(GP)$human <- ..
```

### Taxa operations


Filter samples

```{r preprocess3, warning=FALSE, message=FALSE}
f1 <- filterfun_sample(topp(0.1))
taxa <- genefilter_sample(pseq, f1, A = round(0.5 * nsamples(pseq)))
```

Prune taxa

```{r preprocess3b, warning=FALSE, message=FALSE}
# With given taxon names
ex2 <- prune_taxa(taxa, pseq)

# Taxa with positive sum across samples
ex3 <- prune_taxa(taxa_sums(pseq) > 0, pseq)
```

Subset taxa

```{r preprocess3c, warning=FALSE, message=FALSE}
pseq <- subset_taxa(pseq, Phylum == "Bacteroidetes")
```


Filter by user-specified function values (here variance)

```{r preprocess3d, warning=FALSE, message=FALSE}
f <- filter_taxa(r, function(x) var(x) > 1e-05, TRUE)
```

Number of taxa

```{r preprocess3e, warning=FALSE, message=FALSE}
ntaxa(pseq)
```


Names

```{r preprocess3f, warning=FALSE, message=FALSE}
rank_names(pseq)
taxa_names(pseq)
```


Pick taxa by rank

```{r preprocess3g, warning=FALSE, message=FALSE}
# Unique phyla
head(get_taxa_unique(pseq, "Phylum"))

Pick taxa by sample

```{r preprocess3gg, warning=FALSE, message=FALSE}
get_taxa(pseq, sample_names(pseq)[1])
```


Taxa sums

```{r preprocess3h, warning=FALSE, message=FALSE}
head(taxa_sums(pseq))
```


### Merging operations for phyloseq objects

```{r preprocess-merge, warning=FALSE, message=FALSE}
# merge_phyloseq()
# merge_taxa()
# tax_glom()
```


### Rarification

```{r preprocess-rarify, warning=FALSE, message=FALSE}
pseq.rarified <- rarefy_even_depth(pseq)
```

