### RDA analysis and visualization. 

Load the package and example data:

```{r rda, warning=FALSE, message=FALSE, fig.path = "figure/"}
library(microbiome)
# Data from https://peerj.com/articles/32/
pseq <- download_microbiome("peerj32")$physeq
```

### Standard RDA 

Standard RDA for microbiota profiles versus the 'time' variable from 
sample metadata:

```{r rda2, warning=FALSE, message=FALSE}
rdatest <- rda_physeq(pseq, "time")
```

### RDA significance test

```{r rda2b, warning=FALSE, message=FALSE}
permutest(rdatest) 
```

### RDA visualization

Visualizing the standard RDA output:

```{r rda4, warning=FALSE, message=FALSE, fig.width=8, fig.height=8, fig.path = "figure/"}
plot(rdatest, choices = c(1,2), type = "points", pch = 15, scaling = 3, cex = 0.7, col = meta$time)
points(rdatest, choices = c(1,2), pch = 15, scaling = 3, cex = 0.7, col = meta$time)
pl <- ordihull(rdatest, meta$time, scaling = 3, label = TRUE)
title("RDA")
```

### Bagged RDA

Fitting bagged RDA on a phyloseq object:

```{r rda5, warning=FALSE, message=FALSE, fig.show='none'}
res <- bagged_rda(pseq, "group", sig.thresh=0.05, nboot=100)
```

Visualizing bagged RDA:

```{r rda6, warning=FALSE, message=FALSE, fig.width=8, fig.height=8}
plot_bagged_rda(res)
```


### Controlling confounding variables with RDA

For more complex scenarios, use the vegan package directly:

```{r rda3, warning=FALSE, message=FALSE, fig.width=8, fig.height=8, fig.path = "figure/"}
# Pick microbiota profiling data from the phyloseq object
otu <- otu_table(pseq)@.Data

# Sample annotations
meta <- sample_data(pseq)

# RDA with confounders
rdatest2 <- rda(t(otu) ~ meta$time + Condition(meta$subject + meta$gender))
```



