---
title: "RDA"
bibliography: 
- bibliography.bib
- references.bib
output: 
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---
<!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{microbiome tutorial - rda}
  %\usepackage[utf8]{inputenc}
  %\VignetteEncoding{UTF-8}  
-->


## RDA analysis and visualization. 

Load the package and example data:

```{r rda, warning=FALSE, message=FALSE, eval=TRUE}
library(microbiome)
data(peerj32) # Data from https://peerj.com/articles/32/
pseq <- peerj32$phyloseq # phyloseq data
pseq.trans <- transform(pseq, "hell") # Hellinger transform
```


### Bagged RDA

Bagged RDA provides added robustness in the analysis compared to the standard RDA. Fit bagged (bootstrap aggregated) RDA on a phyloseq object (alternatively you could apply it to the abundance matrix and covariates directly):

```{r rda5, warning=FALSE, message=FALSE, eval=TRUE}
# In any real study, use bs.iter = 100 or higher
# to achieve meaningful benefits from the bagged version.
# In this example we use bs.iter = 2 just to speed up the
# example code for educational purposes
res <- rda_bagged(pseq.trans, "group", bs.iter=2)
```

Visualizing bagged RDA:

```{r rda6, warning=FALSE, message=FALSE, fig.width=8, fig.height=8, eval=TRUE}
plot_rda_bagged(res)
```



### Standard RDA 

Standard RDA for microbiota profiles versus the given (here 'time')
variable from sample metadata (see also the RDA method in
phyloseq::ordinate)

```{r rda2, warning=FALSE, message=FALSE}
x <- pseq.trans
otu <- abundances(x)
metadata <- meta(x)

library(vegan)
rda.result <- vegan::rda(t(otu) ~ factor(metadata$time),
                         na.action = na.fail, scale = TRUE)
```

Proportion explained by the given factor

```{r rda2b, warning=FALSE, message=FALSE}
summary(rda.result)$constr.chi/summary(rda.result)$tot.chi
```


### RDA visualization

Visualize the standard RDA output.

```{r rda4, warning=FALSE, message=FALSE, fig.width=8, fig.height=8}
plot(rda.result, choices = c(1,2), type = "points", pch = 15, scaling = 3, cex = 0.7, col = metadata$time)
points(rda.result, choices = c(1,2), pch = 15, scaling = 3, cex = 0.7, col = metadata$time)
pl <- ordihull(rda.result, metadata$time, scaling = 3, label = TRUE)
```


### RDA significance test

```{r rda2bc, warning=FALSE, message=FALSE, eval=TRUE}
permutest(rda.result) 
```


### RDA with confounding variables 

For more complex RDA scenarios, use the standard RDA available via the
vegan R package.

```{r rda3, warning=FALSE, message=FALSE, fig.width=8, fig.height=8, eval=TRUE}
# Pick microbiota profiling data from the phyloseq object
otu <- abundances(pseq.trans)

# Sample annotations
metadata <- meta(pseq.trans)

# RDA with confounders using the vegan function
rda.result2 <- vegan::rda(t(otu) ~ metadata$time + Condition(metadata$subject + metadata$gender))
```




```{r cleanup, echo=FALSE, warning=FALSE, message=FALSE}
    # Must do to clean up some space
    pkgs0 <- c("ade4",
      	       "compositions",
    	       "MASS",
    	       "moments",
    	       "scales",
    	       "tgp",
    	       "WGCNA",
    	       "diptest",
    	       "FD",
    	       "gcookbook",
    	       "GGally",
    	       "ggnet",
    	       "Hmisc",
    	       "hrbrthemes",
    	       "igraph",
    	       "intergraph",
    	       "limma",
    	       "lme4",
    	       "netresponse",
    	       "network",
    	       "RColorBrewer",
    	       "sna",
    	       "SpiecEasi",
    	       "tidyverse",
    	       "viridis")
    for (pkg in pkgs0) {
        print(pkg)
        try(a <- unloadNamespace(pkg))
    }
```


