### RDA analysis and visualization. 

Load the package and example data:

```{r rda, warning=FALSE, message=FALSE, fig.path = "figure/"}
library(microbiome)
# Data from https://peerj.com/articles/32/
#pseq <- download_microbiome("peerj32")$phyloseq
data("peerj32")
pseq <- peerj32$phyloseq
```

### Standard RDA 

Standard RDA for microbiota profiles versus the given (here 'time')
variable from sample metadata:

```{r rda2, warning=FALSE, message=FALSE}
# If x has zeroes we can use log(1 + x) transformation
pseq.log10 <- transform_sample_counts(pseq, function (x) {log10(x)})
rda.result <- rda_physeq(pseq.log10, "time")
```

### RDA visualization

Visualizing the standard RDA output:

```{r rda4, warning=FALSE, message=FALSE, fig.width=8, fig.height=8, fig.path = "figure/"}
library(phyloseq)
meta <- sample_data(pseq.log10)
plot(rda.result, choices = c(1,2), type = "points", pch = 15, scaling = 3, cex = 0.7, col = meta$time)
points(rda.result, choices = c(1,2), pch = 15, scaling = 3, cex = 0.7, col = meta$time)
library(vegan)
pl <- ordihull(rda.result, meta$time, scaling = 3, label = TRUE)
title("RDA")
```

See also the RDA method in phyloseq ordinate function, which is calculated without the formula.


### RDA significance test

```{r rda2b, warning=FALSE, message=FALSE}
library(vegan)
permutest(rda.result) 
```

### Bagged RDA

Fitting bagged RDA on a phyloseq object:

```{r rda5, warning=FALSE, message=FALSE, fig.show='none'}
res <- bagged_rda(pseq.log10, "group", sig.thresh=0.05, nboot=100)
```

Visualizing bagged RDA:

```{r rda6, warning=FALSE, message=FALSE, fig.width=8, fig.height=8}
plot_bagged_rda(res)
```


### Controlling confounding variables with RDA

For more complex scenarios, use the vegan package directly:

```{r rda3, warning=FALSE, message=FALSE, fig.width=8, fig.height=8, fig.path = "figure/"}
# Pick microbiota profiling data from the phyloseq object
otu <- otu_table(pseq.log10)@.Data

# Sample annotations
meta <- sample_data(pseq.log10)

# RDA with confounders
rda.result2 <- rda(t(otu) ~ meta$time + Condition(meta$subject + meta$gender))
```



