### Stability analysis within group of samples

Calculate stability as the average correlation between samples and their mean for a given phylotypes vs. samples matrix:

```{r stability-example, message=FALSE}
library(microbiome)

data.directory <- system.file("extdata", package = "microbiome")

genus.matrix.log10.simulated <- read.profiling(level = "oligo",		
			     	  	      method = "frpa", 
			                    data.dir = data.directory, 
			     	      	       log10 = TRUE)  

# Stability: 
stability <- estimate.stability(genus.matrix.log10.simulated)$stability
```

### Stability analysis in time

Calculate correlations between time points 1 and 2 for each subject,
then calculate stability as the average correlation of these
subject-specific correlations. Done as above, but instead of one
matrix, give two matrices as input. These are phylotypes vs. samples
matrices and correspond to time points 1 and 2, respectively. The rows
(phylotypes) and columns (subjects) should be in the same order in
both matrices.



### Version information

```{r versioninfo}
sessionInfo()
```

