
### Stability analysis 

Calculate stability as the average correlation between samples and their
mean for a given phylotypes vs. samples matrix:

```{r bistability}
# Download example data (Lahti et al. Nat. Comm. 5:4344, 2014)
library(microbiome)
data.atlas1006 <- download_microbiome("atlas1006")

# Quantify temporal stability across the abundance range
stability <- intermediate_stability(data.atlas1006, method = "correlation")
intermediate.stability <- sapply(stability, function (x) {x$stability})
hist(intermediate.stability, main = "Intermediate stability")
```


## Bimodality quantification

Calculate coefficient of bimodality (used in [Shade et
al.](http://mbio.asm.org/content/5/4/e01371-14)) for taxa in an
example data set, and plot the taxa with the lowest and highest score:

```{r bimodality, message=FALSE, warning=FALSE, fig.path = "figure/"}
# Pick samples from zero time point (cross-sectional analysis)
pseq0 <- subset_samples(data.atlas1006, time == 0 & DNA_extraction_method == "r")

# Calculate coefficient of bimodality for each taxa
# (calculated for log10 abundances)
bimodality <- multimodality(pseq0, method = "potential.bootstrap")
bimodality.coef <- multimodality(pseq0, method = "coefficient.of.bimodality")

# Compare the two bimodality scores
# plot(bimodality, bimodality.coef)

# Pick log10 data
pseq0log <- transform_sample_counts(pseq0, function(x) log10(x))
otu.log10 <- otu_table(pseq0log)@.Data

# Pick the most and least bimodal taxa as examples
unimodal <- names(which.min(bimodality))
bimodal <- names(which.max(bimodality))

# Visualize population frequencies
par(mfrow = c(1,2))
plot(density(otu.log10[unimodal, ]), main = unimodal)
plot(density(otu.log10[bimodal, ]), main = bimodal)
```

## Bimodality versus intermediate stability

The analysis suggests that bimodal taxa tend to have instable intermediate abundances (following [Lahti et al. Nat. Comm. 5:4344, 2014](http://www.nature.com/ncomms/2014/140708/ncomms5344/full/ncomms5344.html) but with a different subset, parameters and model):


```{r bimodalitybistability, message=FALSE, warning=FALSE, fig.path = "figure/"}
# For clarity, visualize only the prevalent taxa seen with HITChip signal >250
# (note HITChip signal reflects read count but is conceptually different) 
# in at least 10% of the samples. 
pseqf <- filter_taxa(pseq0, function(x) sum(x > 300) > (0.2*length(x)), TRUE)

# Taxa that have bistability and bimodality estimates
s <- taxa_names(pseqf)

plot(intermediate.stability[s], bimodality[s], xlab = "Intermediate stability", ylab = "Bimodality", type = "n")
text(intermediate.stability[s], bimodality[s], label = s, cex = 0.7)
```


### Variability within group of samples (inter-individual stability)

Assess 'inter-individual stability' as in [Salonen et al. ISME J 2014](http://www.nature.com/ismej/journal/v8/n11/full/ismej201463a.html). This is defined as the average correlation between samples and their mean for a given samples vs phylotypes matrix. For the illustration, calculate inter-individual stability (variability) separately for Placebo and LGG groups.

```{r variability-example2, message=FALSE}

# Example data
library(microbiome)
x <- download_microbiome("dietswap")

# Add time field (two time points needed within each group for the 
# intraindividual method)
sample_data(x)$time <- sample_data(x)$timepoint.within.group

# Estimate inter-individual variability
res <- estimate_variability(x, "interindividual")

# Estimate intra-individual variability
res <- estimate_variability(x, "intraindividual")

# Visualize
library(ggplot2)
theme_set(theme_bw(20))
p <- ggplot(res$data, aes(x = group, y = correlation))
p <- p + geom_boxplot()
p <- p + ggtitle(paste("Inter-individual variability (p=", round(res$p.value, 6), ")"))
p <- p + ylab("Correlation")
print(p)
```


### Variability within subjects over time (intra-individual stability)

Assess 'intra-individual stability' as in [Salonen et al. ISME J 2014](http://www.nature.com/ismej/journal/v8/n11/full/ismej201463a.html). This is defined as the average correlation between two time points within subjects, calculated separately within each group. For illustration, check intra-individual stability (variability) separately for Placebo and LGG groups.

```{r variability-intra, message=FALSE}
res <- estimate_variability(x, m, "intraindividual")
library(ggplot2)
theme_set(theme_bw(20))
p <- ggplot(res$data, aes(x = group, y = correlation))
p <- p + geom_boxplot()
p <- p + ggtitle(paste("Intra-individual variability (p=", round(res$p.value, 6), ")"))
p <- p + ylab("Correlation")
print(p)
```



### Version information

```{r stabilitysessioninfo}
sessionInfo()
```

