# Microbiome stability analysis

Download example data - HITChip Atlas of 130 genus-like taxa across 1006 healthy western adults from [Lahti et al. Nat. Comm. 5:4344, 2014](http://www.nature.com/ncomms/2014/140708/ncomms5344/full/ncomms5344.html). A subset of 76 subjects have also short time series available for temporal stability analysis:

```{r bistability}
library(microbiome)
pseq <- download_microbiome("atlas1006")

# Let us keep only prevalent taxa
# (HITChip signal >3 in >20 percent of the samples)
pseq <- filter_prevalent(pseq, detection.threshold = 10^3,
     			       prevalence.threshold = 0.2)
```



## Quantify intermediate stability 

It has been reported that certain microbial groups exhibit bi-stable
abundance distributions with distinct peaks at low and high
abundances, and an instable intermediate abundance range.

Instability at the intermediate abundance range is hence one indicator
of bi-stability.

[Lahti et
al. 2014](http://www.nature.com/ncomms/2014/140708/ncomms5344/full/ncomms5344.html))
use straightforward correlation analysis to quantify how the distance
from the intermediate abundance region (50% quantile) is associated
with the observed shifts between consecutive time points. This can be
calculated with:

```{r bistability2}
intermediate.stability <- intermediate_stability(pseq, output = "scores")
```


## Quantify bimodality 

Bistable systems often exhibit bimodal population patterns. Hence,
bimodality of the abundance distribution provides another indicator of
bistability, although other explanations for bimodality (sampling
biases; heterogeneous population structure etc.) are also possible.

Multiple bimodality scores are available. Perform cross-sectional
analysis of bimodality by including only the samples from the zero
time point:

```{r bimodality, message=FALSE, warning=FALSE, fig.path = "figure/"}
pseq0 <- subset_samples(pseq, time == 0 & DNA_extraction_method == "r")
```


[Multimodality score with potential analysis with
bootstrap](http://www.nature.com/ncomms/2014/140708/ncomms5344/full/ncomms5344.html)


```{r bimodality2, message=FALSE, warning=FALSE, fig.path = "figure/"}
bimodality.pb <- bimodality(pseq0, method = "potential_bootstrap")
```

Sarle's bimodality coefficient (see help(coefficient_of_bimodality)):

```{r bimodality3, message=FALSE, warning=FALSE, fig.path = "figure/"}
bimodality.sarle <- bimodality(pseq0, method = "Sarle.finite.sample")
```


DIP test for multimodality (not included in the microbiome package):

```{r bimodality-dip, message=FALSE, warning=FALSE, fig.path = "figure/"}
bimodality.dip <- bimodality(pseq0, method = "dip")
```


Visualize population densities 

```{r stability2, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
# Pick the most and least bimodal taxa as examples
bimodality <- bimodality.pb
unimodal <- names(which.min(bimodality))
bimodal  <- names(which.max(bimodality))

# Visualize population frequencies
p1 <- plot_density(pseq, otu.name = unimodal, log10 = TRUE) 
p2 <- plot_density(pseq, otu.name = bimodal,  log10 = TRUE) 
library(gridExtra)
grid.arrange(p1, p2, nrow = 1)
```


## Comparing bimodality and intermediate stability

The analysis suggests that bimodal population distribution across individuals is often associated with instable intermediate abundances within individuals. The specific bi-stable groups in the upper left corner were suggested in [Lahti et al. Nat. Comm. 5:4344, 2014](http://www.nature.com/ncomms/2014/140708/ncomms5344/full/ncomms5344.html):

```{r bimodalitybistability, message=FALSE, warning=FALSE, fig.path = "figure/"}
df <- data.frame(group = tax,
                 intermediate.stability = intermediate.stability[tax],
		 bimodality = bimodality.pb[tax])
theme_set(theme_bw(20))
p <- ggplot(df, aes(x = intermediate.stability, y = bimodality, label = group))
p <- p + geom_text(size = 3)
p
```


### TODO

 * tipping element heatmap

 * stability line plot

 * classification of the taxa into the distinct abundance types

### Version information

```{r stabilitysessioninfo}
sessionInfo()
```

