
### Stability analysis 

Calculate stability as the average correlation between samples and their
mean for a given phylotypes vs. samples matrix:

```{r bistability}
# Example data
library(microbiome)
data.atlas1006 <- download_microbiome("atlas1006")
x <- data.atlas1006$microbes
m <- data.atlas1006$meta

# Convert data into phyloseq format
pseq <- hitchip2physeq(x, m, taxonomy = NULL)

# Quantify temporal stability across the abundance range
stability <- estimate_stability(pseq, method = "correlation")
intermediate.stability <- sapply(stability, function (x) {x$stability})
hist(intermediate.stability, main = "Intermediate stability")
```


## Bimodality quantification

Calculate coefficient of bimodality (used in [Shade et
al.](http://mbio.asm.org/content/5/4/e01371-14)) for taxa in an
example data set, and plot the taxa with the lowest and highest score:

```{r bimodality, message=FALSE, warning=FALSE, fig.path = "figure/"}
# Pick samples from zero time point (cross-sectional analysis)
s <- subset(m, time == 0)$sample
m0 <- m[s,]
x0 <- x[s,]

# Calculate coefficient of bimodality for each taxa
bimodality <- apply(x0, 2, coefficient.of.bimodality)

# Pick the most and least bimodal taxa as examples
unimodal <- names(which.min(bimodality))
bimodal <- names(which.max(bimodality))

# Visualize
par(mfrow = c(1,2))
plot(density(x[, unimodal]), main = unimodal)
plot(density(x[, bimodal]), main = bimodal)
```

## Bimodality versus intermediate stability

The analysis suggests that bimodal taxa tend to have instable intermediate abundances:

```{r bimodalitybistability, message=FALSE, warning=FALSE, fig.path = "figure/"}
s <- intersect(names(intermediate.stability), names(bimodality))
plot(intermediate.stability[s], bimodality[s], xlab = "Intermediate stability", ylab = "Bimodality")
```


### Variability within group of samples (inter-individual stability)

Assess 'inter-individual stability' as in [Salonen et al. ISME J 2014](http://www.nature.com/ismej/journal/v8/n11/full/ismej201463a.html). This is defined as the average correlation between samples and their mean for a given samples vs phylotypes matrix. For the illustration, calculate inter-individual stability (variability) separately for Placebo and LGG groups.

```{r variability-example2, message=FALSE}
# Example data
library(microbiome)
data.peerj32 <- download_microbiome("peerj32")
x <- data.peerj32$microbes
m <- data.peerj32$meta

# Estimate inter-individual variability
res <- estimate_variability(x, m, "interindividual")
library(ggplot2)
theme_set(theme_bw(20))
p <- ggplot(res$data, aes(x = group, y = correlation))
p <- p + geom_boxplot()
p <- p + ggtitle(paste("Inter-individual variability (p=", round(res$p.value, 6), ")"))
p <- p + ylab("Correlation")
print(p)
```


### Variability within subjects over time (intra-individual stability)

Assess 'intra-individual stability' as in [Salonen et al. ISME J 2014](http://www.nature.com/ismej/journal/v8/n11/full/ismej201463a.html). This is defined as the average correlation between two time points within subjects, calculated separately within each group. For illustration, check intra-individual stability (variability) separately for Placebo and LGG groups.

```{r variability-intra, message=FALSE}
res <- estimate_variability(x, m, "intraindividual")
library(ggplot2)
theme_set(theme_bw(20))
p <- ggplot(res$data, aes(x = group, y = correlation))
p <- p + geom_boxplot()
p <- p + ggtitle(paste("Intra-individual variability (p=", round(res$p.value, 6), ")"))
p <- p + ylab("Correlation")
print(p)
```



### Version information

```{r stabilitysessioninfo}
sessionInfo()
```

