---
title: "Project Template"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  md_document:
    toc: false
  html_document:
    toc: true
    number_sections: true
    theme: united
    highlight: pygments
---
<!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Project Template}
  %\usepackage[utf8]{inputenc}
-->


Minimal example
===========

To test this example, do the following:

 1. Copy the [Template.Rmd](https://raw.githubusercontent.com/microbiome/microbiome/master/vignettes/Template.Rmd) (this file) on your computer. NOTE: change the file ending from .txt to .Rmd !!
 1. Start [RStudio](http://www.rstudio.com/)
 1. Open the [Template.Rmd](https://raw.githubusercontent.com/microbiome/microbiome/master/vignettes/Template.Rmd) file in RStudio
 1. Convert the Rmd file with the 'knit HTML' button
 1. Start expanding this file to make your own report. See examples from the [example workflow](https://github.com/microbiome/microbiome/blob/master/vignettes/Atlas.Rmd) and [microbiome tutorial](https://github.com/microbiome/microbiome/blob/master/vignettes/vignette.md)


### Update the microbiome package

```{r install2}
library(devtools)
install_github("microbiome/microbiome")
```

### Load the microbiome package

```{r loading, eval=TRUE}
library(microbiome)  
```

