### Inter-individual homogeneity (within group of samples)

Assess 'inter-individual stability', or homogeneity, as in [Salonen et al. ISME J 2014](http://www.nature.com/ismej/journal/v8/n11/full/ismej201463a.html). This is defined as the average correlation between the samples and their mean for a given samples vs phylotypes matrix. For illustration, calculate inter-individual homogeneity separately for Placebo and LGG groups. Note that this homogeneity measure is affected by sample size.

Load example data

```{r heterogeneity-example2, message=FALSE}
library(microbiome)
data("dietswap")
x <- dietswap

# Add time field (two time points needed within each group for the 
# intraindividual method)
sample_data(x)$time <- sample_data(x)$timepoint.within.group
```


Heterogeneity across subjects within a group

```{r heterogeneity-example2b, message=FALSE}
res <- estimate_homogeneity(x, "interindividual")
```


Visualize

```{r homogeneity-example2d, message=FALSE}
library(ggplot2)
theme_set(theme_bw(20))
p <- ggplot(res$data, aes(x = group, y = correlation))
p <- p + geom_boxplot()
p <- p + ggtitle(paste("Inter-individual homogeneity (p=", round(res$p.value, 6), ")", sep = ""))
p <- p + ylab("Correlation")
print(p)
```


### Intra-individual stability

Homogeneity within subjects over time (also called intra-individual stability in [Salonen et al. ISME J 2014](http://www.nature.com/ismej/journal/v8/n11/full/ismej201463a.html)). Defined as the average correlation between two time points within subjects within each group. For illustration, check intra-individual stability (homogeneity) separately for Placebo and LGG groups.

```{r homogeneity-example2c, message=FALSE}
res <- estimate_homogeneity(x, "intraindividual")
```


Visualize

```{r homogeneity-intra, message=FALSE, warning=FALSE, fig.height=5, fig.height=5}
library(ggplot2)
theme_set(theme_bw(20))
p <- ggplot(res$data, aes(x = group, y = correlation))
p <- p + geom_boxplot()
p <- p + ggtitle(paste("Intra-individual homogeneity (p=", round(res$p.value, 6), ")"))
p <- p + ylab("Correlation")
print(p)
```


### Time series

```{r homogeneity-timeseries, message=FALSE, warning=FALSE, fig.height=5, fig.height=5}
data("atlas1006")
pseq <- atlas1006
pseq <- subset_samples(pseq, DNA_extraction_method == "r")
pseq <- transform_phyloseq(pseq, "relative.abundance")
p <- plot_timeseries(pseq, "Dialister", subject = "831", tipping.point = 0.5)
print(p)
```


## Further visualization tools

Draw regression curve with smoothed error bars based on
the [Visually-Weighted Regression](http://www.fight-entropy.com/2012/07/visually-weighted-regression.html) by Solomon M. Hsiang. The sorvi implementation extends [Felix Schonbrodt's original code](http://www.nicebread.de/visually-weighted-watercolor-plots-new-variants-please-vote/).

```{r variability-regressionline, message=FALSE, eval=TRUE, fig.width=10, fig.height=5, warning=FALSE, dev="CairoPNG"}
data(atlas1006)
p <- plot_regression(diversity ~ age, sample_data(atlas1006))
print(p)
```

### Version information

```{r stabilitysessioninfo}
sessionInfo()
```

