### Inter-individual variability (variability within group of samples)

Assess 'inter-individual stability' as in [Salonen et al. ISME J 2014](http://www.nature.com/ismej/journal/v8/n11/full/ismej201463a.html). This is defined as the average correlation between samples and their mean for a given samples vs phylotypes matrix. For the illustration, calculate inter-individual stability (variability) separately for Placebo and LGG groups.

Load example data

```{r variability-example2, message=FALSE}
library(microbiome)
x <- download_microbiome("dietswap")
# Add time field (two time points needed within each group for the 
# intraindividual method)
sample_data(x)$time <- sample_data(x)$timepoint.within.group
```


Variability across subjects within a group

```{r variability-example2b, message=FALSE}
res <- estimate_variability(x, "interindividual")
```


Visualize

```{r variability-example2d, message=FALSE}
library(ggplot2)
theme_set(theme_bw(20))
p <- ggplot(res$data, aes(x = group, y = correlation))
p <- p + geom_boxplot()
p <- p + ggtitle(paste("Inter-individual variability (p=", round(res$p.value, 6), ")"))
p <- p + ylab("Correlation")
print(p)
```


### Intra-individual variability

Variability within subjects over time (intra-individual stability). Assess 'intra-individual stability' as in [Salonen et al. ISME J 2014](http://www.nature.com/ismej/journal/v8/n11/full/ismej201463a.html). This is defined as the average correlation between two time points within subjects, calculated separately within each group. For illustration, check intra-individual stability (variability) separately for Placebo and LGG groups.

```{r variability-example2c, message=FALSE}
res <- estimate_variability(x, "intraindividual")
```


Visualize

```{r variability-intra, message=FALSE}
library(ggplot2)
theme_set(theme_bw(20))
p <- ggplot(res$data, aes(x = group, y = correlation))
p <- p + geom_boxplot()
p <- p + ggtitle(paste("Intra-individual variability (p=", round(res$p.value, 6), ")"))
p <- p + ylab("Correlation")
print(p)
```


### Version information

```{r stabilitysessioninfo}
sessionInfo()
```

