---
title: "microbiome vignette"
author: "Leo Lahti and Jarkko Salojarvi"
date: "`r Sys.Date()`"
bibliography: 
- bibliography.bib
- references.bib
output: 
  md_document:
    variant: markdown_github
---
<!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{microbiome tutorial}
  %\usepackage[utf8]{inputenc}
-->

```{r, echo=FALSE, message=FALSE}
# Handle citations
require(knitcitations)
cleanbib()
#options("citation_format" = "pandoc")
bib <- read.bibtex("bibliography.bib")
#opts_chunk$set(fig.width=4, fig.height=3, par=TRUE, out.width='2in', fig.pos='H')
#knitr::opts_chunk$set(fig.path = "figure/")
```


microbiome R package
===========

The microbiome package contains general-purpose tools for microarray-based analysis of microbiome profiling data sets in R `r citep(bib[["rcore13"]])`. 

The analyses are based on the phyloseq class from the independent [phyloseq](https://github.com/joey711/phyloseq) R package. Convert your data into phyloseq format before testing these examples.


### How to start 

Example on reproducible document generation:  

 * [Installing R tools](Installation.md)
 * [How to get started](Template.md)  
 * [Example workflow](Atlas.md)


### Data

* [Download example data sets](Data.md)
* [Extract data from HITChip database](https://github.com/microbiome/HITChipDB/blob/master/vignettes/vignette.md)
* [Get phylogeny](Phylogeny.md)


### Visualization and related tools

Download some [example data sets](Data.md) to try these tools:

* [Barplots](Barplots.md)
* [Boxplots](Boxplots.md)
* [Density](Density.md)
* [Heatmaps](Heatmap.md)
* [Motion charts](Motionchart.md)
* [PCA and other ordination/projection methods](PCA.md)
* [Cross hybridization](Crosshyb.md)


### Clustering 
* [Bimodality](Stability.md)
* [Clustering](Clustering.md)


### Microbiota composition
* [Bistability analysis](Stability.md)
* [Core microbiota](Core.md)
* [Diversity](Diversity.md)
* [Probe level studies](Probelevel.md)
* [RelativeAbundance](RelativeAbundance.md)
* [Tipping elements](Stability.md)
* [Variability](Variability.md)


### Linear models, comparisons, and association studies
* [Linear models](limma.md)
* [Pairwise comparisons](Comparisons.md)
* [Cross correlations](Crosscorrelation.md)


### Other statistical analysis
* [ROC curves](ROC.md)
* [RDA](RDA.md)
* [Phyloseq tools](Phyloseq.md)


### Output
* [Generating output files/figures](Output.md)


### Misc
* [Miscellaneous](misc.md)


### Licensing and Citations

This work can be freely used, modified and distributed under the 
[Two-clause FreeBSD license](http://en.wikipedia.org/wiki/BSD\_licenses).

Kindly cite the work as 'Leo Lahti and Jarkko Salojarvi
(2014). microbiome R package. URL: http://microbiome.github.com'.


### Dependencies

The package utilizes tools from a number of other CRAN and
Bioconductor extensions, including:

 * df2json `r citep(citation("df2json"))`
 * rjson `r citep(citation("rjson"))`
 * ade4 `r citep(citation("ade4"))`
 * mixOmics `r citep(citation("mixOmics"))`
 * RCurl `r citep(citation("RCurl"))`
 * vegan `r citep(citation("vegan"))`
 * reshape `r citep(citation("reshape"))`
 * WGCNA `r citep(citation("WGCNA"))`
 * ggplot2 `r citep(citation("ggplot2"))`
 * RPA `r citep(citation("RPA"))`
 * minet `r citep(citation("minet"))`
 * fastcluster `r citep(citation("fastcluster"))`
 * plyr `r citep(citation("plyr"))`


### References

```{r, echo=FALSE, message=FALSE}
#You can embed citations, for example: `r citep(bib[["lahti14natcomm"]])`
#You can embed citations, for example2: @lahti14natcomm
#Cite with DOI: `r citep("10.1890/11-0011.1")`
#Cite URL `r citep("http://knowledgeblog.org/greycite")`
#For automated markdown citations, check [this](http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html), [this](https://github.com/cboettig/knitcitations), and [this](http://www.carlboettiger.info/2012/03/24/citations-in-markdown-using-knitr.html). 
#write.bibtex(file="references.bib")
```

```{r, echo=FALSE, message=FALSE, results='asis'}
bibliography()
```

### Session info

This vignette was created with

```{r sessioninfo}
sessionInfo()
```




