---
title: "microbiome vignette"
author: "Leo Lahti and Jarkko Salojarvi"
date: "`r Sys.Date()`"
bibliography: 
- bibliography.bib
- references.bib
output:
  md_document:
    toc: true
---
<!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{microbiome tutorial}
  %\usepackage[utf8]{inputenc}
-->


microbiome R package
===========

The microbiome package contains general-purpose tools for
microarray-based analysis of microbiome profiling data sets. 

### Example workflows
* [Minimal example](Template.Rmd)
* [Atlas](Atlas.Rmd)

### Installation, example data sets and preprocessing
* [Installation](Installation.Rmd)
* [Data](Data.Rmd)
* [RPA](RPA.Rmd)
* [Preprocessing](Preprocessing.Rmd)
* [Phylogeny](Phylogeny.Rmd)

### Visualization and related tools

* [Barplots](Barplots.Rmd)
* [Boxplots](Boxplots.Rmd)
* [Heatmaps](Heatmap.Rmd)
* [Matrix visualization](Matrix-visualization.Rmd)
* [Motion charts](Motionchart.Rmd)
* [Ordination](Projections.Rmd)
* [Oligo heatmap](Oligoheatmap.Rmd)
* [Cross hybridization](Crosshyb.Rmd)

### Clustering 
* [Bimodality](Bimodality.Rmd)
* [Clustering](Clustering.Rmd)
* [Distance metrics](Metrics.Rmd)

### Microbiota composition
* [Core microbiota](Core.Rmd)
* [Diversity](Diversity.Rmd)
* [Probe level studies](Probelevel.Rmd)
* [Stability](Stability.Rmd)

### Linear models, comparisons, and association studies
* [Linear models](limma.Rmd)
* [Pairwise comparisons](Comparisons.Rmd)
* [Cross correlations](Crosscorrelation.Rmd)

### Other statistical analysis
* [ROC curves](ROC.Rmd)
* [RDA](RDA.Rmd)

### Miscellaneous
* [leaveout](leaveout.Rmd)
* [misc](misc.Rmd)



### Licensing and Citations

This work can be freely used, modified and distributed under the 
[Two-clause FreeBSD license](http://en.wikipedia.org/wiki/BSD\_licenses).

Kindly cite the work as 'Leo Lahti and Jarkko Salojarvi
(2014). microbiome R package. URL: http://microbiome.github.com'.


### References

```{r, echo=FALSE, message=FALSE}
require(knitcitations)
cleanbib()
#options("citation_format" = "pandoc")
bib <- read.bibtex("bibliography.bib")
```

The package utilizes tools from a number of other CRAN and
Bioconductor extensions, including:

 * df2json `r citep(citation("df2json"))`
 * rjson `r citep(citation("rjson"))`
 * ade4 `r citep(citation("ade4"))`
 * mixOmics `r citep(citation("mixOmics"))`
 * RCurl `r citep(citation("RCurl"))`
 * vegan `r citep(citation("vegan"))`
 * reshape `r citep(citation("reshape"))`
 * WGCNA `r citep(citation("WGCNA"))`
 * ggplot2 `r citep(citation("ggplot2"))`
 * RPA `r citep(citation("RPA"))`
 * minet `r citep(citation("minet"))`
 * fastcluster `r citep(citation("fastcluster"))`

```{r, echo=FALSE, message=FALSE}
#You can embed citations, for example: `r citep(bib[["lahti14natcomm"]])`
#You can embed citations, for example2: @lahti14natcomm
#Cite with DOI: `r citep("10.1890/11-0011.1")`
#Cite URL `r citep("http://knowledgeblog.org/greycite")`
#For automated markdown citations, check [this](http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html), [this](https://github.com/cboettig/knitcitations), and [this](http://www.carlboettiger.info/2012/03/24/citations-in-markdown-using-knitr.html). 
write.bibtex(file="references.bib")
```

The preprocessing of HITChip data is based on RPA [`r citep(citation("RPA"))`; `r citep("10.1109/TCBB.2009.38")`]

The peerj32 example data set is from `r citep("10.7717/peerj.32")`.


```{r, echo=FALSE, message=FALSE, results='asis'}
bibliography()
```



 * L. Lahti et al. Associations between the human intestinal microbiota, Lactobacillus rhamnosus GG and serum lipids indicated by integrated analysis of high-throughput profiling data. [PeerJ 1:e32, 2013](http://dx.doi.org/10.7717/peerj.32).
Lahti13provasI

 * R Core Team (2013). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. [ISBN 3-900051-07-0](http://www.R-project.org/)

 * W. N. Venables and B. D. Ripley (2002) Modern Applied Statistics with S. Fourth Edition. Springer, New York. ISBN 0-387-95457-0

 * H. Wickham (2011). The Split-Apply-Combine Strategy for Data Analysis. [Journal of Statistical Software, 40(1), 1-29](http://www.jstatsoft.org/v40/i01/)


### Session info

This vignette was created with

```{r sessioninfo}
sessionInfo()
```




