---
title: "microbiome vignette"
author: "Leo Lahti, Tineka Blake and Jarkko Salojarvi"
date: "`r Sys.Date()`"
bibliography: 
- bibliography.bib
- references.bib
output: 
  rmarkdown::html_vignette
---
<!--
  %\VignetteIndexEntry{microbiome tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
  %\VignetteEncoding{UTF-8}
-->

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Handle citations
require(knitcitations)
cleanbib()
#options("citation_format" = "pandoc")
bib <- read.bibtex("bibliography.bib")
#opts_chunk$set(fig.width=4, fig.height=3, par=TRUE, out.width='2in', fig.pos='H')
knitr::opts_chunk$set(fig.path = "figure/", dev="CairoPNG")
```


microbiome R package
===========

Tools for microbiome analysis in R `r citep(bib[["rcore13"]])`.

The microbiome R package provides analysis tools, diagnostic plots,
and other utilities for microbiome analyses, as well as multiple
example data sets.

We use the [phyloseq](http://joey711.github.io/phyloseq/import-data)
class, a standard representation format in R for taxonomic
profiling. This package provides extensions and convenient wrappers
for many standard tasks encountered in microbiome studies. 


### Getting started

* [Installation](Template.md) 
* [Example data](Data.md)
* [Data manipulation](Preprocessing.md)


### Visualization and related tools

* [Community composition](Composition.md)
* [Heatmaps](Heatmap.md)
* [Networks](Networks.md)
* [Ordination (PCA, PCoA, NMDS etc.)](Ordination.md)
* [Density](Density.md)


### Microbiota analysis

* [Core microbiota](Core.md)
* [Diversity](Diversity.md)
* [Linear models](linearmodels.md)
* [Pairwise comparisons](Comparisons.md)
* [Stability, bimodality, tipping elements](Stability.md)
* [Variability](Variability.md)
* [Experimental](Experimental.md)



### Licensing and Citations

This work can be freely used, modified and distributed under the
[Two-clause FreeBSD license](http://en.wikipedia.org/wiki/BSD\_licenses).

Kindly cite this work as follows:

```{r sotkanetIndicators, warning=FALSE, message=FALSE, eval=TRUE}
citation('microbiome')
```


### Dependencies

The package utilizes tools from a number of other R extensions,
including ade4 `r citep(citation("ade4"))`, dplyr `r citep(citation("dplyr"))`, ggplot2 `r citep(citation("ggplot2"))`, MASS `r citep(citation("MASS"))`, moments `r citep(citation("moments"))`, phyloseq `r citep(citation("phyloseq"))`, RColorBrewer `r citep(citation("RColorBrewer"))`, scales `r citep(citation("scales"))`, stats `r citep(citation("stats"))`, tidyr `r citep(citation("tidyr"))`, vegan `r citep(citation("vegan"))`.


### References

```{r, echo=FALSE, message=FALSE}
#You can embed citations, for example: `r citep(bib[["lahti14natcomm"]])`
#You can embed citations, for example2: @lahti14natcomm
#Cite with DOI: `r citep("10.1890/11-0011.1")`
#Cite URL `r citep("http://knowledgeblog.org/greycite")`
#For automated markdown citations, check [this](http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html), [this](https://github.com/cboettig/knitcitations), and [this](http://www.carlboettiger.info/2012/03/24/citations-in-markdown-using-knitr.html). 
#write.bibtex(file="references.bib")
```

```{r, echo=FALSE, message=FALSE, results='asis'}
bibliography()
```

