<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{An R Markdown Vignette made with knitr}
-->

microbiome R package
===========

### Installing and loading the release version (not available yet)

For general users:

```{r install, eval=FALSE}
source("http://www.bioconductor.org/biocLite.R")
biocLite("microbiome")
library(microbiome)
```

### Installing and loading the development version

For developers:

```{r install2, eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("microbiome", "microbiome")
library(sorvi)
```

### Examples

Further installation and usage instructions can be found at the
project [wiki](https://github.com/microbiome/microbiome/wiki/). 


### Licensing and Citations

This work can be freely used, modified and distributed under the 
[Two-clause FreeBSD license](http://en.wikipedia.org/wiki/BSD\_licenses).

Kindly cite the work, if appropriate, as 'Leo Lahti and Jarkko Salojärvi (2013). URL: http://microbiome.github.com'. 


### Session info


This vignette was created with

```{r sessioninfo}
sessionInfo()
```




