---
title: "Differential abundance testing"
author: "`r Sys.Date()`"
bibliography: 
- bibliography.bib
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: true
    use_bookdown: false
    highlight: haddock
---
<!--
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{microbiome tutorial - atlas}
  %\usepackage[utf8]{inputenc}
-->



```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Handle citations
require(knitcitations)
require(bookdown)
# cleanbib()
# options("citation_format" = "pandoc")
bib <- read.bibtex("bibliography.bib")
#opts_chunk$set(fig.width=4, fig.height=3, par=TRUE, out.width='2in', fig.pos='H')
library(knitr)
knitr::opts_chunk$set(fig.path = "figure/", dev="CairoPNG")
theme_set(theme_bw(20))
```


# Differential abundance testing: univariate data

This section covers basic univariate tests for two-group comparison,
covering t-test, Wilcoxon test, and multiple testing.

The following example compares the abundance of a selected bug between
two conditions. Let us assume that the data is already properly
normalized.

Let us load example data

```{r univariate1, warning=FALSE, message=FALSE}
library(microbiome)
data(dietswap)
d <- dietswap

# Pick microbial abundances for a given taxonomic group
taxa <- "Dialister"

# Construct a data.frame with the selected
# taxonomic group and grouping
df <- data.frame(Abundance = abundances(d)[taxa,],
                 Group = meta(d)$nationality)

```


Compare the groups visually using a boxplot (left). However, we
observe that the abundances are in absolute scale and therefore the
comparison is not clear.  Let us try the log10 transformation.  Now,
the data contains many zeros and taking log10 will yield infinite
values. Hence we choose the commonly used, although somewhat
problematic, log10(1+x) transformation (right).


```{r univariate_boxplot, warning=FALSE, message=FALSE, fig.width=10, fig.height=5, out.width="200px"}
p1 <- ggplot(df, aes(x = Group, y = Abundance)) +
       geom_boxplot() +
       labs(title = "Absolute abundances", y = "Abundance (read count)")

# Let us add the log10(1+x) version:
df$Log10_Abundance <- log10(1 + df$Abundance)
p2 <- ggplot(df, aes(x = Group, y = Log10_Abundance)) +
       geom_boxplot() +
       labs(title = "Log10 abundances", y = "Abundance (log10(1+x) read count)")       

library(patchwork)
p1 + p2
```



The groups seem to differ. Let us test the difference statistically.
First, let us perform t-test, which is based on Gaussian assumptions.
Each group is expected to follow Gaussian distribution.

Significance p-value with t-test:

```{r univariate2b, warning=FALSE, message=FALSE}
print(t.test(Log10_Abundance ~ Group, data = df)$p.value)
```


Now let us investigate the Gaussian assumption in more
detail. Boxplots may not show deviations from Gaussian assumptions
very clearly Let us try another visualization; the density plot.

```{r univariate_densityplot, warning=FALSE, message=FALSE}
p <- ggplot(df, aes(fill = Group, x = Log10_Abundance)) +
       geom_density(alpha = 0.5)
print(p)
```


Apparently, the data is not Gaussian distributed.  In such cases, a common
procedure is to use non-parametric tests. These do not make
assumptions of the data distribution but instead compare the ordering
of the samples.



So, let us look at the significance p-value with Wilcoxon test (log10 data):

```{r univariate_wilcoxon, warning=FALSE, message=FALSE}
print(wilcox.test(Log10_Abundance ~ Group, data = df)$p.value)
```


But since the test is non-parametric, we can as well use the original absolute abundances since the 
log transformation does not change sample ordering on which the Wilcoxon test is based.

Let us verify that the absolute abundances yield the same p-value for Wilcoxon test:

```{r univariate_wilcoxon2, warning=FALSE, message=FALSE}
print(wilcox.test(Abundance ~ Group, data = df)$p.value)
```


Let us compare how much the results would differ in the whole data
between t-test and Wilcoxon test. To remove non-varying taxa that
would demand extra scripting, let us for demonstration purposes now
focus on core taxa that are observed in more than 20% of the samples
with more than 3 reads.

```{r univariate4, warning=FALSE, message=FALSE}
# Core taxa to be tested
test.taxa <- core_members(d, prevalence = 20/100, detection = 3)

# Calculate p-values with the two different methods for each taxonomic unit
pvalue.ttest <- c()
pvalue.wilcoxon <- c()
for (taxa in test.taxa) {
  # Create a new data frame for each taxonomic group
  df <- data.frame(Abundance = abundances(d)[taxa,],
                   Log10_Abundance = log10(1 + abundances(d)[taxa,]),  
                   Group = meta(d)$nationality)

  pvalue.ttest[[taxa]] <- t.test(Log10_Abundance ~ Group, data = df)$p.value
  pvalue.wilcoxon[[taxa]] <- wilcox.test(Abundance ~ Group, data = df)$p.value  
}
# Arrange the results in a data.frame
pvalues <- data.frame(taxon = test.taxa,
	              pvalue.ttest = pvalue.ttest,
                      pvalue.wilcoxon = pvalue.wilcoxon)

# Note that multiple testing occurs.
# We must correct the p-values.
# let us apply the standard Benjamini-Hochberg False Discovery Rate (FDR)
# correction
pvalues$pvalue.ttest.adjusted <- p.adjust(pvalues$pvalue.ttest)
pvalues$pvalue.wilcoxon.adjusted <- p.adjust(pvalues$pvalue.wilcoxon)
```


Compare the p-value histograms between raw and adjusteed p-values.

```{r univariate5, warning=FALSE, message=FALSE, fig.width=10, fig.height=5, out.width="200px"}
p1 <- ggplot(pvalues, aes(x = pvalue.wilcoxon)) +
        geom_histogram() +
        labs(title = "Raw p-values") +
	ylim(c(0, 80))

p2 <- ggplot(pvalues, aes(x = pvalue.wilcoxon.adjusted)) +
        geom_histogram() +
        labs(title = "Adjusted p-values") +
	ylim(c(0, 80))	

print(p1 + p2)
```


Now compare these adjusted p-values between t-test and Wilcoxon test. Let us
also highlight the p = 0.05 intervals.

```{r univariate6, warning=FALSE, message=FALSE, fig.width=8, fig.height=8, out.width="200px"}
p <- ggplot(data = pvalues,
         aes(x = pvalue.ttest.adjusted,
	     y = pvalue.wilcoxon.adjusted)) +
       geom_text(aes(label = taxon)) + 
       geom_abline(aes(intercept = 0, slope = 1)) +
       geom_hline(aes(yintercept = 0.05), shape = 2) +
       geom_vline(aes(xintercept = 0.05), shape = 2)
print(p)
```






# Linear models: the role of covariates

This section provides a brief hands-on introduction to the practical motivations
and use linear (and generalized linear) models.


Let us compare two groups with a linear model. We use Log10 abundances
since this is closer to the Gaussian assumptions than the absolute
count data. We can fit a linear model with Gaussian variation as follows:

```{r lm, warning=FALSE, message=FALSE}
res <- glm(Log10_Abundance ~ Group, data = df, family = "gaussian")
```


Let us investigate model coefficients

```{r lm_coefs, warning=FALSE, message=FALSE}
kable(summary(res)$coefficients, digits = 5)
```


The intercept equals to the mean in the first group:


```{r lm_coefs3, warning=FALSE, message=FALSE}
print(mean(subset(df, Group == "AAM")$Log10_Abundance))
```

The group term equals to the difference between group means:

```{r lm_coefs3b, warning=FALSE, message=FALSE}
print(mean(subset(df, Group == "AFR")$Log10_Abundance) -
      mean(subset(df, Group == "AAM")$Log10_Abundance))
```


Note that the linear model (default) significance equals to t-test assuming equal variances.

```{r lm_signif, warning=FALSE, message=FALSE}
print(t.test(Log10_Abundance ~ Group, data = df, var.equal=TRUE)$p.value)
```


An important advantage of linear models, compared to plain t-test is
that they allow incorporating additional variables, such as potential
confounders (age, BMI, gender..):

```{r lm_glm, warning=FALSE, message=FALSE}
# Add a covariate:
df$sex <- meta(d)$sex

# Fit the model:
res <- glm(Log10_Abundance ~ Group + sex, data = df, family = "gaussian")
```


## Generalized linear models: a very brief overview

Let us briefly discuss the ideas underlying generalized linear models
before proceeding to the next section.

The Generalized linear model (GLM) allows a richer family of
probability distributions to describe the data. Intuitively speaking,
GLMs allow the modeling of nonlinear, nonsymmetric, and nongaussian
associations. GLMs consist of three elements:
  - A probability distribution (from exponential family)
  - A linear predictor η = Xβ .
  - A link function g such that $E(Y) = μ = g^{−1}(η)$.

We use Poisson with (its natural) log-link. Fit abundance (read
counts) assuming that the data is Poisson distributed, and the
logarithm of its mean, or expectation, is obtained with a linear
model.

```{r lm_glm2, warning=FALSE, message=FALSE}
res <- glm(Abundance ~ Group, data = df, family = "poisson")
```


Investigate the model output:

```{r lm_glm2b, warning=FALSE, message=FALSE}
kable(summary(res)$coefficients, digits = 5)
```




# Advanced models of differential abundance

GLMs are the basis for advanced testing of differential abundance in
sequencing data. This is necessary, as the sequencing data sets
deviate from symmetric, continuous, Gaussian assumptions in many ways.


Sequencing data consists of discrete counts:

```{r pooled2, warning=FALSE, message=FALSE}
print(abundances(d)[1:5,1:3])
```


The data is sparse:

```{r pooled3, warning=FALSE, message=print, fig.width=8, fig.height=8}
hist(log10(1 + abundances(d)), 100)
```


Long tails of rare taxa:

```{r tail, warning=FALSE, message=print, fig.width=16, fig.height=8, out.width="200px"}
medians <- apply(abundances(d),1,median)/1e3
A <- melt(abundances(d))
A$Var1 <- factor(A$Var1, levels = rev(names(sort(medians))))
p <- ggplot(A, aes(x = Var1, y = value)) +
        geom_boxplot() +
	labs(y = "Abundance (reads)", x = "Taxonomic Group") +
	scale_y_log10()
	
print(p)
```

Overdispersion (variance exceeds the mean):

```{r pooled_overdispersion, warning=FALSE, message=FALSE, fig.width=8, fig.height=8, out.width="200px"}
means <- apply(abundances(d),1,mean)
variances <- apply(abundances(d),1,var)

# Calculate mean and variance over samples for each taxon
library(reshape2)
library(dplyr)
df <- melt(abundances(d))
names(df) <- c("Taxon", "Sample", "Reads")
df <- df %>% group_by(Taxon) %>%
             summarise(mean = mean(Reads),
	               variance = var(Reads))

# Illustrate overdispersion
library(scales)
p <- ggplot(df, aes(x = mean, y = variance)) +
       geom_point() +
       geom_abline(aes(intercept = 0, slope = 1)) +
       scale_x_log10(labels = scales::scientific) +
       scale_y_log10(labels = scales::scientific) +
       labs(title = "Overdispersion (variance > mean)")
print(p)
```


## DESeq2: differential abundance testing for sequencing data

DESeq2 analysis can accommodate those particular assumptions about
sequencing data. 

```{r pooled_deseq, warning=FALSE, message=FALSE}
# Start by converting phyloseq object to deseq2 format
library(DESeq2)
ds2 <- phyloseq_to_deseq2(d, ~ group + nationality)

# Run DESeq2 analysis (all taxa at once!)
dds <- DESeq(ds2)

# Investigate results
res <- results(dds)
deseq.results <- as.data.frame(res)
df <- deseq.results
df$taxon <- rownames(df)
df <- df %>% arrange(log2FoldChange, padj)

# Print the results; flitered and sorted by pvalue and effectsize
library(knitr)
df <- df %>% filter(pvalue < 0.05 & log2FoldChange > 1.5) %>%
             arrange(pvalue, log2FoldChange)
kable(df, digits = 5)
```


For comparison purposes, assess significances and effect sizes based on Wilcoxon test.

```{r pooled_topw, warning=FALSE, message=FALSE, fig.width=10, fig.height=5, out.width="200px"}
test.taxa <- taxa(d)
pvalue.wilcoxon <- c()
foldchange <- c()
for (taxa in test.taxa) {
  # Create a new data frame for each taxonomic group
  df <- data.frame(Abundance = abundances(d)[taxa,],
                   Log10_Abundance = log10(1 + abundances(d)[taxa,]),
                   Group = meta(d)$nationality)
  # Calculate pvalue and effect size (difference beween log means)		 
  pvalue.wilcoxon[[taxa]] <- wilcox.test(Abundance ~ Group, data = df)$p.value
  foldchange[[taxa]] <- coef(lm(Log10_Abundance ~ Group, data = df))[[2]]
}
# Correct p-values for multiple testing
pvalue.wilcoxon.adjusted <- p.adjust(pvalue.wilcoxon)
```



```{r pooled_pcomp, warning=FALSE, message=FALSE, fig.width=10, fig.height=5}
par(mfrow = c(1,2))
plot(deseq.results$padj, pvalue.wilcoxon.adjusted,
  xlab = "DESeq2 adjusted p-value",
  ylab = "Wilcoxon adjusted p-value",
  main = "P-value comparison")
abline(v = 0.05, h = 0.05, lty = 2)

plot(deseq.results$log2FoldChange, foldchange, 
  xlab = "DESeq2",
  ylab = "Linear model",
  main = "Effect size comparison")
abline(0,1)
```






