### Oligoprofile heatmap

Reproduce and modify the oligo heatmap of the [run.profiling.script](profiling). Plots heatmap of the oligo profiles together with phylotype groups and sample clusters. Reload and plot preprocessed data from the [run.profiling.script](profiling) by using the [read.profiling](reading) function, assuming you have stored the output in directory "datadirectory/". Start by reading the data:

```{r heatmap-example2}
# Define data directory (here: simulated data directory)
data.directory <- system.file("extdata", package = "microbiome")

# Read Oligo level data in original domain
oligo.matrix.log10.simulated <- read.profiling(level = "oligo", 
			              data.dir = data.directory, log10 = TRUE)

# Read Oligo-phylogeny mapping table (two methods):
phylogeny.info <- GetPhylogeny("HITChip", "full")
```

Save the image in a file (too large to be opened directly in
R). To prevent ordering of the rows, use hclust.method = NULL in the
function call:

```{r example3, message=FALSE, eval=FALSE}
library(microbiome)
ppcm <- 150
png(filename = "oligoprofileClustering.png", 
     width = max(trunc(ppcm*21), 
                 trunc(ppcm*21*ncol(oligo.matrix.log10.simulated)/70)), 
     height = trunc(ppcm*29.7))


library(HITChipDB)
tmp <- PlotPhylochipHeatmap(oligo.matrix.log10.simulated, 
				phylogeny.info, level = "L1", metric = "pearson")
dev.off()
```

